---
title: "Making a Map in R"
author: "Chris"
date: "2/14/2023"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Here are the pitfalls

* Google API key needed
* Old or defunct packages or package dependencies
* I was avoiding packages that have failed me in the past without thinking of how some of them could be used together

### Website this base code was altered from: https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html

Map of the USA

```{r}

install.packages("mapdata")
install.packages('maps')
library(ggplot2, ggmap, 'maps')

usa <- map_data("usa")
gg1 <-ggplot() + geom_polygon(data = usa, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3) #run code without object to see in this doc

```

Mapping WA State
```{r}
states <- map_data("state")
head(states)

#install.packages("cowplot")

#library("cowplot")

wa_df <- subset(states, region== "washington")
head(wa_df)

counties <- map_data("county")
wa_county <- subset(counties, region== "washington")
head(wa_county)

wa_base <- ggplot(data = wa_df, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
wa_base + theme_nothing()

wac <- wa_base + theme_nothing() + 
  geom_polygon(data = wa_county, fill = NA, color = "white") +
  geom_polygon(color = "black", fill = NA)  # get the state border back on top

```

Mapping Puget Sound and my cage retrieval sites
```{r}

location<- read.csv("mussel.csv")
head(location)
lat <- location[,5]
long <- location[,6]
site <- location[,2]
cage <- data.frame(long, lat, site)
head(cage)

#isolating puget sound area
wac + coord_fixed(xlim = c(-124, -122),  ylim = c(47, 49), ratio = 1.3)

#adding cage sites
wac +
  coord_fixed(xlim = c(-124, -122),  ylim = c(47, 49), ratio = 1.3) +
  geom_point(data = cage, aes(x = long, y = lat), color = "black", size = 3, inherit.aes = FALSE) +
  geom_point(data = cage, aes(x = long, y = lat), color = "yellow", size = 2, inherit.aes = FALSE)


```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
