---
title: "01- Exploratory Statistics: Distribution, Variance and Outlier Detection"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

# double check where you are
# *NOTE* see 00-data_cleaned for directory issue
getwd()

# load data
data<- read.csv("../data/all_lab_data_raw.csv")

# check it out
str(data) # checking df structure
# note that sample_id and site_number are integers and may need to be adjusted to characters for analysis

summary(data) #checking df contents

```

# Histograms - all data
```{r}
# p450 basic histogram + basic density plot
hist(data$p450)
#plot(density(data$p450, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# SOD basic histogram + basic density plot
hist(data$sod)
#plot(density(data$sod, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Condition_factor basic histogram + basic density plot
hist(data$ci_wet_volume)
hist(data$ci_wet_shell_weight)
hist(data$ci_pollution)
#plot(density(data$condition_factor, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Avg_thickness basic histogram + basic density plot
hist(data$shell_thickness)
#plot(density(data$avg_thickness, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

```

# Histograms - outliers removed - Don't Run
```{r}
# p450
filtered_data <- data[!data$p450_outlier, ]
hist(filtered_data$p450, main="Histogram of p450 (Outliers Removed)", xlab="p450 Value")
plot(density(filtered_data$p450, na.rm=TRUE), main="Density Plot (Outliers Removed)", xlab="p450 Value")


# SOD 
filtered_datas <- data[!data$sod_outlier, ]
hist(filtered_datas$sod)
plot(density(filtered_datas$sod, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Condition_factor
filtered_datac <- data[!data$condition_factor_outlier, ]
hist(filtered_datac$condition_factor)
plot(density(filtered_datac$condition_factor, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Avg_thickness
filtered_dataa <- data[!data$avg_thickness_outlier, ]
hist(filtered_datac$avg_thickness)
plot(density(filtered_datac$avg_thickness, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

```

# Significance
## Variance, p= 0.7175, we can use PERMANOVA for the stats
```{r}

# when we looked at our plots, the data is highly variable and because of that it may incorrectly inflate our PERMANOVA results. Reminder, we are running PERMANOVAs on p450, sod and condition_factor because they aren't normally distributed). We check the variance of each group to determine if the PERMANOVA is the correct test. I used the 'bray' method for the matrix because it is ecological data and that's the standard. Note: variance can be evaluated with either bray or euclidian methods, bray is the most common until we get to permanova where euclidian method is used for continuous data. After testing both methods, the result is the same. 

# packages for these tests
library(vegan)
library(cluster)

data$site_name <- as.factor(data$site_name) # change site names to a factor instead of a character

metric_matrix <- as.matrix(data[, 10:15]) # pull out my metrics - confirm column numbers are still correct

dist_matrix_variance <- vegdist(metric_matrix, na.rm= TRUE, method = "bray") # create a distance matrix

dispersion_test <- betadisper(dist_matrix_variance, group = data$site_name) # check homogeneity of group variances


anova(dispersion_test) # Perform permutation test for homogeneity. A p-value > .05 means we need to use PERMANOVA. p= 0.7359

## *NOTE*: I don't actually know what it was supposed to look like, so I can followup in the future. Visualize results (boxplot of dispersions)
boxplot(dispersion_test) # visualize results (boxplot of dispersions)

```

# PERMANOVA & Post Hoc
## Site (n=74)
### No significant results, p= 0.083
```{r}

# we can proceed with the PERMANOVA based on the analysis of variance above. Let's check if Site has a statistically significant relationship with our metrics. Instead of creating a distance matrix for each individual metric, I am going to create one that housing all metrics of interest (biomarkers & morphometrics). This is not an ideal process with many metrics, but 4 metrics makes an easy test case for the process.

#install.packages("RVAideMemoire") # only need to install once

library(vegan)
library(RVAideMemoire)

dist_matrix_full <- vegdist(as.matrix(data[, c("p450", "sod", "ci_wet_volume", "ci_wet_shell_weight", "ci_pollution", "shell_thickness")]), na.rm= TRUE, method = "euclidean") # create distance matrix

permanova_full <- adonis2(dist_matrix_full ~ site_name, data = data, permutations = 999) # run PERMANOVA

pairwise_results <- list() # list to store post hoc result
significant_results <- list() # list to store only significant post hoc results

pairwise_model <- pairwise.perm.manova(dist_matrix_full, data$site_name, nperm = 999, p.method = "bonferroni") # run post hoc 

pairwise_df <- as.data.frame(pairwise_model$p.value) # Extract the results table correctly from `pairwise.htest` object
colnames(pairwise_df) <- c("p.value")  # Ensure column name is consistent

pairwise_results[["Site"]] <- pairwise_df # Store full results

significant_results[["Site"]] <- pairwise_df[pairwise_df$p.value < 0.05, , drop = FALSE] # filter significant results

cat("\n### PERMANOVA results for All Sites ###\n")
print(permanova_full) # view permanova results

if (nrow(significant_results[["Site"]]) > 0) {
  cat("\n### Significant pairwise post hoc results (p < 0.05) ###\n")
  print(significant_results[["Site"]])
} else {
  cat("\n### No significant pairwise differences found in All Sites ###\n") # view significant post hoc results
}

# write out significant results
#write.csv(significant_results[["Full Data"]], "significant_pairwise_results_full_data.csv", row.names = FALSE)

```

## Reporting Area (n=9)
### No significant results, p= 0.32
```{r}

library(vegan)
library(RVAideMemoire)

# don't need to recreate the matrix, we will use the same matrix created in the first step
#dist_matrix_full <- vegdist(as.matrix(data[, c("p450", "sod", "ci_wet_volume", "ci_wet_shell_weight", "ci_pollution", "shell_thickness")]), na.rm= TRUE, method = "euclidean") # create distance matrix

data$reporting_area <- as.factor(data$reporting_area) # make a factor for analysis

permanova_full <- adonis2(dist_matrix_full ~ reporting_area, data = data, permutations = 999) # run PERMANOVA

pairwise_results <- list() # list to store post hoc result
significant_results <- list() # list to store only significant post hoc results

pairwise_model <- pairwise.perm.manova(dist_matrix_full, data$reporting_area, nperm = 999, p.method = "bonferroni") # run post hoc 

pairwise_df <- as.data.frame(pairwise_model$p.value) # Extract the results table correctly from `pairwise.htest` object
colnames(pairwise_df) <- c("p.value")  # Ensure column name is consistent

pairwise_results[["Reporting Area"]] <- pairwise_df # Store full results

significant_results[["Reporting Area"]] <- pairwise_df[pairwise_df$p.value < 0.05, , drop = FALSE] # filter significant results

cat("\n### PERMANOVA results for All Reporting Areas ###\n")
print(permanova_full) # view permanova results

if (nrow(significant_results[["Reporting Area"]]) > 0) {
  cat("\n### Significant pairwise post hoc results (p < 0.05) ###\n")
  print(significant_results[["Reporting Area"]])
} else {
  cat("\n### No significant pairwise differences found in All Reporting Areas ###\n") # view significant post hoc results
}

# write out significant results
#write.csv(significant_results[["Full Data"]], "significant_pairwise_results_full_data.csv", row.names = FALSE)

```

## PCB Groups (n=8)
### No significant results, p= 0.26
```{r}

library(vegan)
library(RVAideMemoire)

# don't need to recreate the matrix, we will use the same matrix created in the first step
#dist_matrix_full <- vegdist(as.matrix(data[, c("p450", "sod", "ci_wet_volume", "ci_wet_shell_weight", "ci_pollution", "shell_thickness")]), na.rm= TRUE, method = "euclidean") # create distance matrix

data<- read.csv("../data/cleaned/cleaned_with_pcb_groups.csv")
data$pcb_group <- as.factor(data$pcb_group) # make a factor for analysis

permanova_full <- adonis2(dist_matrix_full ~ pcb_group, data = data, permutations = 999) # run PERMANOVA

pairwise_results <- list() # list to store post hoc result
significant_results <- list() # list to store only significant post hoc results

pairwise_model <- pairwise.perm.manova(dist_matrix_full, data$pcb_group, nperm = 999, p.method = "bonferroni") # run post hoc 

pairwise_df <- as.data.frame(pairwise_model$p.value) # Extract the results table correctly from `pairwise.htest` object
colnames(pairwise_df) <- c("p.value")  # Ensure column name is consistent

pairwise_results[["pcb_group"]] <- pairwise_df # Store full results

significant_results[["pcb_group"]] <- pairwise_df[pairwise_df$p.value < 0.05, , drop = FALSE] # filter significant results

cat("\n### PERMANOVA results for All pcb_groups ###\n")
print(permanova_full) # view permanova results

if (nrow(significant_results[["pcb_group"]]) > 0) {
  cat("\n### Significant pairwise post hoc results (p < 0.05) ###\n")
  print(significant_results[["pcb_group"]])
} else {
  cat("\n### No significant pairwise differences found in All pcb_groups ###\n") # view significant post hoc results
}

# write out significant results
#write.csv(significant_results[["Full Data"]], "significant_pairwise_results_full_data.csv", row.names = FALSE)

```

## PAH Group
### No significant results, p= 0.836
```{r}

library(vegan)
library(RVAideMemoire)

# don't need to recreate the matrix, we will use the same matrix created in the first step
#dist_matrix_full <- vegdist(as.matrix(data[, c("p450", "sod", "ci_wet_volume", "ci_wet_shell_weight", "ci_pollution", "shell_thickness")]), na.rm= TRUE, method = "euclidean") # create distance matrix

data$pah_group <- as.factor(data$pah_group) # make a factor for analysis

permanova_full <- adonis2(dist_matrix_full ~ pah_group, data = data, permutations = 999) # run PERMANOVA

pairwise_results <- list() # list to store post hoc result
significant_results <- list() # list to store only significant post hoc results

pairwise_model <- pairwise.perm.manova(dist_matrix_full, data$pah_group, nperm = 999, p.method = "bonferroni") # run post hoc 

pairwise_df <- as.data.frame(pairwise_model$p.value) # Extract the results table correctly from `pairwise.htest` object
colnames(pairwise_df) <- c("p.value")  # Ensure column name is consistent

pairwise_results[["pah_group"]] <- pairwise_df # Store full results

significant_results[["pah_group"]] <- pairwise_df[pairwise_df$p.value < 0.05, , drop = FALSE] # filter significant results

cat("\n### PERMANOVA results for All pah_groups ###\n")
print(permanova_full) # view permanova results

if (nrow(significant_results[["pah_group"]]) > 0) {
  cat("\n### Significant pairwise post hoc results (p < 0.05) ###\n")
  print(significant_results[["pah_group"]])
} else {
  cat("\n### No significant pairwise differences found in All pah_groups ###\n") # view significant post hoc results
}

# write out significant results
#write.csv(significant_results[["Full Data"]], "significant_pairwise_results_full_data.csv", row.names = FALSE)

```

## Geo Group
### No significant results, p= 0.903
```{r}

library(vegan)
library(RVAideMemoire)

# don't need to recreate the matrix, we will use the same matrix created in the first step
#dist_matrix_full <- vegdist(as.matrix(data[, c("p450", "sod", "ci_wet_volume", "ci_wet_shell_weight", "ci_pollution", "shell_thickness")]), na.rm= TRUE, method = "euclidean") # create distance matrix

data$geo_group <- as.factor(data$geo_group) # make a factor for analysis

permanova_full <- adonis2(dist_matrix_full ~ geo_group, data = data, permutations = 999) # run PERMANOVA

pairwise_results <- list() # list to store post hoc result
significant_results <- list() # list to store only significant post hoc results

pairwise_model <- pairwise.perm.manova(dist_matrix_full, data$geo_group, nperm = 999, p.method = "bonferroni") # run post hoc 

pairwise_df <- as.data.frame(pairwise_model$p.value) # Extract the results table correctly from `pairwise.htest` object
colnames(pairwise_df) <- c("p.value")  # Ensure column name is consistent

pairwise_results[["geo_group"]] <- pairwise_df # Store full results

significant_results[["geo_group"]] <- pairwise_df[pairwise_df$p.value < 0.05, , drop = FALSE] # filter significant results

cat("\n### PERMANOVA results for All geo_groups ###\n")
print(permanova_full) # view permanova results

if (nrow(significant_results[["geo_group"]]) > 0) {
  cat("\n### Significant pairwise post hoc results (p < 0.05) ###\n")
  print(significant_results[["geo_group"]])
} else {
  cat("\n### No significant pairwise differences found in All geo_groups ###\n") # view significant post hoc results
}

# write out significant results
#write.csv(significant_results[["Full Data"]], "significant_pairwise_results_full_data.csv", row.names = FALSE)

```

# KW & Post Hoc
## Site
### No significant results, p= 0.083
```{r}

```

