---
  title: "011 - P450 Plots"
output:
  html_document: 
  toc: true
toc_float:
  collapsed: false
smooth_scroll: true
fig_width: 20
pdf_document: 
  fig_width: 20
fig_height: 9
---
  
# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" )# Align plots to the center

```

# Load data
```{r}

data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/p450data.csv")
indv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgP450_analytes.csv")

```

# p450
## dumbbell - original data in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_p450_range <- data %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_p450 = min(p450), max_p450 = max(p450), mean_p450 = mean(p450))

# Order by site name
site_p450_range <- site_p450_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_p450_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_p450, xend = max_p450, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_p450, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_p450, color = as.factor(reporting_area)), size = 3) +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "P450 Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - log transformed data in alphabetical order
```{r}
library(ggplot2)
library(dplyr) 

# Plot
ggplot(site_p450_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_p450, xend = max_p450, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_p450, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_p450, color = as.factor(reporting_area)), size = 3) +
  scale_x_log10() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein) (Log Scale)") , y = "Site Name", title = "P450 Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - log transformed data grouped by reporting area with indicators of mean value (triangles)
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
site_p450_range <- site_p450_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_p450_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean added
ggplot(site_p450_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_p450, xend = max_p450, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_p450, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_p450, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_p450), shape = 17, size = 3, color = "black") +
  scale_x_log10() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein) (Log Scale)"), y = "Site Name", title = "P450 Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Psite_p450_range <- data %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_p450 = min(p450), max_p450 = max(p450), mean_p450 = mean(p450))

# Create a unique identifier for each combination of site_name and PAHgroup8
Psite_p450_range <- Psite_p450_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Psite_p450_range <- Psite_p450_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Psite_p450_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Psite_p450_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_p450, xend = max_p450, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_p450, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_p450, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_p450), shape = 17, size = 3, color = "black") +
  scale_x_log10() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein) (Log Scale)"), y = "Site Name", title = "P450 Activity by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}
library(ggplot2)
library(dplyr)

# Order by site name
#data <- data %>%
#  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

# Plot
ggplot(data, aes(x = site_name, y = p450, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein)"), title = "P450 Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## boxplot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- data %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = p450, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein)"), title = "P450 Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
pahbox <- data %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

pahbox <- pahbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
pahbox <- pahbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(pahbox, aes(x = site_name_ordered, y = p450, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = bquote("P450 Activity (Unit / mg"^"-1"~"protein)"), title = "P450 Activity by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```


## correlation plot
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
corr_results <- sapply(indv[ , 3:ncol(indv)], function(x) {
  test <- cor.test(indv$avg_value, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
spearman_cor_matrix <- cor(indv[ , 3:ncol(indv)], use = "complete.obs", method = "spearman")
spearman_cor_avg_value <- spearman_cor_matrix["avg_value", ]

# Filter for significance using corr_results for p-values
significant_analytes <- names(corr_results[corr_results < 0.05])

# New correlation matrix with only significant analytes
significant_cor_matrix <- spearman_cor_matrix[significant_analytes, significant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(significant_cor_matrix, method = "circle")

```

## correlation table
```{r}

library(dplyr)
library(tibble)

# Spearman correlation 
cor_results <- sapply(indv[ , 3:ncol(indv)], function(x) {
  test <- cor.test(indv$avg_value, x, method = "spearman")
  c(correlation = test$estimate, p_value = test$p.value)
})

# Convert to data frame, fix column names and remove first row of old column names
cor_results_df <- as.data.frame(t(cor_results))

colnames(cor_results_df) <- c("correlation", "p_value")

cor_results_df <- cor_results_df %>%
  rownames_to_column(var = "analyte")

cor_results_df <- cor_results_df %>%
  slice(-1) 

# Filter for significance and round the results
significant_results_df <- cor_results_df %>%
  filter(p_value < 0.05)

significant_results_df <- significant_results_df %>%
  mutate(correlation = round(correlation, 3),
         p_value = round(p_value, 3))

# View table
print(significant_results_df)

# Write to table
#write.csv(significant_results_df,"/Users/cmantegna/Documents/WDFWmussels/output/tables/VIZp450correlation.csv", row.names = FALSE)
```

## linear regression - signficant analytes
```{r}

library(ggplot2)
library(gridExtra)

indv$reporting_area <- as.character(indv$reporting_area)

# Custom colors to match other plot colors
reporting_area_colors <- c(
  "6" = "red",
  "7" = "blue",
  "8.1" = "green",
  "8.2" = "purple",
  "9" = "orange",
  "10" = "yellow",
  "11" = "brown",
  "12" = "pink",
  "13" = "grey"
  
)

# Only significant analytes
significant_analytes <- significant_results_df$analyte

# Create an empty list to store plots
plot_list <- list()

# Loop through each significant analyte and create a linear regression plot
for (analyte in significant_analytes) {
  
  # Ensure the analyte exists in df before proceeding
  if (!analyte %in% colnames(indv)) {
    warning(paste("Analyte", analyte, "is not present in the data frame. Skipping..."))
    next
  }
  
  # Linear Model
  model <- lm(indv[[analyte]] ~ indv$avg_value)
  
  # Plot
p <- ggplot(indv, aes_string(x = "avg_value", y = analyte, color = "reporting_area")) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black", se = FALSE) +  
  scale_color_manual(values = reporting_area_colors) +  
  labs(title = paste("Linear Regression: Average P450 Activity vs", analyte),
       x = bquote("P450 Activity (Unit / mg"^"-1"~"protein)"),
       y = analyte) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend

# Store the plot in the list
plot_list[[analyte]] <- p

}

# Display the plots in a grid
do.call(grid.arrange, c(plot_list, ncol = 2))  # Adjust 'ncol' to set the number of columns in the grid

```






