---
title: "02- Analysis Groups"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

# double check where you are
# *NOTE* see 00-data_cleaned for directory issue
getwd()

# load data
data<- read.csv("../data/cleaned/all_cleaned.csv")

# check it out
str(data) # checking df structure
# note that sample_id and site_number are integers and may need to be adjusted to characters for analysis

summary(data) #checking df contents

#colnames(data) <- tolower(colnames(data))

```

# Adding Identifiers for Grouping
## PCB
### Note: Reporting Area is already in the df
```{r}

library(classInt)
library(dplyr)

df<- read.csv("../data/not_for_analysis/pcb_group_workbook.csv")

# Extract the DryValue column
dry_values <- df$DryValue

# Compute Fisher-Jenks breaks for 6 groups
num_classes <- 6
breaks <- classIntervals(dry_values, num_classes, style = "fisher")$brks

# Print the breakpoints
print(breaks)

# Create group assignments based on the breaks
df$DryValue_Group <- cut(dry_values, breaks = breaks, labels = 1:num_classes, include.lowest = TRUE)

# Check the first few rows with the new grouping
head(df)

# Assign the lowest group number for each site
df <- df %>%
  group_by(SiteName) %>%
  summarise(pcb_group = min(as.numeric(DryValue_Group))) %>%
  ungroup()

df <- df %>%
  rename(site_name = SiteName)

# Check results
print(df)

# Merge while keeping only matching site names
data <- left_join(data, df, by = "site_name")

# Check if each site in data now has a single assigned group
table(data$pcb_group)  # Ensure no duplicates within a site

# Check the first few rows
head(data)

# write out new table

write.csv(data, "all_cleaned_with_groups.csv", row.names = FALSE)

```


# Grouping Options: Individual Samples
#### Note: we will proceed with the groupings using the cleaned data with outliers removed

## Site (n=74)
### Note: Statistics (/01-stats.Rmd) confirm this grouping is an ineffective tool of analysis. Box plot added to illustrate difficulties
```{r}

# save plot
#ggsave(plot=geopgraphic_boxplot, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/geographic_boxplot_p450.png", width=15, height=12)
```

## Geographic (lat/ long) (n=5)

```{r}

# make geographic clusters based on lat/ long coordinates in 5 groups since it looks like there are 5 distinct areas of sites

# use k-means clustering 
set.seed(123)
clusters <- kmeans(data[, c("latitude", "longitude")], centers = 5)
data <- data %>%
  mutate(latlong_group = as.factor(clusters$cluster))

# lat/long ranges for each group
latlong_ranges <- data %>%
  group_by(latlong_group) %>%
  summarize(
    lat_range = paste0(round(min(latitude), 2), " - ", round(max(latitude), 2)),
    long_range = paste0(round(min(longitude), 2), " - ", round(max(longitude), 2))
  ) %>%
  mutate(label = paste0("Group ", latlong_group, "\nLat: ", lat_range, "\nLong: ", long_range))

# add lat/long ranges back to the df
data <- data %>%
  left_join(latlong_ranges, by = "latlong_group")

# remove outliers
filtered_data <- data %>% filter(!is_outlier)

# box plot
geographic_boxplot<- ggplot(filtered_data, aes(x = reorder(latlong_group, as.numeric(latlong_group)), y = p450)) +
  geom_boxplot(outlier.shape = NA, color = "gray50", show.legend = FALSE) + # Boxplot only
  theme_minimal() +
  theme(
    panel.grid = element_blank(), # Remove all grid lines
    panel.border = element_blank(), # Remove border
    axis.line = element_line(color = "black"), # Show only axis lines
    axis.text.x = element_blank(), # Remove x-axis text
    axis.text.y = element_text(face = "bold"), # Bold y-axis text
    axis.title = element_text(face = "bold"), # Bold axis titles
    axis.ticks.x = element_blank() # Remove x-axis ticks
  ) +
  labs(
    title = "P450 by Geographic Location",
    x = "Geographic Group",
    y = expression("P450 Activity (Unit / mg"^-1~"protein)"),
    shape = "Lat/Long Group"
  ) +
  scale_shape_manual(
    values = seq(0, length(unique(filtered_data$latlong_group)) - 1), # Unique shapes for legend
    labels = latlong_ranges$label # Group labels
  ) +
  guides(
    shape = guide_legend(override.aes = list(size = 3)) # Restore shapes in the legend
  ) +
  geom_point(
    data = data.frame(
      latlong_group = unique(filtered_data$latlong_group),
      p450 = NA
    ),
    aes(
      x = latlong_group,
      y = min(filtered_data$p450) - 0.1 * diff(range(filtered_data$p450)), # Adjust y position below plot
      shape = latlong_group
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  coord_cartesian(clip = "off") # Ensure shapes outside the plot area are visible

print(geographic_boxplot)

# save plot
#ggsave(plot=geopgraphic_boxplot, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/geographic_boxplot_p450.png", width=15, height=12)


```

## Reporting Area
```{r}

```

