---
title: "020- SOD tests for significance and correlation"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}
library(tinytex)
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

```{r}

#getwd()
data <- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/soddata.csv")
sdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD.csv")
indv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD_analytes.csv")
summed<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD_analytes2.csv")
smetal<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD_metals.csv")

```

# Adjust SOD individual values to reflect LOQ limitations
```{r}

# Data contains numbers below 0 that must be adjusted.These numbers represent samples whose values were below the limit of quantification using the SOD kit. There was some activity in the raw sample, but it is too far below the limit of the standard curve to be accurately quantified.


#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
data$sod[data$sod <= 0] <- 0.0025
both$sod[both$sod <= 0] <- 0.0025

```
# Check data

```{r}
summary(both)
```

# Shapiro-Wilkes 
## Data is not normally distributed
```{r}

#test for normality 
shapiro.test(data$sod)

```

# Kruskal-Wallis
## full data
Reporting Areas Are:\
6 - East Juan de Fuca Strait\
7 - San Juan Islands\
8.1 - Deception Pass, Hope Island, and Skagit Bay\
8.2 - Port Susan and Port Gardner\
9 - Admiralty Inlet\
10 - Seattle-Bremerton\
11 - Tacoma-Vashon\
12 - Hood Canal\
13 - South Puget Sound

```{r}

#test for significant interaction

data$PAHgroup8 <- as.character(data$PAHgroup8)

kruskal.test(sod ~ site_name, data = data)
kruskal.test(sod ~ reporting_area, data = data)
kruskal.test(sod ~ PAHgroup8, data = data)

```


# Dunn Test (post hoc)

```{r}
library(dunn.test)

# Site
dunn_site <- dunn.test(data$sod, data$site_name, method = "bonferroni", list = TRUE)
mc_site <- data.frame(
    Comparison = dunn_site$comparisons,
    Z = dunn_site$Z,
    P.adjusted = dunn_site$P.adjusted
)

# Write results to a table
#write.csv(mc_site, "/Users/cmantegna/Documents/WDFWmussels/output/tables/SignificantSODsite.csv", row.names = FALSE)

# PAHgroup8
dunn_pah8 <- dunn.test(data$sod, data$PAHgroup8, method = "bonferroni", list = TRUE)
mc_site <- data.frame(
    Comparison = dunn_pah8$comparisons,
    Z = dunn_pah8$Z,
    P.adjusted = dunn_pah8$P.adjusted
)

#write.csv(mc_site, "/Users/cmantegna/Documents/WDFWmussels/output/tables/SignificantSODpah8.csv", row.names = FALSE)
```

# Correlation

## biomarkers
```{r}

# negative correlation (corr= -0.3256127, p-value = 0.004821)
correlation_result <- cor.test(summed$avg_p450, summed$avg_value, method = "spearman")
print(correlation_result)

```

## sumPAH, lmwPAH, hmwPAH and PAH16
### No correlations
```{r}

# correlated
correlation_result <- cor.test(summed$avg_value, summed$sumPAH, method = "spearman")
print(correlation_result)

# correlated
correlation_result <- cor.test(summed$avg_value, summed$lmwPAH, method = "spearman")
print(correlation_result)

# correlated
correlation_result <- cor.test(summed$avg_value, summed$hmwPAH, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(summed$avg_value, summed$sum16PAH, method = "spearman")
print(correlation_result)

```

## sum42PAH and sumPCB
```{r}

# correlated
correlation_result <- cor.test(summed$avg_value, summed$sum42PAH, method = "spearman")
print(correlation_result)

# correlated
correlation_result <- cor.test(summed$avg_value, summed$sumPCB, method = "spearman")
print(correlation_result)

```

## other summed analytes
```{r}

# correlated
correlation_result <- cor.test(summed$avg_value, summed$sum40CB, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(summed$avg_value, summed$sumBDE, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(summed$avg_value, summed$sumCHLD, method = "spearman")
print(correlation_result)

# correlated
correlation_result <- cor.test(summed$avg_value, summed$sumDDT, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(summed$avg_value, summed$sumHCH, method = "spearman")
print(correlation_result)
```

## mercury, arsenic, cadmium, copper, lead and zinc
```{r}

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$mercuryTotal, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$arsenic, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$cadmium, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$copper, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$lead, method = "spearman")
print(correlation_result)

# no correlation
correlation_result <- cor.test(smetal$avg_value, smetal$Zinc, method = "spearman")
print(correlation_result)

```

