---
title: "031- Morphometrics Grouping Plots"
output:
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
  pdf_document: 
    fig_width: 20
    fig_height: 9
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}
library(tinytex)
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

```{r}

getwd()
mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphometricdata.csv")

```

# Check data

```{r}
summary(mdata)
```

# Plot data - boxplot, all condition factor values by site name
# *All analytes were plotted by site and reporting area in p450 flow.*

```{r}

mplot<- ggplot(mdata, aes(x = site_name, y = cf)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(mplot)
ggsave(plot=mplot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/CFboxplotSITE.png", width=20, height=9)

```

# Plot data - boxplot, all shell thickness values by site name

```{r}

mplot<- ggplot(mdata, aes(x = site_name, y = at)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(mplot)
ggsave(plot=mplot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/ATboxplotSITE.png", width=20, height=9)
```

# Plot data - ranked box plot, all condition factor by site name
```{r}

#order the sites by value

data_ordered <- mdata[order(mdata$cf),]

#create a factor with the ordered site names
data_ordered$site_name <- factor(data_ordered$site_name, levels = unique(data_ordered$site_name))

#plot with ordered site names
ranked<- ggplot(data_ordered, aes(x = site_name, y = cf)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels if needed

print(ranked)
ggsave(plot=ranked, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/CFboxplotSITEranked.png", width=20, height=9)
```

# Plot data - ranked box plot, all shell thickness by site name
```{r}

#order the sites by value

data_ordered <- mdata[order(mdata$at),]

#create a factor with the ordered site names
data_ordered$site_name <- factor(data_ordered$site_name, levels = unique(data_ordered$site_name))

#plot with ordered site names
ranked<- ggplot(data_ordered, aes(x = site_name, y = at)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels if needed

print(ranked)
ggsave(plot=ranked, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/ATboxplotSITE.png", width=20, height=9)
```

# Plot data - boxplot, all condition factor values by reporting area
```{r}

mdata$reporting_area <- as.character(mdata$reporting_area)


mplot<- ggplot(mdata, aes(x = reporting_area, y = cf)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(mplot)
ggsave(plot=mplot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/CFboxplotRA.png", width=20, height=9)

```

# Plot data - boxplot, all shell thickness values by reporting area
```{r}

mdata$reporting_area <- as.character(mdata$reporting_area)


mplot<- ggplot(mdata, aes(x = reporting_area, y = at)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(mplot)
ggsave(plot=mplot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures/ATboxplotRA.png", width=20, height=9)

```

