---
title: "031- Shell Thickness & Condition Factor Plots"
output:
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
  pdf_document: 
    fig_width: 20
    fig_height: 9
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load data
```{r}
mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphdata.csv")
Mindv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgMORPH_analytes.csv")
summed<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgMORPH_analytes2.csv")

```

# shell thickness (at)
## dumbbell - original data in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_at_range <- mdata %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_at = min(at), max_at = max(at), mean_at = mean(at))

# Order by site name
site_at_range <- site_at_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_at_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_at, xend = max_at, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(reporting_area)), size = 3) +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - reporting area with indicators of mean value (triangles)
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
site_at_range <- site_at_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_at_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean added
ggplot(site_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_at, xend = max_at, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_at), shape = 17, size = 2, color = "black") +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Asite_at_range <- mdata %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_at = min(at), max_at = max(at), mean_at = mean(at))

# Create a unique identifier for each combination of site_name and PAHgroup8
Asite_at_range <- Asite_at_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Asite_at_range <- Asite_at_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Asite_at_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Asite_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_at, xend = max_at, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_at), shape = 17, size = 2, color = "black") +
  labs(x = bquote("Shell Thickness"), y = "Site Name", title = "Shell Thickness by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}

library(ggplot2)
library(dplyr)

# Order by site names
data <- mdata %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(data, aes(x = site_name, y = at, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Site Name", y = "Shell Thickness (mm)", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +  
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1) 
  )

```

## box plot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- mdata %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = at, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Site Name", y = "Shell Thickness (mm)", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
atbox <- mdata %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

atbox <- atbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
atbox <- atbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(atbox, aes(x = site_name_ordered, y = at, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = "Site Name", y= "Shell Thickness (mm)", title = "Shell Thickness by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## correlation plot - individual analytes
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
Mcor_results <- sapply(Mindv[ , 3:ncol(Mindv)], function(x) {
  test <- cor.test(Mindv$avg_at, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
Mspearman_cor_matrix <- cor(Mindv[ , 3:ncol(Mindv)], use = "complete.obs", method = "spearman")
Mspearman_cor_avg_value <- Mspearman_cor_matrix["avg_at", ]

# Filter for significance using corr_results for p-values
Msignificant_analytes <- names(Mcor_results[Mcor_results < 0.05])

# New correlation matrix with only significant analytes
Msignificant_cor_matrix <- Mspearman_cor_matrix[Msignificant_analytes, Msignificant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(Msignificant_cor_matrix, method = "circle", type= "lower", diag= FALSE)

```

## correlation plot summed analytes
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
cor_results <- sapply(summed[ , 3:ncol(summed)], function(x) {
  test <- cor.test(summed$avg_at, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
cor_matrix <- cor(summed[ , 3:ncol(summed)], use = "complete.obs", method = "spearman")
cor_avg_value <- cor_matrix["avg_at", ]

# Filter for significance using corr_results for p-values
significant_sum <- names(cor_results[cor_results < 0.05])

# New correlation matrix with only significant analytes
significant_sum <- cor_matrix[significant_sum, significant_sum]

# Plot the correlation matrix for significant analytes
corrplot(significant_sum, method = "circle", type= "lower", diag= FALSE)

```

## correlation table - individual analytes
```{r}

library(dplyr)
library(tibble)

# Spearman correlation 
Mcor_results <- sapply(Mindv[ , 3:ncol(Mindv)], function(x) {
  test <- cor.test(Mindv$avg_at, x, method = "spearman")
  c(correlation = test$estimate, p_value = test$p.value)
})

# Convert to data frame, fix column names and remove first row of old column names
Mcor_results_df <- as.data.frame(t(Mcor_results))

colnames(Mcor_results_df) <- c("correlation", "p_value")

Mcor_results_df <- Mcor_results_df %>%
  rownames_to_column(var = "analyte")

Mcor_results_df <- Mcor_results_df %>%
  slice(-1) 

# Filter for significance and round the results
Msignificant_results_df <- Mcor_results_df %>%
  filter(p_value < 0.05)

Msignificant_results_df <- Msignificant_results_df %>%
  mutate(correlation = round(correlation, 3),
         p_value = round(p_value, 3))

# View table
print(Msignificant_results_df)

# Write to table
#write.csv(Msignificant_results_df,"/Users/cmantegna/Documents/WDFWmussels/output/tables/VIZatcorrelation.csv", row.names = FALSE)

```

## correlation table - summed analytes
```{r}

library(dplyr)
library(tibble)

# Spearman correlation tests to extract correlation coefficients and p-values
cor_results <- sapply(summed[, 4:ncol(summed)], function(x) {
  test <- cor.test(summed[, 3], x, method = "spearman")
  c(correlation = test$estimate, p_value = test$p.value)  
})

# Convert the results to a data frame
cor_results_df <- as.data.frame(t(cor_results))

colnames(cor_results_df) <- c("correlation", "p_value")

# Add analyte names as a column from rownames
cor_results_df <- cor_results_df %>%
  rownames_to_column(var = "analyte")

# Filter for significance and round the results
significant_results_df <- cor_results_df %>%
  filter(p_value < 0.05) %>%
  mutate(correlation = round(correlation, 3),
         p_value = round(p_value, 3))

# View table
print(significant_results_df)

# Write to CSV file
#write.csv(significant_results_df, "/Users/cmantegna/Documents/WDFWmussels/output/tables/VIZmorphSUMMEDcorrelation.csv", row.names = FALSE)

```

## linear regression with signficant analyte correlation
```{r}

library(ggplot2)
library(gridExtra)

Mindv$reporting_area <- as.character(Mindv$reporting_area)

# Custom colors to match other plot colors
reporting_area_colors <- c(
  "6" = "red",
  "7" = "blue",
  "8.1" = "green",
  "8.2" = "purple",
  "9" = "orange",
  "10" = "yellow",
  "11" = "brown",
  "12" = "pink",
  "13" = "grey"
  
)


# Only significant analytes
Msignificant_analytes <- Msignificant_results_df$analyte

# Create an empty list to store plots
plot_list <- list()

# Loop through each significant analyte and create a linear regression plot
for (analyte in Msignificant_analytes) {
  
  # Ensure the analyte exists in df before proceeding
  if (!analyte %in% colnames(Mindv)) {
    warning(paste("Analyte", analyte, "is not present in the data frame. Skipping..."))
    next
  }
  
  # Linear Model
  model <- lm(Mindv[[analyte]] ~ Mindv$avg_at)
  
  # Plot
p <- ggplot(Mindv, aes_string(x = "avg_at", y = analyte, color = "reporting_area")) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black", se = FALSE) +  
  scale_color_manual(values = reporting_area_colors) +  
  labs(title = paste("Linear Regression: Shell Thickness vs", analyte),
       x = "Shell Thickness (mm)",
       y = analyte) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend

# Store the plot in the list
plot_list[[analyte]] <- p

}

# Display the plots in a grid
do.call(grid.arrange, c(plot_list, ncol = 4)) 

```

## linear regression - summed analytes
```{r}

library(ggplot2)
library(gridExtra)

summed$reporting_area <- as.character(summed$reporting_area)

# Custom colors to match other plot colors
reporting_area_colors <- c(
  "6" = "red",
  "7" = "blue",
  "8.1" = "green",
  "8.2" = "purple",
  "9" = "orange",
  "10" = "yellow",
  "11" = "brown",
  "12" = "pink",
  "13" = "grey"
  
)


# Only significant analytes
significant_analytes <- significant_results_df$analyte

# Create an empty list to store plots
plot_list <- list()

# Loop through each significant analyte and create a linear regression plot
for (analyte in significant_analytes) {
  
  # Ensure the analyte exists in df before proceeding
  if (!analyte %in% colnames(summed)) {
    warning(paste("Analyte", analyte, "is not present in the data frame. Skipping..."))
    next
  }
  
  # Linear Model
  model <- lm(summed[[analyte]] ~ summed$avg_at)
  
  # Plot
p <- ggplot(summed, aes_string(x = "avg_at", y = analyte, color = "reporting_area")) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black", se = FALSE) +  
  scale_color_manual(values = reporting_area_colors) +  
  labs(title = paste("Linear Regression: Shell Thickness vs", analyte),
       x = "Shell Thickness (mm)",
       y = analyte) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend

# Store the plot in the list
plot_list[[analyte]] <- p

}

# Display the plots in a grid
do.call(grid.arrange, c(plot_list, ncol = 3)) 

```

# condition factor (cf)
## dumbbell - original data in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_cf_range <- mdata %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_cf = min(cf), max_cf = max(cf), mean_cf = mean(cf))

# Order by site name
site_cf_range <- site_cf_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_cf_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(reporting_area)), size = 3) +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - reporting area with indicators of mean value (triangles)
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
site_cf_range <- site_cf_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_cf_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean added
ggplot(site_cf_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_cf), shape = 17, size = 1, color = "black") +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Csite_at_range <- mdata %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_cf = min(cf), max_cf = max(cf), mean_cf = mean(cf))

# Create a unique identifier for each combination of site_name and PAHgroup8
Csite_at_range <- Csite_at_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Csite_at_range <- Csite_at_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Csite_at_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Csite_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_cf), shape = 17, size = 2, color = "black") +
  labs(x = bquote("Condition Factor"), y = "Site Name", title = "Condition Factor by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}

library(ggplot2)
library(dplyr)

# Order by site names
data <- mdata %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(data, aes(x = site_name, y = cf, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +  
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1) 
  )

```

## box plot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- mdata %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = cf, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
cfbox <- mdata %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

cfbox <- cfbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
cfbox <- cfbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(cfbox, aes(x = site_name_ordered, y = cf, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = "Condition Factor", title = "Condition Factor by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## correlation plot - summed analytes (no sig individual analytes)
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
Mcor_results <- sapply(summed[ , 4:ncol(summed)], function(x) {
  test <- cor.test(summed$avg_cf, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
Mspearman_cor_matrix <- cor(summed[ , 4:ncol(summed)], use = "complete.obs", method = "spearman")
Mspearman_cor_avg_value <- Mspearman_cor_matrix["avg_cf", ]

# Filter for significance using corr_results for p-values
Msignificant_analytes <- names(Mcor_results[Mcor_results < 0.05])

# New correlation matrix with only significant analytes
Msignificant_cor_matrix <- Mspearman_cor_matrix[Msignificant_analytes, Msignificant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(Msignificant_cor_matrix, method = "circle", type= "lower", diag= FALSE)

```
