---
title: "04- Pairwise Comparisons, KW + Dunn Post Hoc"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

# libraries
library(knitr)
library(tidyr)
library(dplyr)
library(vegan)
library(pgirmess)
library(ggplot2)

# global options
knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

# double check where you are
# *NOTE* see 00-data_cleaned for directory issue
getwd()

# using ALL individual samples, not averages
data<- read.csv("../data/cleaned/complete_with_groups.csv")
pdata<- read.csv("../data/cleaned/p450_complete_with_groups_for_kw.csv")
# check it out
#str(data) # checking df structure

#summary(data) #checking df contents

#colnames(data) <- tolower(colnames(data)) # if needed

```

# making groups characters
```{r}
# make analysis groups a factor (reporting area, geographic group, pcb group, site, pah group)
# change to a factor - all data
data$site_name <- as.character(data$site_name)
data$reporting_area <- as.character(data$reporting_area)
data$geo_group <- as.character(data$geo_group)
data$p16 <- as.character(data$p16)
data$p42 <- as.character(data$p42)
data$plmw <- as.character(data$plmw)
data$phmw <- as.character(data$phmw)
data$pcb <- as.character(data$pcb)

#p450 data
pdata$site_name <- as.character(pdata$site_name)
pdata$reporting_area <- as.character(pdata$reporting_area)
pdata$geo_group <- as.character(pdata$geo_group)
pdata$p16 <- as.character(pdata$p16)
pdata$p42 <- as.character(pdata$p42)
pdata$plmw <- as.character(pdata$plmw)
pdata$phmw <- as.character(pdata$phmw)
pdata$pcb <- as.character(pdata$pcb)

```

# KW & Post Hoc (individual metrics for each group)
## p450
### All to post hoc except geo_group
```{r}

kruskal.test(p450 ~ site_name, data = pdata) # p= 6.442e-6
kruskal.test(p450 ~ reporting_area, data = pdata) # p= 0.0001749
kruskal.test(p450 ~ geo_group, data = pdata) # p= 0.309
kruskal.test(p450 ~ p16, data = pdata)  # p= 0.002045
kruskal.test(p450 ~ p42, data = pdata) # p= 0.000756
kruskal.test(p450 ~ plmw, data = pdata) # p= 0.001638
kruskal.test(p450 ~ phmw, data = pdata) # p= 0.002376
kruskal.test(p450 ~ pcb, data = pdata) # p= 0.0076

```
### 10-13 reporting area significance
### 1-2 p42 significance
### 1-2 phmw significance
```{r}

mc_site<- as.data.frame(kruskalmc(p450 ~ site_name, data = pdata, method = "bonferroni"))
mc_reporting<- as.data.frame(kruskalmc(p450 ~ reporting_area, data = pdata, method = "bonferroni")) 
mc_16<- as.data.frame(kruskalmc(p450 ~ p16, data = pdata, method = "bonferroni"))
mc_42<- as.data.frame(kruskalmc(p450 ~ p42, data = pdata, method = "bonferroni"))
mc_low<- as.data.frame(kruskalmc(p450 ~ plmw, data = pdata, method = "bonferroni"))
mc_high<- as.data.frame(kruskalmc(p450 ~ phmw, data = pdata, method = "bonferroni"))
mc_pcb<- as.data.frame(kruskalmc(p450 ~ pcb, data = pdata, method = "bonferroni"))


head(mc_site)
head(mc_reporting)
head(mc_16)
head(mc_42)
head(mc_low)
head(mc_high)
head(mc_pcb)

```

## sod
### Site and PCB group to posthoc
```{r}

kruskal.test(sod ~ site_name, data = data) # p= 0.001777
kruskal.test(sod ~ reporting_area, data = data) # p= 0.3136
kruskal.test(sod ~ geo_group, data = data) # p= 0.6698
kruskal.test(sod ~ p16, data = data)  # p= 0.2081
kruskal.test(sod ~ p42, data = data) # p= 0.1482
kruskal.test(sod ~ plmw, data = data) # p= 0.1077
kruskal.test(sod ~ phmw, data = data) # p= 0.06894
kruskal.test(sod ~ pcb, data = data) # p= 0.0289

```
### No significant results 
```{r}

mc_site<- as.data.frame(kruskalmc(sod ~ site_name, data = data, method = "bonferroni"))
mc_pcb<- as.data.frame(kruskalmc(sod ~ pcb, data = data, method = "bonferroni"))

head(mc_site)
head(mc_pcb)

```

## ci1
### Only site to post hoc
```{r}

kruskal.test(ci1 ~ site_name, data = data) # p= 4.216e-5
kruskal.test(ci1 ~ reporting_area, data = data) # p= 0.1622
kruskal.test(ci1 ~ geo_group, data = data) # p= 0.386
kruskal.test(ci1 ~ p16, data = data)  # p= 0.3497
kruskal.test(ci1 ~ p42, data = data) # p= 1.3653
kruskal.test(ci1 ~ plmw, data = data) # p= 0.285
kruskal.test(ci1 ~ phmw, data = data) # p= 0.4321
kruskal.test(ci1 ~ pcb, data = data) # p= 0.1851

```
### No significant result
```{r}

mc_site<- as.data.frame(kruskalmc(ci1 ~ site_name, data = data, method = "bonferroni"))

head(mc_site)

```

## ci2
### Site and plmw to post hoc
```{r}

kruskal.test(ci2 ~ site_name, data = data) # p= 1.255e-8
kruskal.test(ci2 ~ reporting_area, data = data) # p= 0.2123
kruskal.test(ci2 ~ geo_group, data = data) # p= 0.3529
kruskal.test(ci2 ~ p16, data = data)  # p= 0.08039
kruskal.test(ci2 ~ p42, data = data) # p= 0.1298
kruskal.test(ci2 ~ plmw, data = data) # p= 0.03071
kruskal.test(ci2 ~ phmw, data = data) # p= 0.1718
kruskal.test(ci2 ~ pcb, data = data) # p= 0.1205

```

### No significant result
```{r}

mc_site<- as.data.frame(kruskalmc(ci2 ~ site_name, data = data, method = "bonferroni"))
mc_plmw<- as.data.frame(kruskalmc(sod ~ plmw, data = data, method = "bonferroni"))

head(mc_site)
head(mc_plmw)

```

## ci3
### Site, RA and geo_group to post hoc
```{r}

kruskal.test(ci3 ~ site_name, data = data) # p= 1.673e-5
kruskal.test(ci3 ~ reporting_area, data = data) # p= 0.001512
kruskal.test(ci3 ~ geo_group, data = data) # p= 0.001814
kruskal.test(ci3 ~ p16, data = data)  # p= 0.2777
kruskal.test(ci3 ~ p42, data = data) # p= 0.9086
kruskal.test(ci3 ~ plmw, data = data) # p= 0.6458
kruskal.test(ci3 ~ phmw, data = data) # p= 0.9712
kruskal.test(ci3 ~ pcb, data = data) # p= 0.813

```

### No signficant result
```{r}

mc_site<- as.data.frame(kruskalmc(ci3 ~ site_name, data = data, method = "bonferroni"))
mc_reporting<- as.data.frame(kruskalmc(ci3 ~ reporting_area, data = data, method = "bonferroni")) 
mc_geo<- as.data.frame(kruskalmc(ci3 ~ geo_group, data = data, method = "bonferroni"))



head(mc_site)
head(mc_reporting)
head(mc_geo)

```

# ANOVA 
## shell thickness
### Significant result for site (p= 5.71e-16 ), RA (p= 1.09e-6), plmw (p= 0.00751)
```{r}

library(car)
library(multcomp)

# Define grouping variables
grouping_vars <- c("site_name", "reporting_area", "geo_group", "p16", "p42", "phmw", "plmw", "pcb")

# Function to run ANOVA and post hoc test
run_anova_posthoc <- function(group_var, data) {
  
  # Run ANOVA
  anova_model <- aov(shell ~ get(group_var), data = data)
  anova_summary <- summary(anova_model)

  # Extract p-value from ANOVA table
  p_value <- anova_summary[[1]]["Pr(>F)"][1,1]  

  # Only proceed if ANOVA is significant
  if (p_value < 0.05) {
    cat("\n### Significant ANOVA results for shell thickness by", group_var, "###\n")
    print(anova_summary)

    # Run Tukey's HSD post hoc test
    posthoc_result <- TukeyHSD(anova_model)

    # Convert to data frame for filtering
    posthoc_df <- as.data.frame(posthoc_result[[1]])
    posthoc_df <- cbind(comparison = rownames(posthoc_df), posthoc_df)  # Add comparison column

    # Filter only significant results (p < 0.05)
    significant_posthoc <- posthoc_df[posthoc_df$`p adj` < 0.05, , drop = FALSE]

    if (nrow(significant_posthoc) > 0) {
      cat("\n### Significant Tukey Post Hoc Comparisons for", group_var, "###\n")
      print(significant_posthoc)
    } else {
      cat("\n### No significant post hoc differences for", group_var, "###\n")
    }

    return(list(ANOVA = anova_summary, PostHoc = significant_posthoc))
  } else {
    cat("\n### No significant ANOVA differences for shell thickness by", group_var, "###\n")
    return(NULL)
  }
}

# Run ANOVA + post hoc for each grouping variable and store only significant results
anova_results <- lapply(grouping_vars, function(var) run_anova_posthoc(var, data))
names(anova_results) <- grouping_vars

# Remove NULL results (where ANOVA was not significant)
anova_results <- anova_results[!sapply(anova_results, is.null)]

```
