---
title: "04- Spatial Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: 
    fig_width: 20
    fig_height: 9
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}
library(tinytex)
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data
Weight = mg\
Length, width, height = mm\
p450, SOD = activity/ (mg/protein)\
Condition factor, economic factor = unitless
```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarkerfull.csv")

#alldata has the site names, biomarkers, condition factor, average thickness and analyte data - each row is an individual sample
alldata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/alldata.csv")

```

## fix zero's in the data frame and alldata frame

```{r}

# Data contains 0's and must be adjusted in this order to preserve all usable data.

#sod
#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
data$SOD[data$SOD <= 0] <- 0.0025
alldata$SOD[alldata$SOD <= 0] <- 0.0025
#p450
#remove any p450 values that are 0 - those are true 0's not non-detectable. I am replacing with na so I don't lose the entire row of data, including the SOD values.
data$p450[data$p450 <= 0] <- NA
alldata$p450[alldata$p450 <= 0] <- NA

#write.csv(alldata, "/Users/cmantegna/Documents/WDFWmussels/data/alldata.csv")
```

#Impute missing p450 and condition_factor values
### Using the 'mice' package, the most common imputation method is predictive mean matching (pmm) based on the package documentation, so that is used here. 
```{r}
#install.packages("mice")
library(mice)

# Perform multiple imputation by chained equations
imputed_data <- mice(alldata[, c("p450", "condition_factor")], method = 'pmm', m = 5)  # m is the number of datasets

# Create a complete dataset from the imputations
completed_data <- complete(imputed_data, 1)  # choosing the first imputed dataset for simplicity

# Replace the original missing values
alldata$p450 <- ifelse(is.na(alldata$p450), completed_data$p450, alldata$p450)
alldata$condition_factor <- ifelse(is.na(alldata$condition_factor), completed_data$condition_factor, alldata$condition_factor)

```

```{r}
# check the data frame
#summary(alldata)
#str(alldata)
```

# K-means clustering
## This spatial analysis test puts the sites into three (because I chose 3 and not 4) clusters that are distinct from the other clusters based on averaging the data to find centroid values.
```{r}
library(dplyr)

# Prepare the data for clustering - only numeric columns should be included
clustering_data <- alldata %>%
  select(p450, SOD, condition_factor, avg_thickness) %>%
  na.omit()  # Removing rows with NA values

# Scale the data
clustering_data <- scale(clustering_data)

# Perform K-means clustering with a chosen number of clusters, let's say 3 for this example
set.seed(123)  # Setting seed for reproducibility
kmeans_result <- kmeans(clustering_data, centers = 3)

# Add the cluster assignments to the original data
alldata$cluster <- kmeans_result$cluster[match(rownames(clustering_data), rownames(alldata))]

# Now let's list the unique site names within each cluster
clustered_sites <- alldata %>%
  select(site_name, cluster) %>%
  group_by(cluster) %>%
  summarise(unique_sites = list(unique(site_name)))  # Using list to keep it all within one dataframe

print(clustered_sites)

# If you want to save the result to a CSV
#write_csv(clustered_sites, "clustered_sites.csv")

```

# Defining the centroids that define the clusters.
### these centroids define each cluster (1-3) and assign a scaled z-score that produces negative and positive values. Negative values indicate a centroid value below the mean, and a positive number indicates a centroid value above the mean.
### Cluster 1 has p450 and condition_factor values below the mean, and SOD and avg_thickness values above the mean.
### Cluster 2 has SOD values below the mean, and p450, condition_factor and avg_thickness values above the mean.
### Cluster 3 has all four values below their means.
```{r}
library(dplyr)

# Selecting and scaling the columns (standardize them because K-means is affected by scale)
data_selected <- alldata %>% 
  select(p450, SOD, condition_factor, avg_thickness) %>% 
  na.omit() %>%  # Removing any NA values to avoid errors in kmeans
  scale()  # Scale the data to standardize

# Perform K-means clustering, let's say we determine that there are 3 clusters
set.seed(123)  # For reproducibility
kmeans_result <- kmeans(data_selected, centers = 3)

# The centroids (cluster centers) are stored in the 'centers' attribute
centroids <- kmeans_result$centers

# Print the centroids
print(centroids)

```

# Defining the actual values by going through the centroid assignment in reverse
```{r}

```


# Kruskal-Wallis & kruskalmc post hoc tests to verify if there are any statistically significant differences between the clusters
### K-W shows a significant relationship between p450, SOD, condition_factor and avg_thickness and clustered sites. A post hoc test on each of these metrics will determine if the relationship is truly significant. 
```{r}
library(dplyr)
library(pgirmess)

# Perform Kruskal-Wallis test for each variable
kw_p450 <- kruskal.test(p450 ~ cluster, data = alldata)
kw_SOD <- kruskal.test(SOD ~ cluster, data = alldata)
kw_condition_factor <- kruskal.test(condition_factor ~ cluster, data = alldata)
kw_avg_thickness <- kruskal.test(avg_thickness ~ cluster, data = alldata)

# Print the Kruskal-Wallis test results
print(kw_p450)
print(kw_SOD)
print(kw_condition_factor)
print(kw_avg_thickness)

```
# p450 kruskalmc post hoc
### p450 values show a significant difference between clusters 1-2 and 2-3 but not between 1-3.
```{r}
#p450

if(kw_p450$p.value < 0.05) {
  posthoc_p450 <- kruskalmc(p450 ~ cluster, data = alldata)
  print(posthoc_p450)
}

```

# SOD kruskalmc post hoc
### SOD shows a significant difference between clusters 1-2 and 1-3 but not between 2-3.
```{r}
#SOD

if(kw_SOD$p.value < 0.05) {
  posthoc_SOD <- kruskalmc(SOD ~ cluster, data = alldata)
  print(posthoc_SOD)
}

```

# condition_factor kruskalmc post hoc
### Condition_factor shows no statistically significant difference between the clusters
```{r}
#condition_factor

if(kw_condition_factor$p.value < 0.05) {
  posthoc_condition_factor <- kruskalmc(condition_factor ~ cluster, data = alldata)
  print(posthoc_condition_factor)
}

```

# avg_thickness kruskalmc post hoc
### Avg_thickness shows a significant difference between clusters 1-3 and 2-3 but not between 1-2.
```{r}
#avg_thickness

if(kw_avg_thickness$p.value < 0.05) {
  posthoc_avg_thickness <- kruskalmc(avg_thickness ~ cluster, data = alldata)
  print(posthoc_avg_thickness)
}

```


