---
title: "040- boxplots for p450, sod, at and cf in reporting area and pah ranking"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load data
```{r}

data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/p450data.csv")
sdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/soddata.csv")
mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphdata.csv")

```

# remove significant outlier from p450
```{r}

# outlier (Sample 23) is more than 3 sd's from the mean for the set - going to remove it.
data <- data[data$sample_id != 23, ]

# confirm it's gone
data[data$sample_id == 23, ]

```

# Adjust SOD individual values to reflect LOQ limitations
```{r}

# Data contains numbers below 0 that must be adjusted.These numbers represent samples whose values were below the limit of quantification using the SOD kit. There was some activity in the raw sample, but it is too far below the limit of the standard curve to be accurately quantified.


#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
sdata$sod[sdata$sod <= 0] <- 0.0025

```

# p450
## RA - box plot - full data
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
data$reporting_area <- as.factor(data$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
data$reporting_area <- factor(data$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2
p <- ggplot(data, aes(x = reporting_area, y = p450)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "P450 Activity by Reporting Area", 
       x = "Reporting Area", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(p)

#ggsave(plot=p, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450raBoxAll.png", width=15, height=12)

```

# RA - box plot with annotation bars
```{r}

# Load necessary libraries
library(ggplot2)
library(ggsignif)  # For adding significance annotations
library(svglite)

# Convert reporting_area to factor (for proper ordering)
data$reporting_area <- as.factor(data$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
data$reporting_area <- factor(data$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2 and add significance annotations
p <- ggplot(data, aes(x = reporting_area, y = p450)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "P450 Activity by Reporting Area", 
       x = "Reporting Area", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  ) +
  
  # Add significance annotations using geom_signif
  geom_signif(comparisons = list(c("11", "12"), c("12", "13"), c("12", "8.2"), c("10", "13")),
              map_signif_level = TRUE,  # Automatically map significance levels to asterisks
              y_position = c(2.5e+07, 2.8e+07, 3.0e+07, 3.2e+07),  # Adjust these values for proper annotation heights
              tip_length = 0.02,  # Length of the lines for annotations
              textsize = 3)  # Size of the significance text

# Display the plot
print(p)

# Optionally save the plot (adjust the file path as needed)
#ggsave(plot=p, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450RA_all_with_significance.png", width=15, height=12)

```


## RA - box plot - significant pairs grid based on stats the following pairs are significantly different: 11-12, 12-13, 12-8.2, 10-13
```{r}

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(gridExtra)

# Manually specify the significant pairs (based on your description)
significant_pairs <- list(
  c("11", "12"),
  c("12", "8.2"),
  c("12", "13"),
  c("10", "13")
)

# Create an empty list to hold the plots
plot_list <- list()

# Loop through each significant pair and create a boxplot
for (i in 1:length(significant_pairs)) {
  group1 <- significant_pairs[[i]][1]
  group2 <- significant_pairs[[i]][2]
  
  # Filter the data for the two groups
  subset_data <- data %>%
    filter(reporting_area %in% c(group1, group2))
  
  # Create the boxplot for the pair with standardized y-axis
  q <- ggplot(subset_data, aes(x = reporting_area, y = p450)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    labs(title = paste(group1, "vs", group2),
         x = "Reporting Area", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
    ylim(0, 3e+07) +  # Standardize y-axis limits
    theme_minimal()
  
  # Add the plot to the list
  plot_list[[i]] <- q
  
  # Save each individual plot (adjust the file path as needed)
  ggsave(filename = paste0("plot_", group1, "_vs_", group2, ".png"), plot = p, width = 5, height = 4)
}

# Arrange all the plots in a grid (2 columns for now, can adjust as needed)
grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 2))

# Save the grid of plots as a single image (adjust the file path as needed)
#ggsave(plot=grid_plot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450raGrid.png", width=15, height=12)

```

## Rank - box plot - full data
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
data$PAHgroup8 <- as.factor(data$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
data$PAHgroup8 <- factor(data$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2
r <- ggplot(data, aes(x = PAHgroup8, y = p450)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "P450 Activity by PAH Ranking", 
       x = "PAH Ranking", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(r)

#ggsave(plot=r, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450pahBoxAll.png", width=15, height=12)
```

# Rank box plot with annotation bars
```{r}

# Load necessary libraries
library(ggplot2)
library(ggsignif)  # For adding significance annotations
library(svglite)

# Convert reporting_area to factor (for proper ordering)
data$PAHgroup8 <- as.factor(data$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
data$PAHgroup8 <- factor(data$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2 and add significance annotations
b <- ggplot(data, aes(x = PAHgroup8, y = p450)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "P450 Activity by PAH Ranking", 
       x = "PAH Ranking", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  ) +
  
  # Add significance annotations using geom_signif
  geom_signif(comparisons = list(c("1", "2")),
              map_signif_level = TRUE,  # Automatically map significance levels to asterisks
              y_position = c(2.5e+07, 2.8e+07, 3.0e+07, 3.2e+07),  # Adjust these values for proper annotation heights
              tip_length = 0.02,  # Length of the lines for annotations
              textsize = 3)  # Size of the significance text

# Display the plot
print(b)

# Optionally save the plot (adjust the file path as needed)
#ggsave(plot=b, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450pah_all_with_significance.png", width=15, height=12)

```

## Rank - box plot - significant pairs based on stats the following pairs are significantly different: 1 - 2
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Manually specify the significant pair
significant_pairs <- list(
  c("1", "2")
)

# Create an empty list to hold the plot
plot_list <- list()

# Loop through the significant pair and create a boxplot
for (i in 1:length(significant_pairs)) {
  group1 <- significant_pairs[[i]][1]
  group2 <- significant_pairs[[i]][2]
  
  # Filter the data for the two groups
  subset_data <- data %>%
    filter(PAHgroup8 %in% c(group1, group2))
  
  # Create the boxplot for the pair with standardized y-axis
  q <- ggplot(subset_data, aes(x = PAHgroup8, y = p450)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    labs(title = paste(group1, "vs", group2),
         x = "PAH Ranking", y = "P450 Activity (Unit / mg"^"-1"~"protein)") +
    theme_minimal()
  
  # Add the plot to the list (even though it's only one pair for now)
  plot_list[[i]] <- q
  
}

# If there's only one plot, print it directly
print(q)


# Save the grid of plots as a single image (adjust the file path as needed)
#ggsave(plot=q, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450pahGrid.png", width=15, height=12)

```

# sod
## RA - box plot - full data; there are no significant pairwise comparisons
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
sdata$reporting_area <- as.factor(sdata$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
sdata$reporting_area <- factor(sdata$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2
s <- ggplot(sdata, aes(x = reporting_area, y = sod)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "SOD Activity by Reporting Area", 
       x = "Reporting Area", y = "SOD Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(s)

#ggsave(plot=s, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/SODraBoxAll.png", width=15, height=12)

```

## Rank - box plot - full data
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
sdata$PAHgroup8 <- as.factor(sdata$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
sdata$PAHgroup8 <- factor(sdata$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2
r <- ggplot(sdata, aes(x = PAHgroup8, y = sod)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "SOD Activity by PAH Ranking", 
       x = "PAH Ranking", y = "SOD Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(r)

#ggsave(plot=r, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/SODpahBoxAll.png", width=15, height=12)

```

# Rank box plot with annotation bars
```{r}

# Load necessary libraries
library(ggplot2)
library(ggsignif)  # For adding significance annotations
library(svglite)

# Convert reporting_area to factor (for proper ordering)
sdata$PAHgroup8 <- as.factor(sdata$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
sdata$PAHgroup8 <- factor(sdata$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2 and add significance annotations
c <- ggplot(sdata, aes(x = PAHgroup8, y = sod)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "SOD Activity by PAH Ranking", 
       x = "PAH Ranking", y = "SOD Activity (Unit / mg"^"-1"~"protein)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  ) +
  
  # Add significance annotations using geom_signif
  geom_signif(comparisons = list(c("7", "1"), c("7", "2")),
              map_signif_level = TRUE,  # Automatically map significance levels to asterisks
              y_position = c(50, 55, 60, 65),  # Adjust these values for proper annotation heights
              tip_length = 0.02,  # Length of the lines for annotations
              textsize = 3)  # Size of the significance text

# Display the plot
print(c)

# Optionally save the plot (adjust the file path as needed)
#ggsave(plot=c, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/SODpah_all_with_significance.png", width=15, height=12)
```


## Rank - box plot - significant pairs based on stats the following pairs are significantly different: 7 - 1, 7 - 2
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(gridExtra)

# Manually specify the significant pairs (based on your description)
significant_pairs <- list(
  c("1", "7"),
  c("2", "7")
)

# Create an empty list to hold the plots
plot_list <- list()

# Loop through each significant pair and create a boxplot
for (i in 1:length(significant_pairs)) {
  group1 <- significant_pairs[[i]][1]
  group2 <- significant_pairs[[i]][2]
  
  # Filter the data for the two groups
  subset_data <- sdata %>%
    filter(PAHgroup8 %in% c(group1, group2))
  
  # Create the boxplot for the pair with standardized y-axis
  t <- ggplot(subset_data, aes(x = PAHgroup8, y = sod)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    labs(title = paste(group1, "vs", group2),
         x = "PAH Ranking", y = "SOD Activity (Unit / mg"^"-1"~"protein)") +
    theme_minimal()
  
  # Add the plot to the list
  plot_list[[i]] <- t
  
  # Save each individual plot (adjust the file path as needed)
  #ggsave(filename = paste0("plot_", group1, "_vs_", group2, ".png"), plot = p, width = 5, height = 4)
}

# Arrange all the plots in a grid (2 columns for now, can adjust as needed)
grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 2))

# Save the grid of plots as a single image (adjust the file path as needed)
#ggsave(plot=t, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/SODpahGrid.png", width=15, height=12)

```

# shell thickness
## RA - box plot - full data
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
mdata$reporting_area <- as.factor(mdata$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
mdata$reporting_area <- factor(mdata$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2
u <- ggplot(mdata, aes(x = reporting_area, y = at)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "Shell Thickness by Reporting Area", 
       x = "Reporting Area", y = "Shell Thickness (mm)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(u)

#ggsave(plot=u, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/ATraBoxAll.png", width=15, height=12)

```

# RA - box plot with annotation bars
```{r}

# Load necessary libraries
library(ggplot2)
library(ggsignif)  # For adding significance annotations
library(svglite)

# Convert reporting_area to factor (for proper ordering)
mdata$reporting_area <- as.factor(mdata$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
mdata$reporting_area <- factor(mdata$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2 and add significance annotations
d <- ggplot(mdata, aes(x = reporting_area, y = at)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "Shell Thickness by Reporting Area", 
       x = "Reporting Area", y = "Shell Thickness (mm)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  ) +
  

  # Add significance annotations using geom_signif
  geom_signif(comparisons = list(c("11", "6"), c("11", "7"), c("11", "8.1"), c("11", "10"), c("11", "12"), c("13", "6"), c("13", "6")),
              map_signif_level = TRUE,  # Automatically map significance levels to asterisks
              y_position = c(1.15, 1.20, 1.25, 1.29),  # Adjust these values for proper annotation heights
              tip_length = 0.02,  # Length of the lines for annotations
              textsize = 3)  # Size of the significance text

# Display the plot
print(d)

# Optionally save the plot (adjust the file path as needed)
ggsave(plot=d, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/atRA_all_with_significance.png", width=15, height=12)

```


## RA - box plot - significant pairs grid based on stats the following pairs are significantly different: 11-10, 11-12, 11-6, 11-7, 11-8.1, 13-12, 13-6
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(gridExtra)

# Manually specify the significant pairs (based on your description)
significant_pairs <- list(
  c("11", "6"),
  c("11", "7"),
  c("11", "8.1"),
  c("11", "10"),
  c("11", "12"),
  c("13", "6"),
  c("13", "12")
)

# Create an empty list to hold the plots
plot_list <- list()

# Loop through each significant pair and create a boxplot
for (i in 1:length(significant_pairs)) {
  group1 <- significant_pairs[[i]][1]
  group2 <- significant_pairs[[i]][2]
  
  # Filter the data for the two groups
  subset_data <- mdata %>%
    filter(reporting_area %in% c(group1, group2))
  
  # Create the boxplot for the pair with standardized y-axis
  v <- ggplot(subset_data, aes(x = reporting_area, y = at)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    labs(title = paste(group1, "vs", group2),
         x = "Reporting Area", y = "Shell Thickness (mm)") +
    theme_minimal()
  
  # Add the plot to the list
  plot_list[[i]] <- v
  
  # Save each individual plot (adjust the file path as needed)
  #ggsave(filename = paste0("plot_", group1, "_vs_", group2, ".png"), plot = v, width = 5, height = 4)
}

# Arrange all the plots in a grid (2 columns for now, can adjust as needed)
grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 2))

# Save the grid of plots as a single image (adjust the file path as needed)
#ggsave(plot=grid_plot, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/ATraGrid.png", width=15, height=12)

```

## Rank - box plot - full data
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
mdata$PAHgroup8 <- as.factor(mdata$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
mdata$PAHgroup8 <- factor(mdata$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2
w <- ggplot(mdata, aes(x = PAHgroup8, y = at)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "Shell Thickness by PAH Ranking", 
       x = "PAH Ranking", y = "Shell Thickness (mm)") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(w)

#ggsave(plot=w, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/p450pahBoxAll.png", width=15, height=12)

```

## Rank - box plot - significant pairs based on stats the following pairs are significantly different: 3-1, 3-6, 7-1
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Manually specify the significant pair
significant_pairs <- list(
  c("3", "1"),
  c("3", "6"),
  c("7", "1")
)

# Create an empty list to hold the plot
plot_list <- list()

# Loop through the significant pair and create a boxplot
for (i in 1:length(significant_pairs)) {
  group1 <- significant_pairs[[i]][1]
  group2 <- significant_pairs[[i]][2]
  
  # Filter the data for the two groups
  subset_data <- mdata %>%
    filter(PAHgroup8 %in% c(group1, group2))
  
  # Create the boxplot for the pair with standardized y-axis
  a <- ggplot(subset_data, aes(x = PAHgroup8, y = at)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    labs(title = paste(group1, "vs", group2),
         x = "PAH Ranking", y = "Shell Thickness (mm)") +
    theme_minimal()
  
  # Add the plot to the list (even though it's only one pair for now)
  plot_list[[i]] <- a
  
}

grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 2))

# Save the grid of plots as a single image (adjust the file path as needed)
#ggsave(plot=a, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/ATpahGrid.png", width=15, height=12)

```

# condition factor
## RA - box plot - full data only, no significant comparisons
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
mdata$reporting_area <- as.factor(mdata$reporting_area)

# Custom order for the reporting areas
custom_order <- c('6', '7', '8.1', '8.2', '9', '10', '11', '12', '13')
mdata$reporting_area <- factor(mdata$reporting_area, levels = custom_order)

# Create the boxplot with ggplot2
u <- ggplot(mdata, aes(x = reporting_area, y = cf)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "Condition Factor by Reporting Area", 
       x = "Reporting Area", y = "Condition Factor") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(u)

#ggsave(plot=u, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/CFraBoxAll.png", width=15, height=12)

```

## Rank - box plot - full data only, no significant comparisons
```{r}

library(ggplot2)
library(svglite)

# Convert reporting_area to factor (for proper ordering)
mdata$PAHgroup8 <- as.factor(mdata$PAHgroup8)

# Custom order for the reporting areas
custom_order <- c('1', '2', '3', '4', '5', '6', '7', '8')
mdata$PAHgroup8 <- factor(mdata$PAHgroup8, levels = custom_order)

# Create the boxplot with ggplot2
w <- ggplot(mdata, aes(x = PAHgroup8, y = cf)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(title = "Condition Factor by PAH Ranking", 
       x = "PAH Ranking", y = "Condition Factor") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),  # X-axis labels at 0 angle, centered, and larger size
    axis.text.y = element_text(size = 12),  # Y-axis label size
    axis.title.x = element_text(size = 14),  # X-axis title font size
    axis.title.y = element_text(size = 14),  # Y-axis title font size
    plot.title = element_text(hjust = 0.5, size = 16),  # Center the title and adjust its size
    panel.grid = element_blank(),  # Remove grey grid
    panel.background = element_rect(fill = "white", color = NA),  # Set background to white
    axis.line = element_line(color = "black")  # Add black axis lines
  )

# Display the plot
print(w)

#ggsave(plot=w, filename="/Users/cmantegna/Documents/WDFWmussels/output/figures_manuscript/CFpahBoxAll.png", width=15, height=12)

```