---
title: "041- updated biomarker plots"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load data
```{r}

data<- read.csv("/Users/cmantegna/Documents/Github/WDFWmussels/data/p450data.csv")
sdata<- read.csv("/Users/cmantegna/Documents/Github/WDFWmussels/data/soddata.csv")
#mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphdata.csv")

```

# remove significant outlier from p450
```{r}

# outlier (Sample 23) is more than 3 sd's from the mean for the set - going to remove it.
data <- data[data$sample_id != 23, ]

# confirm it's gone
data[data$sample_id == 23, ]

```


# P450
## Outlier adjustment, geographic grouping, box plot
```{r}

library(tidyverse)
library(ggplot2)
library(cluster)
library(vegan) 

# identify and remove outliers
iqr_threshold <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  x < lower_bound | x > upper_bound
}

data <- data %>%
  mutate(is_outlier = iqr_threshold(p450))

# after filtering out the outliers, we go from 299 data points to 285

# make geographic clusters based on lat/ long coordinates in 5 groups since it looks like there are 5 distinct areas of sites

# use k-means clustering 
set.seed(123)
clusters <- kmeans(data[, c("latitude", "longitude")], centers = 5)
data <- data %>%
  mutate(latlong_group = as.factor(clusters$cluster))

# lat/long ranges for each group
latlong_ranges <- data %>%
  group_by(latlong_group) %>%
  summarize(
    lat_range = paste0(round(min(latitude), 2), " - ", round(max(latitude), 2)),
    long_range = paste0(round(min(longitude), 2), " - ", round(max(longitude), 2))
  ) %>%
  mutate(label = paste0("Group ", latlong_group, "\nLat: ", lat_range, "\nLong: ", long_range))

# add lat/long ranges back to the df
data <- data %>%
  left_join(latlong_ranges, by = "latlong_group")

# remove outliers
filtered_data <- data %>% filter(!is_outlier)

# box plot
geographic_boxplot<- ggplot(filtered_data, aes(x = reorder(latlong_group, as.numeric(latlong_group)), y = p450)) +
  geom_boxplot(outlier.shape = NA, color = "gray50", show.legend = FALSE) + # Boxplot only
  theme_minimal() +
  theme(
    panel.grid = element_blank(), # Remove all grid lines
    panel.border = element_blank(), # Remove border
    axis.line = element_line(color = "black"), # Show only axis lines
    axis.text.x = element_blank(), # Remove x-axis text
    axis.text.y = element_text(face = "bold"), # Bold y-axis text
    axis.title = element_text(face = "bold"), # Bold axis titles
    axis.ticks.x = element_blank() # Remove x-axis ticks
  ) +
  labs(
    title = "P450 by Geographic Location",
    x = "Geographic Group",
    y = expression("P450 Activity (Unit / mg"^-1~"protein)"),
    shape = "Lat/Long Group"
  ) +
  scale_shape_manual(
    values = seq(0, length(unique(filtered_data$latlong_group)) - 1), # Unique shapes for legend
    labels = latlong_ranges$label # Group labels
  ) +
  guides(
    shape = guide_legend(override.aes = list(size = 3)) # Restore shapes in the legend
  ) +
  geom_point(
    data = data.frame(
      latlong_group = unique(filtered_data$latlong_group),
      p450 = NA
    ),
    aes(
      x = latlong_group,
      y = min(filtered_data$p450) - 0.1 * diff(range(filtered_data$p450)), # Adjust y position below plot
      shape = latlong_group
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  coord_cartesian(clip = "off") # Ensure shapes outside the plot area are visible

print(geographic_boxplot)

# save plot
#ggsave(plot=geopgraphic_boxplot, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/geographic_boxplot_p450.png", width=15, height=12)

```

## PERMANOVA
```{r}

library(tidyverse)
library(ggplot2)
library(cluster)
library(vegan) 

# PERMANOVA to determine/ confirm any statistically significant comparisons
# PERMANOVA prep: create a distance matrix (Bray-Curtis is the standard)
distance_matrix <- vegdist(data$p450, method = "euclidean")

permanova_results <- adonis2(distance_matrix ~ latlong_group, data = data)

# view
print(permanova_results)

# Function for pairwise comparisons - built by ChatGPT
pairwise.adonis2 <- function(dist, group) {
  combs <- combn(unique(group), 2, simplify = FALSE)
  results <- map_df(combs, function(pair) {
    sub_data <- group %in% pair
    ad <- adonis2(as.dist(as.matrix(dist)[sub_data, sub_data]) ~ group[sub_data])
    tibble(
      Group1 = pair[1],
      Group2 = pair[2],
      F = ad$F[1],
      R2 = ad$R2[1],
      p_value = ad$`Pr(>F)`[1]
    )
  })
  return(results)
}

# post-hoc
pairwise_permanova <- pairwise.adonis2(distance_matrix, data$latlong_group)

# view
print(pairwise_permanova)

```

## Mapping by geographic group
```{r}

library(tidyverse)
library(ggplot2)
library(cluster)
library(vegan) 

# map lat/long groups
world_map <- map_data("world")

# bound the map
map_bounds <- data %>%
  summarize(
    min_lat = min(latitude) - 1.0,
    max_lat = max(latitude) + 1.0,
    min_long = min(longitude) - 1.0,
    max_long = max(longitude) + 1.0
  )

# plot map + lat/long group with shapes
ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "gray70") +
  geom_point(data = data, aes(x = longitude, y = latitude, shape = latlong_group), size = 3) +
  scale_shape_manual(
    values = seq(0, length(unique(data$latlong_group)) - 1), # Assign unique shapes to each group
    labels = latlong_ranges$label # Add group labels
  ) +
  coord_cartesian(
    xlim = c(map_bounds$min_long, map_bounds$max_long),
    ylim = c(map_bounds$min_lat, map_bounds$max_lat)
  ) +
  theme_minimal() +
  labs(
    title = "Map of Latitude/Longitude Groups",
    x = "Longitude",
    y = "Latitude",
    shape = "Lat/Long Group"
  )

```

## Geographic grouping table
```{r}
# Create a readable table of sites included in each group
site_table <- data %>%
  select(site_name, latlong_group) %>%
  distinct() %>% # Ensure unique combinations of site_name and latlong_group
  arrange(latlong_group, site_name) %>%
  group_by(latlong_group) %>%
  summarise(
    sites = paste(site_name, collapse = ", "),
    .groups = "drop"
  )

#getwd()
# save the table
#write.csv(site_table, "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/latlong_group_sites_table.csv", row.names = FALSE)

# view
print(site_table)

```

# SOD
## Adjust SOD individual values to reflect LOQ limitations
```{r}

# Data contains numbers below 0 that must be adjusted.These numbers represent samples whose values were below the limit of quantification using the SOD kit. There was some activity in the raw sample, but it is too far below the limit of the standard curve to be accurately quantified.


#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
sdata$sod[sdata$sod <= 0] <- 0.0025

```

## Outlier adjustment, geographic grouping, box plot
```{r}

library(tidyverse)

# Use the clustering from the first dataset (data)
# clusters$centers and clusters$cluster are already calculated from your first dataset.

# Function to assign clusters from k-means centroids to new data
assign_clusters <- function(new_data, centroids) {
  # Calculate the distance of each point in `new_data` to each centroid
  distances <- as.matrix(dist(rbind(centroids, new_data[, c("latitude", "longitude")])))
  distances <- distances[-seq_len(nrow(centroids)), 1:nrow(centroids)]
  # Assign each point to the nearest centroid
  apply(distances, 1, which.min)
}

# Assign clusters from the previously calculated k-means to `sdata`
sdata <- sdata %>%
  mutate(latlong_group = as.factor(assign_clusters(sdata, clusters$centers)))

# Detect outliers in the second dataset (sdata)
iqr_threshold <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  x < lower_bound | x > upper_bound
}

sdata <- sdata %>%
  mutate(is_outlier = iqr_threshold(sod))

# remove outliers
filtered_sdata <- sdata %>% filter(!is_outlier)

# create lat/long ranges
latlong_ranges <- sdata %>%
  group_by(latlong_group) %>%
  summarize(
    lat_range = paste0(round(min(latitude), 2), " - ", round(max(latitude), 2)),
    long_range = paste0(round(min(longitude), 2), " - ", round(max(longitude), 2))
  ) %>%
  mutate(label = paste0("Group ", latlong_group, "\nLat: ", lat_range, "\nLong: ", long_range))

# add lat/long ranges back to the data
sdata <- sdata %>%
  left_join(latlong_ranges, by = "latlong_group")

# plot
geographic_boxplot_sod<- ggplot(filtered_sdata, aes(x = reorder(latlong_group, as.numeric(latlong_group)), y = sod)) +
  geom_boxplot(outlier.shape = NA, color = "gray50", show.legend = FALSE) + 
  theme_minimal() +
  theme(
    panel.grid = element_blank(), # Remove all grid lines
    panel.border = element_blank(), # Remove border
    axis.line = element_line(color = "black"), # Show only axis lines
    axis.text.x = element_blank(), # Remove x-axis text
    axis.text.y = element_text(face = "bold"), # Bold y-axis text
    axis.title = element_text(face = "bold"), # Bold axis titles
    axis.ticks.x = element_blank() # Remove x-axis ticks
  ) +
  labs(
    title = "SOD by Geographic Location",
    x = "Geographic Group",
    y = expression("SOD Activity (Unit / mg"^-1~"protein)"),
    shape = "Lat/Long Group"
  ) +
  scale_shape_manual(
    values = seq(0, length(unique(filtered_sdata$latlong_group)) - 1), # Unique shapes for legend
    labels = latlong_ranges$label # Group labels
  ) +
  guides(
    shape = guide_legend(override.aes = list(size = 3)) # Restore shapes in the legend
  ) +
  geom_point(
    data = data.frame(
      latlong_group = unique(filtered_sdata$latlong_group),
      sod = NA
    ),
    aes(
      x = latlong_group,
      y = min(filtered_sdata$sod) - 0.1 * diff(range(filtered_sdata$sod)), # Adjust y position below plot
      shape = latlong_group
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  coord_cartesian(clip = "off") # Ensure shapes outside the plot area are visible

print(geographic_boxplot_sod)

# save plot
#ggsave(plot=geopgraphic_boxplot_sod, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/geographic_boxplot_sod.png", width=15, height=12)

```

## PERMANOVA
```{r}

# Create a distance matrix for SOD data
distance_matrix_sod <- vegdist(sdata$sod, method = "euclidean")

permanova_results_sod <- adonis2(distance_matrix_sod ~ latlong_group, data = sdata)

# view
print(permanova_results_sod)

# post-hoc
# NO significant pairwise comparisons
pairwise_permanova_sod <- pairwise.adonis2(distance_matrix_sod, sdata$latlong_group)

# view
print(pairwise_permanova_sod)

# Map of SOD lat/long groups
# bound the map
map_bounds_sod <- sdata %>%
  summarize(
    min_lat = min(latitude) - 1.0,
    max_lat = max(latitude) + 1.0,
    min_long = min(longitude) - 1.0,
    max_long = max(longitude) + 1.0
  )

# map
ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "gray70") +
  geom_point(data = sdata, aes(x = longitude, y = latitude, color = latlong_group), size = 3) +
  scale_color_discrete(labels = latlong_ranges$label) +
  coord_cartesian(
    xlim = c(map_bounds_sod$min_long, map_bounds_sod$max_long),
    ylim = c(map_bounds_sod$min_lat, map_bounds_sod$max_lat)
  ) +
  theme_minimal() +
  labs(
    title = "Map of Latitude/Longitude Groups for SOD",
    x = "Longitude",
    y = "Latitude",
    color = "Lat/Long Group"
  )

# table of sites included in each SOD group
site_table_sod <- sdata %>%
  select(site_name, latlong_group) %>%
  arrange(latlong_group, site_name) %>%
  group_by(latlong_group) %>%
  summarise(sites = paste(unique(site_name), collapse = ", "))

# view
print(site_table_sod)

```
# P450
## Analyte Plots
## PCBs
```{r}

library(ggplot2)

avg_data<- read.csv("/Users/cmantegna/Documents/Github/WDFWmussels/data/avg_all_pcb.csv")

# Example: Your data frame is called 'data', and you want to plot 'Analyte_X' vs 'Biomarker'
a23_p450<- ggplot(avg_data, aes(x = a23, y = avg_p450)) +
  geom_point() + # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "blue") + # Linear regression line with confidence interval
  labs(
    title = "Analyte_PCB_a23 vs Biomarker Activity",
    x = "Analyte Concentration (ng/g tissue, dry weight)",
    y = "Biomarker Activity"
  ) +
  theme_minimal()

ggsave(plot=a23_p450, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/a23_p450.png", width=15, height=12)

```

# SOD
## Analyte Plots
## PCBs
```{r}

library(ggplot2)

#avg_data<- read.csv("/Users/cmantegna/Documents/Github/WDFWmussels/data/avg_all_pcb.csv")

# Example: Your data frame is called 'data', and you want to plot 'Analyte_X' vs 'Biomarker'
a23_sod<- ggplot(avg_data, aes(x = a23, y = avg_sod)) +
  geom_point() + # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "blue") + # Linear regression line with confidence interval
  labs(
    title = "Analyte_PCB_a23 vs Biomarker Activity",
    x = "Analyte Concentration (ng/g tissue, dry weight)",
    y = "Biomarker Activity"
  ) +
  theme_minimal()

print(a23_sod)
#ggsave(plot=a23_sod, filename="/Users/cmantegna/Documents/Github/WDFWmussels/output/figures_manuscript/a23_sod.png", width=15, height=12)

```

## Analyte plots - all individuals loop
```{r}

avg_sdata<- read.csv("/Users/cmantegna/Documents/Github/WDFWmussels/data/sodcorrelationMAN2.csv")

# Save all analyte names
analyte_names2 <- colnames(avg_sdata)[4:15] 

# Create a folder to save plots
getwd()
#dir.create("/Users/cmantegna/Documents/GitHub/WDFWmussels/output/analyte_plots")

# Loop through analytes
for (analyte in analyte_names2) {
  p <- ggplot(avg_sdata, aes_string(x = analyte, y = "avg_sod")) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = paste(analyte, "vs SOD Activity 2"),
      x = paste(analyte, "Concentration"),
      y = "SOD Activity"
    ) +
    theme_minimal()
  
  # Save the plot
  ggsave(filename = paste0("analyte_plots/", analyte, "_v_SOD2.png"), plot = p, width = 6, height = 4)
}

```

