---
title: "07- Correlations with Averaged Data, n=74"
subtitle: "Move to code file 10, this is no longer applicable"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

# libraries
library(knitr)
library(tidyr)
library(dplyr)
library(vegan)
library(pgirmess)
library(ggplot2)
library(FSA)           
library(rstatix)       
library(car)
library(RVAideMemoire)
library(rcompanion)

# global options
knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

getwd()

# ALL individual samples
#df1<- read.csv("../data/cleaned/all_samples_all_groups.csv")
#pdata<- read.csv("../data/cleaned/p450_samples_all_groups.csv")
# Averaged
avgdata<- read.csv("../data/cleaned/avg_samples_all_groups.csv") #all data + groups without contaminants data
avgsummed<-  read.csv("../data/cleaned/avg_all_summed.csv")
avgpcb<-  read.csv("../data/cleaned/avg_all_pcb.csv")
avgpbde<-  read.csv("../data/cleaned/avg_all_pbde.csv")
avgmetal<-  read.csv("../data/cleaned/avg_all_metal.csv")
avglmw<-  read.csv("../data/cleaned/avg_all_lmw.csv")
avghmw<-  read.csv("../data/cleaned/avg_all_hmw.csv")
avgddt<-  read.csv("../data/cleaned/avg_all_ddtPesticide.csv")
avghch<-  read.csv("../data/cleaned/avg_all_chlordanesHch.csv")
avgall<-  read.csv("../data/cleaned/avg_all_analytes_individual.csv")


```

# Spearman Correlations
## Individual analytes - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:150

# Convert grouping metrics to factors
df[, grouping_cols] <- lapply(df[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df[, biomarker_cols]
analyte_data <- df[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0("Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

# Individual analytes by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:150

# Convert grouping metrics to factors
df[, grouping_cols] <- lapply(df[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df[, biomarker_cols])
analytes <- colnames(df[, analyte_cols])
grouping_vars <- colnames(df[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df[df[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0("Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```


## avgDDT - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df1 <- avgddt  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:31

# Convert grouping metrics to factors
df1[, grouping_cols] <- lapply(df1[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df1[, biomarker_cols]
analyte_data <- df1[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df1[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0("Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgDDT by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:31

# Convert grouping metrics to factors
df1[, grouping_cols] <- lapply(df1[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df1[, biomarker_cols])
analytes <- colnames(df1[, analyte_cols])
grouping_vars <- colnames(df1[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df1[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df1[df1[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0("Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgHCH - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df2 <- avghch  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:30

# Convert grouping metrics to factors
df2[, grouping_cols] <- lapply(df2[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df2[, biomarker_cols]
analyte_data <- df2[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df2[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" HCH Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgHCH by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:30

# Convert grouping metrics to factors
df2[, grouping_cols] <- lapply(df2[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df2[, biomarker_cols])
analytes <- colnames(df2[, analyte_cols])
grouping_vars <- colnames(df2[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df2[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df2[df2[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0("HCH Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgHMW - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df3 <- avghmw  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:39

# Convert grouping metrics to factors
df3[, grouping_cols] <- lapply(df3[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df3[, biomarker_cols]
analyte_data <- df3[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df3[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" HMW Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgHMW by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:39

# Convert grouping metrics to factors
df3[, grouping_cols] <- lapply(df3[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df3[, biomarker_cols])
analytes <- colnames(df3[, analyte_cols])
grouping_vars <- colnames(df3[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df3[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df3[df3[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" HMW Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgLMW - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df4 <- avglmw  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:51

# Convert grouping metrics to factors
df4[, grouping_cols] <- lapply(df4[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df4[, biomarker_cols]
analyte_data <- df4[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df4[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" LMW Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgLMW by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:51

# Convert grouping metrics to factors
df4[, grouping_cols] <- lapply(df4[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df4[, biomarker_cols])
analytes <- colnames(df4[, analyte_cols])
grouping_vars <- colnames(df4[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df4[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df4[df1[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" LMW Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgMetal - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df5 <- avgmetal  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:26

# Convert grouping metrics to factors
df5[, grouping_cols] <- lapply(df5[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df5[, biomarker_cols]
analyte_data <- df5[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df5[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" Metal Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgMetal by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:26

# Convert grouping metrics to factors
df5[, grouping_cols] <- lapply(df5[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df5[, biomarker_cols])
analytes <- colnames(df5[, analyte_cols])
grouping_vars <- colnames(df5[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df5[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df5[df5[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" Metal Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgPBDE - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df6 <- avgpbde  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:36

# Convert grouping metrics to factors
df6[, grouping_cols] <- lapply(df6[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df6[, biomarker_cols]
analyte_data <- df6[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df6[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" PBDE Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgPBDE by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:36

# Convert grouping metrics to factors
df6[, grouping_cols] <- lapply(df6[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df6[, biomarker_cols])
analytes <- colnames(df6[, analyte_cols])
grouping_vars <- colnames(df6[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df6[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df6[df6[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" PBDE Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgPCB - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df7 <- avgpcb  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:62

# Convert grouping metrics to factors
df7[, grouping_cols] <- lapply(df7[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df7[, biomarker_cols]
analyte_data <- df7[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df7[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" PCB Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgPCB by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:62

# Convert grouping metrics to factors
df7[, grouping_cols] <- lapply(df7[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df7[, biomarker_cols])
analytes <- colnames(df7[, analyte_cols])
grouping_vars <- colnames(df7[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df7[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df7[df7[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" PCB Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```

## avgSummed - n= 74
```{r}

# Load required package
library(Hmisc)

# Define dataset
df8 <- avgsummed  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:31

# Convert grouping metrics to factors
df8[, grouping_cols] <- lapply(df8[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarker_data <- df8[, biomarker_cols]
analyte_data <- df8[, analyte_cols]

# Get biomarker and analyte names
biomarkers <- colnames(biomarker_data)
analytes <- colnames(analyte_data)

# Initialize storage for results
results_list <- list()

# Run Spearman correlation for each biomarker against all analytes
for (biomarker in biomarkers) {
  spearman_result <- rcorr(as.matrix(df8[[biomarker]]), as.matrix(analyte_data), type = "spearman")
  
  # Extract correlation coefficients and p-values
  corr_matrix <- spearman_result$r[, 1, drop = FALSE]  # Ensure it's a column matrix
  p_matrix <- spearman_result$P[, 1, drop = FALSE]  # Ensure p-values align properly

  # Convert matrices to data frames
  corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
  p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

  # Merge correlation values with p-values
  merged_df <- merge(corr_df, p_df, by = "Analyte")

  # Add the correct biomarker label
  merged_df$Biomarker <- biomarker  # Ensures proper alignment

  # Reorder columns
  merged_df <- merged_df[, c("Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

  # Filter significant results (p < 0.05)
  significant_results <- merged_df[merged_df$P_Value < 0.05, ]

  # Store results
  results_list[[biomarker]] <- significant_results

  # Save as CSV
  write.csv(significant_results, paste0(" Summed Spearman_Significant_", biomarker, ".csv"), row.names = FALSE)
}

# Print message
message("Spearman correlation complete. Significant results saved.")

```

## avgSummed by groups - n= 9 groups
```{r}

# Load required package
library(Hmisc)
library(dplyr)

# Define dataset
#df <- avgall  # Ensure avgall is your dataset

# Define column ranges
biomarker_cols <- 4:9
grouping_cols <- 10:20
analyte_cols <- 21:31

# Convert grouping metrics to factors
df8[, grouping_cols] <- lapply(df8[, grouping_cols], as.factor)

# Extract biomarker and analyte data
biomarkers <- colnames(df8[, biomarker_cols])
analytes <- colnames(df8[, analyte_cols])
grouping_vars <- colnames(df8[, grouping_cols])

# Run Spearman correlation within each group
for (group_var in grouping_vars) {
  message("Running Spearman correlation for group: ", group_var)
  
  # Split data by the grouping variable
  group_levels <- unique(df8[[group_var]])

  # Initialize storage
  group_results <- list()

  for (level in group_levels) {
    subset_df <- df8[df8[[group_var]] == level, ]  # Filter for group level
    
    # Skip if too few observations
    if (nrow(subset_df) < 5) next
    
    for (biomarker in biomarkers) {
      spearman_result <- rcorr(as.matrix(subset_df[[biomarker]]), as.matrix(subset_df[, analyte_cols]), type = "spearman")
      
      # Extract correlation coefficients and p-values
      corr_matrix <- spearman_result$r[, 1, drop = FALSE]
      p_matrix <- spearman_result$P[, 1, drop = FALSE]

      # Convert matrices to data frames
      corr_df <- data.frame(Analyte = rownames(corr_matrix), Spearman_Correlation = corr_matrix[, 1], row.names = NULL)
      p_df <- data.frame(Analyte = rownames(p_matrix), P_Value = p_matrix[, 1], row.names = NULL)

      # Merge correlation values with p-values
      merged_df <- merge(corr_df, p_df, by = "Analyte")
      merged_df$Biomarker <- biomarker  # Ensure correct biomarker labeling
      merged_df$Group <- group_var
      merged_df$Level <- level

      # Reorder columns
      merged_df <- merged_df[, c("Group", "Level", "Biomarker", "Analyte", "Spearman_Correlation", "P_Value")]

      # Filter significant results (p < 0.05)
      significant_results <- merged_df[merged_df$P_Value < 0.05, ]

      # Store results
      group_results[[paste0(biomarker, "_", level)]] <- significant_results
    }
  }

  # Combine all results for the group and save as CSV
  if (length(group_results) > 0) {
    final_results <- do.call(rbind, group_results)
    write.csv(final_results, paste0(" Summed Spearman_Grouped_", group_var, ".csv"), row.names = FALSE)
  }
}

message("Grouped Spearman correlation complete. Significant results saved.")


```
