---
title: "09- Creating Analysis Groups Based on Biomarker Values"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

# libraries
library(knitr)
library(tidyr)
library(tidyverse)
library(dplyr)
library(vegan)
library(cluster)
library(pgirmess)
library(ggplot2)
library(factoextra)
library(FSA)           
library(rstatix)       
library(car)
library(RVAideMemoire)
library(rcompanion)

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

# double check where you are
# *NOTE* see 00-data_cleaned for directory issue
getwd()

# load data
bio_df<- read.csv("../data/raw_complete_no_groups.csv")

# check it out
str(bio_df) # checking df structure
summary(bio_df) #checking df contents

```

```{r }

# replace any values below LOQ with imputed LOQ (.5 x LOD), LOD= 0.05
bio_df$sod[bio_df$sod <= 0] <- 0.0025

# check
summary(bio_df)

```

# Clustering based on p450, sod, and shell thickness
```{r}

# Basic setup
biomarker_data <- bio_df[, 7:8]  # adjust if needed
rownames(biomarker_data) <- bio_df$sample_id  # optional but useful

# remove NAs from p450
bio_clust_df <- bio_df %>%
  filter(!is.na(p450)) %>%
  select(sample_id, p450, sod, shell)

# Scale the biomarkers
scaled_biomarkers <- scale(bio_clust_df[, c("p450", "sod")])

# 2. Run k-means with k = 3 and k = 4
set.seed(42)
kmeans_3 <- kmeans(scaled_biomarkers, centers = 3, nstart = 25)
kmeans_4 <- kmeans(scaled_biomarkers, centers = 4, nstart = 25)

# 3. Add both cluster assignments to bio_clust_df
bio_clust_df$k3_cluster <- as.factor(kmeans_3$cluster)
bio_clust_df$k4_cluster <- as.factor(kmeans_4$cluster)

# 4. Merge into main bio_df using sample_id
bio_df <- bio_df %>%
  left_join(bio_clust_df[, c("sample_id", "k3_cluster", "k4_cluster")], by = "sample_id")

```

# Comparing clusters, box plots
```{r}

# p450
ggplot(bio_clust_df, aes(x = k3_cluster, y = p450, fill = k3_cluster)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "p450 by Cluster")

ggplot(bio_clust_df, aes(x = k4_cluster, y = p450, fill = k4_cluster)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "p450 by Cluster")

# sod
ggplot(bio_clust_df, aes(x = k3_cluster, y = sod, fill = k3_cluster)) +
  geom_boxplot()

ggplot(bio_clust_df, aes(x = k4_cluster, y = sod, fill = k4_cluster)) +
  geom_boxplot()

# shell
ggplot(bio_clust_df, aes(x = k3_cluster, y = shell, fill = k3_cluster)) +
  geom_boxplot()

ggplot(bio_clust_df, aes(x = k4_cluster, y = shell, fill = k4_cluster)) +
  geom_boxplot()

```

# Comparing Clusters, PCA
```{r}

library("ggfortify")

pca3 <- prcomp(scaled_biomarkers)

autoplot(pca3, data = bio_clust_df, colour = 'k3_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of Biomarkers, 3")

pca4 <- prcomp(scaled_biomarkers)

autoplot(pca4, data = bio_clust_df, colour = 'k4_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of Biomarkers, 4")

```

# UMAP- Uniform Manifold Approximation and Projection
## preserves local and global data structure, important for distance matrices and cluster proximity
```{r}

#install.packages("umap")
library("umap")

set.seed(42)
umap_res <- umap(scaled_biomarkers)

# Add UMAP dimensions to your df
bio_clust_df$UMAP1 <- umap_res$layout[,1]
bio_clust_df$UMAP2 <- umap_res$layout[,2]


ggplot(bio_clust_df, aes(x = UMAP1, y = UMAP2, color = k3_cluster)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "UMAP Projection of Biomarker Data (k = 3)")

ggplot(bio_clust_df, aes(x = UMAP1, y = UMAP2, color = k4_cluster)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "UMAP Projection of Biomarker Data (k = 4)")

ggplot(bio_clust_df, aes(x = UMAP1, y = UMAP2, color = shell)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "UMAP of Biomarkers Colored by Shell Thickness")


```

# Looking into the shell thickness v biomarker result
```{r}

# make a detox to stress ratio
bio_df <- bio_df %>%
  mutate(detox_stress_ratio = p450 / sod)

# make a stress index
bio_df <- bio_df %>%
  mutate(stress_index = rowMeans(scale(select(., p450, sod)), na.rm = TRUE))

# plot them against shell thickness
ggplot(bio_df, aes(x = detox_stress_ratio, y = shell)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Shell Thickness vs. Detox-to-Stress Ratio")

ggplot(bio_df, aes(x = stress_index, y = shell)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Shell Thickness vs. Stress Index")

# run a quick model to investigate
summary(lm(shell ~ detox_stress_ratio, data = bio_df))
summary(lm(shell ~ stress_index, data = bio_df))


```

# non-parametric model
```{r}

library(mgcv)
gam_model <- gam(shell ~ s(stress_index), data = bio_df)
summary(gam_model)
plot(gam_model)

```

# Shell as a co-variate
```{r}

ggplot(bio_df, aes(x = shell, y = sod, color = k4_cluster.x)) +
  geom_point() + facet_wrap(~ k4_cluster.x) +
  geom_smooth(method = "lm")


```

# PCA for stress/ detox and stress indices
```{r}

# Example ratios
bio_clust_df$p450_sod_ratio <- bio_clust_df$p450 / bio_clust_df$sod
bio_clust_df$stress_index <- rowMeans(scaled_biomarkers[, c("p450", "sod")], na.rm = TRUE)

# Re-run PCA or clustering with new composite measures included
composite_df <- scale(bio_clust_df[, c("p450_sod_ratio", "stress_index")])

# stress by 3 and 4 clusters
pca_stress3 <- prcomp(composite_df)

autoplot(pca_stress3, data = bio_clust_df, colour = 'k3_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of Stress, 3")

pca_stress4 <- prcomp(composite_df)

autoplot(pca_stress4, data = bio_clust_df, colour = 'k4_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of Stress, 4")

# detox/ stress ration by 3 and 4 clusters
pca_ds3 <- prcomp(composite_df)

autoplot(pca_ds3, data = bio_clust_df, colour = 'k3_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of D/S, 3")

pca_ds4 <- prcomp(composite_df)

autoplot(pca_ds4, data = bio_clust_df, colour = 'k4_cluster', frame = TRUE) +
  theme_minimal() +
  labs(title = "PCA of D/S, 4")


```

