---
title: "09- Data Cleaning Biomarkers and Analytes"
output:
  
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 12
  pdf_document: 
    fig_width: 10
    fig_height: 7
---

# Setup

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 12,       # Set plot width in inches
  fig.height = 8,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r, eval= FALSE}
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

### Note:

For *data* the units are listed below. Weight = g\
Length, width, height = mm\
p450, SOD = activity/ (mg/protein)\
Condition factor, economic factor = unitless\
For *pah*, *indv*, and *allana* the units are ng/g\
For *metal* the units are mg/kg

```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarker_thickness_condition.csv")

#pah has complete site values and different summed pah analyte whole tissue values 
pah<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_pah_analytes.csv")

#indv has complete site values and individual named pah analyte whole tissue values 
#indv<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_analytes.csv") 

metal<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/metal.csv")

all<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/all_analytes.csv")

```
# fix values
```{r}
# Data contains 0's and must be adjusted in this order to preserve all usable data.

#sod
#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
data$SOD[data$SOD <= 0] <- 0.0025

#p450
#remove any p450 values that are 0 - those are true 0's not non-detectable. I am replacing with na so I don't lose the entire row of data, including the SOD values.
data$p450[data$p450 <= 0] <- NA
```

#data for basic stats table for biomarkers, condition factor and shell thickness
```{r}
# Convert 'condition_factor' and 'avg_thickness' to numeric
data$condition_factor <- as.numeric(data$condition_factor)
data$avg_thickness <- as.numeric(data$avg_thickness)

```

#overall data stats
```{r}
library(dplyr)

overall_summary <- data %>%
  summarise(across(c(p450, SOD, condition_factor, avg_thickness),
                   list(
                     median = ~median(.x, na.rm = TRUE),
                     lower_bound = ~quantile(.x, probs = 0.025, na.rm = TRUE),
                     upper_bound = ~quantile(.x, probs = 0.975, na.rm = TRUE),
                     min_value = ~min(.x, na.rm = TRUE),
                     max_value = ~max(.x, na.rm = TRUE)
                   ), .names = "{.col}_{.fn}"))

```

```{r}
library(dplyr)

grouped_summary_with_sd <- data %>%
  group_by(site_name) %>%
  summarise(across(c(p450, SOD, condition_factor, avg_thickness),
                   list(
                     median = ~median(.x, na.rm = TRUE),
                     lower = ~quantile(.x, probs = 0.025, na.rm = TRUE),
                     upper = ~quantile(.x, probs = 0.975, na.rm = TRUE),
                     sd = ~sd(.x, na.rm = TRUE)
                   ), .names = "{.col}_{.fn}"))

```

# trying to get SD for all of the values, not just by site.
```{r}
overall_summary_with_sd <- data %>%
  summarise(across(c(p450, SOD, condition_factor, avg_thickness),
                   list(
                     median = ~median(.x, na.rm = TRUE),
                     lower_bound = ~quantile(.x, probs = 0.025, na.rm = TRUE),
                     upper_bound = ~quantile(.x, probs = 0.975, na.rm = TRUE),
                     min_value = ~min(.x, na.rm = TRUE),
                     max_value = ~max(.x, na.rm = TRUE),
                     sd = ~sd(.x, na.rm = TRUE)
                   ), .names = "{.col}_{.fn}"))

```

# manova for biomarkers, sites, condition_factor and avg_thickness
```{r}

data2<- grouped_summary_with_sd
#if(!require(car)) install.packages("car")
library(car)

# Your dependent variables (biomarkers)
dv <- cbind(data$p450, data$SOD)

# Running MANOVA
manova_result <- manova(dv ~ condition_factor + avg_thickness + site_name, data = data)

# View the summary of the MANOVA results
summary(manova_result, test = "Wilks")

```
# trying the lme
```{r}
install.packages("Matrix")
install.packages("lme4")

library(lme4)

# Example model for Biomarker1
model_p450 <- lmer(p450 ~ condition_factor +avg_thickness + (1|site_name), data = data)

# You can inspect the model summary
summary(model_p450)

```
# sod 
```{r}
library(lme4)

# Example model for Biomarker1
model_SOD <- lmer(SOD ~ condition_factor +avg_thickness + (1|site_name), data = data)

# You can inspect the model summary
summary(model_SOD)
```


# adjusting biomarker values to remove true 0's and replace low values with half of detection limit.

```{r}
library(dplyr)

#merge data and all for per site review 
merged_df <- merge(data, all, by = c("site_name"), all = TRUE)
merged_df2<- merge(merged_df, pah, by= c("site_name"), all= TRUE)
merged_df3<- merge(grouped_summary_with_sd, all, by= c("site_name"), all= TRUE)

head(merged_df)
head(merged_df2)
head(merged_df3)
```

```{r}
summary(merged_df2)
```

```{r}
library(reshape2)

rownames(merged_df3)[rownames(merged_df3) == "Broad Spit (Fisherman's Point)" ] <- "Broad Spit"
rownames(merged_df3)[rownames(merged_df3) == "Penrose Point State Park" ] <- "Pennrose Point State Park"
rownames(merged_df3)[rownames(merged_df3) == "Maristone Island Bridge" ] <- "Maristone Island"
rownames(merged_df3)[rownames(merged_df3) == "Suquamish, Stormwater Outfall" ] <- "Suquamish Stormwater Outfall"


man_df <- dcast(merged_df3, site_name + p450_median + SOD_median + condition_factor_median + avg_thickness_median ~ all_analyte, value.var = "DryValue")
head(man_df)


#all_analyte_df <- dcast(merged_df, DryValue + site_name + site_number +latitude + longitude + p450 + SOD + condition_factor + avg_thickness ~ all_analyte, value.var = "DryValue")

#head(all_analyte_df)

#all_pah_df <- dcast(merged_df2, site_name + site_number +latitude + longitude + p450 + SOD + condition_factor + avg_thickness  ~ pah_analyte, value.var = "DryValue.y")

#head(all_pah_df)

write.csv(man_df, "man_df.csv", row.names = FALSE)
```
#running the MANOVA with all_analytes (not indidivual PAH analytes)
```{r}
library(car)

man_df <- man_df[-c(20, 52, 53, 54), ]

# Your dependent variables (biomarkers)
dv <- cbind(data$p450, data$SOD)

# Running MANOVA
manova_result <- manova(dv ~ site_name + arsenic + cadmium + copper + lead + mercuryTotal + SumBDEs + SumCHLDs + SumDDTs + SumHCHs + SumPAHs16 + SumPAHsHMW + SumPCBs2x17 + Zinc, data = man_df)

# View the summary of the MANOVA results
summary(manova_result, test = "Wilks")
```


# adjusting p450 to only include values in the 2.5 - 97.5 percentile

```{r}
#find quartile values
# Calculate the lower and upper percentiles (2.5th and 97.5th)
lower_limit <- quantile(all_pah_df$p450, probs = 0.025, na.rm= TRUE)
upper_limit <- quantile(all_pah_df$p450, probs = 0.975, na.rm= TRUE)

# Print the lower and upper limits
print(lower_limit)
print(upper_limit)


```

#replace values outside of the quartiles with NA

```{r}
all_pah_df$p450[all_pah_df$p450 <= 1002088] <- NA
all_pah_df$p450[all_pah_df$p450 >= 10831147] <- NA


```

# p450 correlation data frame preparation

```{r}
all_columns <- names(all_pah_df)

# Remove the columns you don't want to include in the model
excluded_columns <- c('latitude', 'longitude', 'site_name', 'site_number', 'SOD', 'NA')
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Enclose each column name in backticks to handle special characters
independent_columns <- sapply(independent_columns, function(x) paste0("`", x, "`"))

# Create a string representing the formula
formula_str <- paste("p450 ~", paste(independent_columns, collapse = " + "))

# Convert the string to a formula object
formula <- as.formula(formula_str)

```

#95% p450 against the analytes

```{r}
library(corrplot)
library(dplyr)

#only assessing the values between 2.5 - 97.5 percentiles of the p450 data
 

# Convert columns to numeric, handling NA values properly
all_pah_df <- all_pah_df %>%
  mutate(
    condition_factor = as.numeric(as.character(condition_factor)),
    avg_thickness = as.numeric(as.character(avg_thickness))
  )


# Extract variable names from the formula
variables <- all.vars(formula)

# Subset the dataframe 'sod_all' using the extracted variables
subset_data <- all_pah_df[, variables]

complete_cases <- subset_data[complete.cases(subset_data), ]
if (nrow(complete_cases) > 1) {
  correlation <- cor(complete_cases, method = "pearson")
} else {
  print("Insufficient complete cases for correlation calculation.")
}


# Compute Pearson correlation for each pair of variables
correlation_results <- cor(subset_data, method = "pearson", use = "pairwise.complete.obs")

# View the correlation matrix
print(correlation_results)

corrplot(correlation_results, method = "shade", type = "upper", tl.col = "black", tl.srt = 90)

```

# pca of those analytes and p450

```{r}
# PCA Plot with biomarkers
library('FactoMineR')
library("factoextra")

all_columns <- names(all_pah_df)

# Remove the columns you don't want to include in the model
excluded_columns <- c('latitude', 'longitude', 'site_name', 'site_number', 'SOD', 'NA')
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Remove NAs from the dataset
df_clean <- na.omit(all_pah_df)

library(FactoMineR)
library(factoextra)  # For PCA visualization

# Use the list 'independent_columns' to subset 'df_clean'
pca_data <- df_clean[, independent_columns]

# Perform PCA on the subset of data
pca_result <- PCA(pca_data, scale.unit = TRUE, ncp = 5, graph = FALSE)

# Visualizations (Optional but helpful for interpreting PCA results)
# Eigenvalues - to see the importance of each principal component
fviz_eig(pca_result)

# Individuals - to see the distribution of samples
fviz_pca_ind(pca_result, 
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             title = "PCA - Individuals (colored by cos2)",
             repel = TRUE) # Avoid text overlapping

# Variables - to see how variables contribute to the principal components
fviz_pca_var(pca_result,
             col.var = "contrib", # Color by contribution to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             title = "PCA - Variables (colored by contribution)",
             repel = TRUE) # Avoid text overlapping


# Selecting the relevant variables for PCA
#pca_data <- df_clean[, independent_columns]

# Performing PCA
#pca_res <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Plotting the PCA
#pcaplot<- fviz_pca_biplot(pca_res, label = "var", col.var = "contrib",
             #   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             #   repel = TRUE)  # Avoid text overlapping (slow if many points)



```
#pearson correlation values
```{r}


```


#glm

```{r}
all_columns <- names(all_pah_df)

# Remove the columns you don't want to include in the model
excluded_columns <- c('latitude', 'longitude', 'site_name', 'site_number', 'SOD', 'NA')
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Enclose each column name in backticks to handle special characters
independent_columns <- sapply(independent_columns, function(x) paste0("`", x, "`"))

# Create a string representing the formula
formula_str <- paste("p450 ~", paste(independent_columns, collapse = " + "))

# Convert the string to a formula object
formula <- as.formula(formula_str)

# Install and load the ppcor package
install.packages("ppcor")
library(ppcor)

library(ppcor)
library(car) # For vif() function

# Assume independent_columns, all_pah_df, and response_var are defined as before

for (predictor in independent_columns) {
  # Variables to potentially control for
  potential_controls <- independent_columns[!independent_columns %in% predictor]
  
  # Check VIF to address multicollinearity among control variables
  # Temporarily include the predictor of interest to calculate VIF
  temp_model <- lm(as.formula(paste(response_var, "~", paste(potential_controls, collapse = "+"), "+", gsub("`", "", predictor))), data = all_pah_df)
  vif_values <- vif(temp_model)
  
  # Exclude variables with high VIF
  low_vif_controls <- names(vif_values)[vif_values < 5] # Choosing 5 as a threshold is somewhat arbitrary; adjust as needed
  
  # Ensure there's enough data for the calculation
  if (length(low_vif_controls) < 2) {
    cat("Not enough control variables for partial correlation between", response_var, "and", predictor, "\n")
    next
  }
  
  # Attempt partial correlation with reduced set of controls
  temp_df <- all_pah_df[, c(response_var, gsub("`", "", predictor), gsub("`", "", low_vif_controls))]
  tryCatch({
    temp_result <- pcor.test(temp_df[,1], temp_df[,2], temp_df[,-c(1,2)])
    results_list[[gsub("`", "", predictor)]] <- temp_result
  }, error = function(e) {
    cat("Error in partial correlation between", response_var, "and", predictor, ": ", e$message, "\n")
  })
}

# Proceed as before to inspect results



glm_model <- glm(formula, family = poisson(), data = all_pah_df)
print(summary(glm_model))

lm_model<- lm(formula, data = all_pah_df)
print(summary(lm_model))
```

# trying a General Additive Model to see if any of the analytes is skewing the relationship

```{r}
library(mgcv)

#rename columns with characters in them
colnames(all_pah_df)[colnames(all_pah_df) == "benz[a]anthracene" ] <- "benzaanthracene"
colnames(all_pah_df)[colnames(all_pah_df) == "benzo[a]pyrene" ] <- "benzoapyrene"
colnames(all_pah_df)[colnames(all_pah_df) == "benzo[b]fluoranthene" ] <- "benzobfluoranthene"
colnames(all_pah_df)[colnames(all_pah_df) == "benzo[e]pyrene" ] <- "benzoepyrene"
colnames(all_pah_df)[colnames(all_pah_df) == "benzo[ghi]perylene" ] <- "benzoghiperylene"
colnames(all_pah_df)[colnames(all_pah_df) == "benzo[k]fluoranthene" ] <- "benzokfluoranthene"
colnames(all_pah_df)[colnames(all_pah_df) == "dibenz[a,h]anthracene" ] <- "dibenzahanthracene"
colnames(all_pah_df)[colnames(all_pah_df) == "indeno[1,2,3-cd]pyrene" ] <- "indeno123cdpyrene"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-benzanthracene/schrysenes" ] <- "C1benzanthraceneschrysenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-dibenzothiophenes" ] <- "C1dibenzothiophenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-fluoranthenes/pyrenes" ] <- "C1fluoranthenespyrenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-fluorenes" ] <- "C1fluorenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-naphthalenes" ] <- "C1naphthalenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C1-phenanthrenes/anthracenes" ] <- "C1phenanthrenesanthracenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-benzanthracenes/chrysenes" ] <- "C2benzanthraceneschrysenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-dibenzothiophenes" ] <- "C2dibenzothiophenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-fluoranthenes/pyrenes" ] <- "C2fluoranthenespyrenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-fluorenes" ] <- "C2fluorenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-naphthalenes" ] <- "C2naphthalenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C2-phenanthrenes/anthracenes" ] <- "C2phenanthrenesanthracenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-benzanthracenes/chrysenes" ] <- "C3benzanthraceneschrysenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-dibenzothiophenes" ] <- "C3dibenzothiophenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-fluoranthenes/pyrenes" ] <- "C3fluoranthenespyrenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-fluorenes" ] <- "C3fluorenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-naphthalenes" ] <- "C3naphthalenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C3-phenanthrenes/anthracenes" ] <- "C3phenanthrenesanthracenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C4-benzanthracenes/chrysenes" ] <- "C4benzanthraceneschrysenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C4-dibenzothiophenes" ] <- "C4dibenzothiophenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C4-fluoranthenes/pyrenes" ] <- "C4fluoranthenespyrenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C4-naphthalenes" ] <- "C4naphthalenes"
colnames(all_pah_df)[colnames(all_pah_df) == "C4-phenanthrenes/anthracenes" ] <- "C4phenanthrenesanthracenes" 

# Fit a GAM model
gam_model <- gam(p450 ~ acenaphthene + acenaphthylene + anthracene + `benzaanthracene` + `benzoapyrene` +  `benzobfluoranthene` + `benzoepyrene` + `benzoghiperylene` + `benzokfluoranthene` + chrysene + `dibenzahanthracene` + dibenzothiophene + `indeno123cdpyrene` + phenanthrene + C1dibenzothiophenes + C1fluoranthenespyrenes + C1fluorenes + C1naphthalenes + C1phenanthrenesanthracenes + C2benzanthraceneschrysenes + C2dibenzothiophenes + C2fluoranthenespyrenes + C2fluorenes + C2naphthalenes + C2phenanthrenesanthracenes + C3benzanthraceneschrysenes + C3dibenzothiophenes + C3fluoranthenespyrenes + C3naphthalenes + C3phenanthrenesanthracenes + C4benzanthraceneschrysenes + C4dibenzothiophenes + C4fluoranthenespyrenes + C4naphthalenes + C4phenanthrenesanthracenes , data = all_pah_df)

# Summary of the model
summary(gam_model)

# Check diagnostics (plots to assess the model)
par(mfrow = c(2, 2))  # Set up the plotting area for a 2 by 2 layout
plot(gam_model)  # Diagnostic plots

# Predict using the model (if necessary)
predictions <- predict(gam_model, newdata = new_data)

# Replace 'new_data' with your new data frame for prediction if needed.

```

Get my data ranges for SOD

```{r}

# Calculate the lower and upper percentiles (2.5th and 97.5th)
lower_limit <- quantile(all_pah_df$SOD, probs = 0.025, na.rm= TRUE)
upper_limit <- quantile(all_pah_df$SOD, probs = 0.975, na.rm= TRUE)

# Print the lower and upper limits
print(lower_limit)
print(upper_limit)

all_pah_df$SOD[all_pah_df$SOD <= .0025] <- NA
all_pah_df$SOD[all_pah_df$SOD >= 51.53] <- NA




```
#trying to get the pca for sod to match the one for p450 but that isn't working

```{r}
# PCA Plot with biomarkers
library('FactoMineR')
library("factoextra")

all_columns <- names(all_analyte_df)

# Remove the columns you don't want to include in the model
excluded_columns <- c('latitude', 'longitude', 'site_name', 'site_number', 'p450', 'NA', "condition_factor")
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Remove NAs from the dataset
df_clean <- na.omit(all_analyte_df)

# Use the list 'independent_columns' to subset 'df_clean'
pca_data <- df_clean[, independent_columns]

# Perform PCA on the subset of data
pca_result <- PCA(pca_data, scale.unit = TRUE, ncp = 5, graph = FALSE)

# Visualizations (Optional but helpful for interpreting PCA results)
# Eigenvalues - to see the importance of each principal component
fviz_eig(pca_result)

# Individuals - to see the distribution of samples
fviz_pca_ind(pca_result, 
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             title = "PCA - Individuals (colored by cos2)",
             repel = TRUE) # Avoid text overlapping

# Variables - to see how variables contribute to the principal components
fviz_pca_var(pca_result,
             col.var = "contrib", # Color by contribution to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             title = "PCA - Variables (colored by contribution)",
             repel = TRUE) # Avoid text overlapping


# Selecting the relevant variables for PCA
#pca_data <- df_clean[, independent_columns]

# Performing PCA
#pca_res <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Plotting the PCA
#pcaplot<- fviz_pca_biplot(pca_res, label = "var", col.var = "contrib",
             #   gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             #   repel = TRUE)  # Avoid text overlapping (slow if many points)



```

# sorting p450 values for results conversation
```{r}

# Sorting the data in ascending order
# Assuming your dataframe is named df
sorted_df <- all_analyte_df[order(all_analyte_df$p450), ]


# Printing the ranked list
print(sorted_df)

```

```{r}
summary(all_analyte_df)
```

