---
title: "099-Manuscript Visualization Options"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 10,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load data
```{r}
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/p450data.csv")
indv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgP450_analytes.csv")

sdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/soddata.csv")
Sindv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD_analytes.csv")

mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphdata.csv")
Mindv<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgMORPH_analytes.csv")
```

# Adjust SOD individual values to reflect LOQ limitations
```{r}

# Data contains numbers below 0 that must be adjusted.These numbers represent samples whose values were below the limit of quantification using the SOD kit. There was some activity in the raw sample, but it is too far below the limit of the standard curve to be accurately quantified.


#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
sdata$sod[sdata$sod <= 0] <- 0.0025

```

# sod
## dumbbell - sites in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_sod_range <- sdata %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_sod = min(sod), max_sod = max(sod), mean_sod = mean(sod))

# Order site names
site_sod_range <- site_sod_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_sod_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_sod, xend = max_sod, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_sod, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_sod, color = as.factor(reporting_area)), size = 3) +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "SOD Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - grouped by reporting area with indicators of mean value (triangles)
```{r}
library(ggplot2)
library(dplyr)


# Order by reporting area
site_sod_range <- site_sod_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_sod_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean
ggplot(site_sod_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_sod, xend = max_sod, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_sod, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_sod, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_sod), shape = 17, size = 3, color = "black") +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "SOD Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Ssite_sod_range <- sdata %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_sod = min(sod), max_sod = max(sod), mean_sod = mean(sod))

# Create a unique identifier for each combination of site_name and PAHgroup8
Ssite_sod_range <- Ssite_sod_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Ssite_sod_range <- Ssite_sod_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Ssite_sod_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Ssite_sod_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_sod, xend = max_sod, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_sod, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_sod, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_sod), shape = 17, size = 3, color = "black") +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "SOD Activity by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}

library(ggplot2)
library(dplyr)

# Order by site names
data <- sdata %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(data, aes(x = site_name, y = sod, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "SOD Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +  
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1) 
  )

```

## box plot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- sdata %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = sod, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), y = "Site Name", title = "SOD Activity by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
sodbox <- sdata %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

sodbox <- sodbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
sodbox <- sodbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(sodbox, aes(x = site_name_ordered, y = sod, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"), title = "SOD Activity by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## correlation matrix with significant analytes only
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
Scor_results <- sapply(Sindv[ , 3:ncol(Sindv)], function(x) {
  test <- cor.test(Sindv$avg_value, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
Sspearman_cor_matrix <- cor(Sindv[ , 3:ncol(Sindv)], use = "complete.obs", method = "spearman")
Sspearman_cor_avg_value <- Sspearman_cor_matrix["avg_value", ]

# Filter for significance using corr_results for p-values
Ssignificant_analytes <- names(Scor_results[Scor_results < 0.05])

# New correlation matrix with only significant analytes
Ssignificant_cor_matrix <- Sspearman_cor_matrix[Ssignificant_analytes, Ssignificant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(Ssignificant_cor_matrix, method = "circle")

```

## correlation table
```{r}

library(dplyr)
library(tibble)

# Spearman correlation 
Scor_results <- sapply(Sindv[ , 3:ncol(Sindv)], function(x) {
  test <- cor.test(Sindv$avg_value, x, method = "spearman")
  c(correlation = test$estimate, p_value = test$p.value)
})

# Convert to data frame, fix column names and remove first row of old column names
Scor_results_df <- as.data.frame(t(Scor_results))

colnames(Scor_results_df) <- c("correlation", "p_value")

Scor_results_df <- Scor_results_df %>%
  rownames_to_column(var = "analyte")

Scor_results_df <- Scor_results_df %>%
  slice(-1) 

# Filter for significance and round the results
Ssignificant_results_df <- Scor_results_df %>%
  filter(p_value < 0.05)

Ssignificant_results_df <- Ssignificant_results_df %>%
  mutate(correlation = round(correlation, 3),
         p_value = round(p_value, 3))

# View table
print(Ssignificant_results_df)

# Write to table
#write.csv(Ssignificant_results_df,"/Users/cmantegna/Documents/WDFWmussels/output/tables/VIZsodcorrelation.csv", row.names = FALSE)

```

## linear regression with signficant analyte correlation
```{r}

library(ggplot2)
library(gridExtra)

Sindv$reporting_area <- as.character(Sindv$reporting_area)

# Custom colors to match other plot colors
reporting_area_colors <- c(
  "6" = "red",
  "7" = "blue",
  "8.1" = "green",
  "8.2" = "purple",
  "9" = "orange",
  "10" = "yellow",
  "11" = "brown",
  "12" = "pink",
  "13" = "grey"
  
)


# Only significant analytes
Ssignificant_analytes <- Ssignificant_results_df$analyte

# Create an empty list to store plots
plot_list <- list()

# Loop through each significant analyte and create a linear regression plot
for (analyte in Ssignificant_analytes) {
  
  # Ensure the analyte exists in df before proceeding
  if (!analyte %in% colnames(Sindv)) {
    warning(paste("Analyte", analyte, "is not present in the data frame. Skipping..."))
    next
  }
  
  # Linear Model
  model <- lm(Sindv[[analyte]] ~ Sindv$avg_value)
  
  # Plot
p <- ggplot(Sindv, aes_string(x = "avg_value", y = analyte, color = "reporting_area")) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black", se = FALSE) +  
  scale_color_manual(values = reporting_area_colors) +  
  labs(title = paste("Linear Regression: Average SOD Activity vs", analyte),
       x = bquote("SOD Activity (Unit / mg"^"-1"~"protein)"),
       y = analyte) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend

# Store the plot in the list
plot_list[[analyte]] <- p

}

# Display the plots in a grid
do.call(grid.arrange, c(plot_list, ncol = 3)) 

```


# shell thickness (at)
## dumbbell - original data in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_at_range <- mdata %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_at = min(at), max_at = max(at), mean_at = mean(at))

# Order by site name
site_at_range <- site_at_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_at_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_at, xend = max_at, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(reporting_area)), size = 3) +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - reporting area with indicators of mean value (triangles)
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
site_at_range <- site_at_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_at_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean added
ggplot(site_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_at, xend = max_at, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_at), shape = 17, size = 3, color = "black") +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Asite_at_range <- mdata %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_at = min(at), max_at = max(at), mean_at = mean(at))

# Create a unique identifier for each combination of site_name and PAHgroup8
Asite_at_range <- Asite_at_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Asite_at_range <- Asite_at_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Asite_at_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Asite_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_at, xend = max_at, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_at, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_at, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_at), shape = 17, size = 2, color = "black") +
  labs(x = bquote("Shell Thickness"), y = "Site Name", title = "Shell Thickness by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}

library(ggplot2)
library(dplyr)

# Order by site names
data <- mdata %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(data, aes(x = site_name, y = at, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +  
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1) 
  )

```

## box plot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- mdata %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = at, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Shell Thickness (mm)", y = "Site Name", title = "Shell Thickness by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
atbox <- mdata %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

atbox <- atbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
atbox <- atbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(atbox, aes(x = site_name_ordered, y = at, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = "Shell Thickness (mm)", title = "Shell Thickness by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## correlation matrix with significant analytes only
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
Mcor_results <- sapply(Mindv[ , 3:ncol(Mindv)], function(x) {
  test <- cor.test(Mindv$avg_at, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
Mspearman_cor_matrix <- cor(Mindv[ , 3:ncol(Mindv)], use = "complete.obs", method = "spearman")
Mspearman_cor_avg_value <- Mspearman_cor_matrix["avg_at", ]

# Filter for significance using corr_results for p-values
Msignificant_analytes <- names(Mcor_results[Mcor_results < 0.05])

# New correlation matrix with only significant analytes
Msignificant_cor_matrix <- Mspearman_cor_matrix[Msignificant_analytes, Msignificant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(Msignificant_cor_matrix, method = "circle")

```

## correlation table
```{r}

library(dplyr)
library(tibble)

# Spearman correlation 
Mcor_results <- sapply(Mindv[ , 3:ncol(Mindv)], function(x) {
  test <- cor.test(Mindv$avg_at, x, method = "spearman")
  c(correlation = test$estimate, p_value = test$p.value)
})

# Convert to data frame, fix column names and remove first row of old column names
Mcor_results_df <- as.data.frame(t(Mcor_results))

colnames(Mcor_results_df) <- c("correlation", "p_value")

Mcor_results_df <- Mcor_results_df %>%
  rownames_to_column(var = "analyte")

Mcor_results_df <- Mcor_results_df %>%
  slice(-1) 

# Filter for significance and round the results
Msignificant_results_df <- Mcor_results_df %>%
  filter(p_value < 0.05)

Msignificant_results_df <- Msignificant_results_df %>%
  mutate(correlation = round(correlation, 3),
         p_value = round(p_value, 3))

# View table
print(Msignificant_results_df)

# Write to table
#write.csv(Msignificant_results_df,"/Users/cmantegna/Documents/WDFWmussels/output/tables/VIZatcorrelation.csv", row.names = FALSE)

```

## linear regression with signficant analyte correlation
```{r}

library(ggplot2)
library(gridExtra)

Mindv$reporting_area <- as.character(Mindv$reporting_area)

# Custom colors to match other plot colors
reporting_area_colors <- c(
  "6" = "red",
  "7" = "blue",
  "8.1" = "green",
  "8.2" = "purple",
  "9" = "orange",
  "10" = "yellow",
  "11" = "brown",
  "12" = "pink",
  "13" = "grey"
  
)


# Only significant analytes
Msignificant_analytes <- Msignificant_results_df$analyte

# Create an empty list to store plots
plot_list <- list()

# Loop through each significant analyte and create a linear regression plot
for (analyte in Msignificant_analytes) {
  
  # Ensure the analyte exists in df before proceeding
  if (!analyte %in% colnames(Mindv)) {
    warning(paste("Analyte", analyte, "is not present in the data frame. Skipping..."))
    next
  }
  
  # Linear Model
  model <- lm(Mindv[[analyte]] ~ Mindv$avg_at)
  
  # Plot
p <- ggplot(Mindv, aes_string(x = "avg_at", y = analyte, color = "reporting_area")) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black", se = FALSE) +  
  scale_color_manual(values = reporting_area_colors) +  
  labs(title = paste("Linear Regression: Shell Thickness vs", analyte),
       x = "Shell Thickness (mm)",
       y = analyte) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove the legend

# Store the plot in the list
plot_list[[analyte]] <- p

}

# Display the plots in a grid
do.call(grid.arrange, c(plot_list, ncol = 4)) 

```

# condition factor (cf)
## dumbbell - original data in alphabetical order
```{r}

library(ggplot2)
library(dplyr) 

# Calculate range + mean
site_cf_range <- mdata %>%
  group_by(site_name, reporting_area) %>%
  summarise(min_cf = min(cf), max_cf = max(cf), mean_cf = mean(cf))

# Order by site name
site_cf_range <- site_cf_range %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(site_cf_range, aes(y = reorder(site_name, site_name))) + 
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name, yend = site_name, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(reporting_area)), size = 3) +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - reporting area with indicators of mean value (triangles)
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
site_cf_range <- site_cf_range %>%
  mutate(site_name_ordered = factor(site_name,
                                    levels = site_cf_range %>%
                                      arrange(desc(reporting_area), site_name) %>%
                                      pull(site_name)))

# Plot w/ mean added
ggplot(site_cf_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name_ordered, yend = site_name_ordered, color = as.factor(reporting_area)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(reporting_area)), size = 3) +
  geom_point(aes(x = mean_cf), shape = 17, size = 1, color = "black") +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## dumbbell - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Calculate range + mean, order by PAH group
Csite_at_range <- mdata %>%
  group_by(site_name, PAHgroup8) %>%
  summarise(min_cf = min(cf), max_cf = max(cf), mean_cf = mean(cf))

# Create a unique identifier for each combination of site_name and PAHgroup8
Csite_at_range <- Csite_at_range %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
Csite_at_range <- Csite_at_range %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = Csite_at_range %>%
                                      arrange(desc(PAHgroup8), site_name) %>%
                                      pull(site_PAHgroup_combined)))

# Plot w/ mean added
ggplot(Csite_at_range, aes(y = site_name_ordered)) +
  geom_segment(aes(x = min_cf, xend = max_cf, y = site_name_ordered, yend = site_name_ordered, color = as.factor(PAHgroup8)), size = 1.5) +
  geom_point(aes(x = min_cf, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = max_cf, color = as.factor(PAHgroup8)), size = 3) +
  geom_point(aes(x = mean_cf), shape = 17, size = 2, color = "black") +
  labs(x = bquote("Condition Factor"), y = "Site Name", title = "Condition Factor by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

## box plot - site
```{r}

library(ggplot2)
library(dplyr)

# Order by site names
data <- mdata %>%
  mutate(site_name = factor(site_name, levels = rev(sort(unique(site_name)))))

# Plot
ggplot(data, aes(x = site_name, y = cf, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +  
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1) 
  )

```

## box plot - reporting area
```{r}

library(ggplot2)
library(dplyr)

# Order by reporting area
data <- mdata %>%
  mutate(site_name_ordered = factor(site_name, 
                                    levels = data %>%
                                      arrange(reporting_area, site_name) %>%
                                      distinct(site_name, reporting_area) %>%  
                                      pull(site_name)))

# Plot using the newly ordered site names
ggplot(data, aes(x = site_name_ordered, y = cf, fill = as.factor(reporting_area))) +
  geom_boxplot() +
  labs(x = "Condition Factor", y = "Site Name", title = "Condition Factor by Site with Reporting Areas") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "Reporting Area") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## boxplot - PAH group 8
```{r}

library(ggplot2)
library(dplyr)

# Create a unique identifier for each combination of site_name and PAHgroup8
cfbox <- mdata %>%
  mutate(site_PAHgroup_combined = paste(site_name, PAHgroup8, sep = "_"))

cfbox <- cfbox %>%
  arrange(desc(PAHgroup8), site_name)

# Order site_name by both PAHgroup8 and site_name to ensure no duplication
cfbox <- cfbox %>%
  mutate(site_name_ordered = factor(site_PAHgroup_combined,
                                    levels = unique(site_PAHgroup_combined)))

# Plot using the newly ordered site names
ggplot(cfbox, aes(x = site_name_ordered, y = cf, fill = as.factor(PAHgroup8))) +
  geom_boxplot() +
  labs(x = "Condition Factor", title = "Condition Factor by Site with PAH Grouping") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set1", name = "PAH Grouping") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

## correlation matrix with significant analytes only - there are none
```{r}

library(corrplot)

# Spearman correlation tests to extract p-values
Mcor_results <- sapply(Mindv[ , 4:ncol(Mindv)], function(x) {
  test <- cor.test(Mindv$avg_cf, x, method = "spearman")
  test$p.value  # Extract p-values
})

# Spearman correlation matrix for plotting
Mspearman_cor_matrix <- cor(Mindv[ , 4:ncol(Mindv)], use = "complete.obs", method = "spearman")
Mspearman_cor_avg_value <- Mspearman_cor_matrix["avg_cf", ]

# Filter for significance using corr_results for p-values
Msignificant_analytes <- names(Mcor_results[Mcor_results < 0.05])

# New correlation matrix with only significant analytes
Msignificant_cor_matrix <- Mspearman_cor_matrix[Msignificant_analytes, Msignificant_analytes]

# Plot the correlation matrix for significant analytes
corrplot(Msignificant_cor_matrix, method = "circle")

```
