---
title: "10- Creating an IBR Index"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

# libraries
library(knitr)
library(tidyr)
library(tidyverse)
library(dplyr)
library(vegan)
library(cluster)
library(pgirmess)
library(ggplot2)
library(factoextra)
library(FactoMineR)
library(FSA)           
library(rstatix)       
library(car)
library(RVAideMemoire)
library(rcompanion)
library(scales)

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

getwd()
setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels")
# load data
df<- read.csv("../data/cleaned/all_with_imputation_weights_site_ra.csv")

# check it out
str(df) # checking df structure
summary(df) #checking df contents

```

```{r }

# replace any values below LOQ with imputed LOQ (.5 x LOD), LOD= 0.05
df$sod[df$sod <= 0] <- 0.0025

# check
summary(df)

```

# Creating the Beliaeff and Burgeot IBR Index with biomarkers and morphometrics
## Normalizing the data
Step 1 - Normalize biomarker values. Use z-scores/ min-max scaling per site or sample to put them on a comparable scale.
Step 2 -Transform morphometrics to reduce scale variability.
Step 3 - Check for outliers. Extreme values may skew integrated scores.
## Optional 
Step 0 - Separate IBRs: IBR biomarkers, IBR morphometrics, IBR both (do this after step 3)
Step 1 - Score biomarkers and morphometrics. Use grand mean or the reference site - Penn Cove Ref Site.
Step 2 - Find deviation from the reference. A positive deviation is more stressed and a negative deviation is less stress.
Step 3 - Plot it. Radar plots and polygon area plots will support calculations.
```{r}

# select and scale biomarkers and raw morphometrics
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

ibr_scaled <- scale(ibr_vars) # standardize (Z-score)

ibr_scores <- abs(as.data.frame(ibr_scaled)) # convert to absolute Z-scores (deviation) 

df$IBR <- rowSums(ibr_scores) # calculate IBR Index (sum of absolute scores)

# plot
ibr_box<- ggplot(df, aes(x = site_name, y = IBR)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Integrated Biomarker Response (IBR) by Site",
       y = "IBR Index", x = "Site") +
  theme(axis.text.x= element_text(angle= 45, hjust= 1))

#save and view
ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibr_box.png", plot= ibr_box, width = 16, height = 14, dpi = 300)

print(ibr_box)

```
# Radar Plots
## Test sample
```{r}

#install.packages("fmsb")
library("fmsb")

# Select and scale
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm, ci1, ci2, ci3)

scaled_df <- scale(ibr_vars)
abs_scores <- abs(as.data.frame(scaled_df))

# Add sample/site identifiers back
abs_scores$site_name <- df$site_name
abs_scores$sample_id <- df$sample_id

# Choose one sample (e.g., row 1)
sample <- abs_scores[1, 1:11]  # only the numeric variables
max_min <- rbind(apply(abs_scores[ ,1:11], 2, max),
                 apply(abs_scores[ ,1:11], 2, min))
radar_data <- rbind(max_min, sample)

radarchart(radar_data,
           axistype = 1,
           pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.8, 0.4),
           plwd = 2,
           title = paste("IBR Radar Plot for Sample", abs_scores$sample_id[1]))

```

## All individual samples
```{r}

library(fmsb)

# Prepare absolute z-score data
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm, ci1, ci2, ci3)

scaled <- scale(ibr_vars)
abs_scores <- abs(as.data.frame(scaled))

# Add ID columns back
abs_scores$sample_id <- df$sample_id
abs_scores$site_name <- df$site_name

# Loop through each sample and create radar plots
for (i in 1:nrow(abs_scores)) {
  sample_data <- abs_scores[i, 1:11]
  max_min <- rbind(apply(abs_scores[ ,1:11], 2, max),
                   apply(abs_scores[ ,1:11], 2, min))
  radar_data <- rbind(max_min, sample_data)
  
  # Save to file
  png(paste0("radar_sample_", abs_scores$sample_id[i], ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.8, 0.4, 0.4),
             plwd = 2,
             title = paste("Sample", abs_scores$sample_id[i], "\nSite:", abs_scores$site_name[i]))
  dev.off()
}

```

## Site plots
```{r}

# Calculate site-level mean absolute Z-scores
site_means <- abs_scores %>%
  group_by(site_name) %>%
  summarise(across(1:11, mean, na.rm = TRUE))

# Loop through each site
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:12]  # skip site_name column
  max_min <- rbind(apply(site_means[ ,2:12], 2, max),
                   apply(site_means[ ,2:12], 2, min))
  radar_data <- rbind(max_min, site_data)
  
  # Save to file
  png(paste0("radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("Mean IBR Profile\nSite:", site_means$site_name[i]))
  dev.off()
}

```


# Creating the Mattos IBRv2i Index with biomarkers and morphometrics
Step 0: Identify reference site (Penn Cove Ref)
Step 1 - Normalize biomarker and morphometric values. Use z-scores/ min-max scaling per site or sample to put them on a comparable scale and retain the directionality of responses based on ref site.
## Optional 
Step 0 - Separate IBRs: IBR biomarkers, IBR morphometrics, IBR both (do this after step 3)
Step 1 - Score biomarkers and morphometrics. Use grand mean or the reference site - Penn Cove Ref Site.
Step 2 - Find deviation from the reference. A positive deviation is more stressed and a negative deviation is less stress.
Step 3 - Plot it. Radar plots and polygon area plots will support calculations.
```{r}

# select and scale biomarkers and raw morphometrics
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

# ID ref site/ samples
ref_df <- df %>% filter(site_number == 3)
ref_vars <- ref_df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

# calculate reference means and SDs
ref_means <- colMeans(ref_vars, na.rm = TRUE)
ref_sds <- apply(ref_vars, 2, sd, na.rm = TRUE)

# Z-score relative to ref site
z_scores <- sweep(ibr_vars, 2, ref_means, FUN = "-")
z_scores <- sweep(z_scores, 2, ref_sds, FUN = "/")

df$IBRv2i <- rowMeans(z_scores, na.rm = TRUE) # IBRv2i = mean of reference-based Z-scores (preserving sign)

#plot
ibrv2i_box<- ggplot(df, aes(x = site_name, y = IBRv2i)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "IBRv2i by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# save and view
ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibrv2i_box.png", plot= ibrv2i_box, width = 16, height = 14, dpi = 300)

print(ibrv2i_box)

```

# Plotting All Samples IBRv2i
```{r}

# Site-averaged Z-scores
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(1:8, mean, na.rm = TRUE))

# Loop through sites
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)

  png(paste0("IBRv2i_radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}


```


```{r}

# Loop through each site
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)
  
  # Save to your custom path
  output_file <- paste0("/Users/cmantegna/Documents/GitHub/WDFWmussels/output/radar_plots_ibrv2i/radar_site_", 
                        gsub(" ", "_", site_means$site_name[i]), ".png")
  
  png(output_file, width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}


```

# Stats
PERMANOVA
```{r}

```

# Visualization
PCA/ NMDS
Correlation heatmaps
```{r}

```

