---
title: "10- Creating an IBR Index"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

# libraries
library(knitr)
library(tidyr)
library(tidyverse)
library(dplyr)
library(vegan)
library(cluster)
library(pgirmess)
library(ggplot2)
library(factoextra)
library(FactoMineR)
library(FSA)           
library(rstatix)       
library(car)
library(RVAideMemoire)
library(rcompanion)
library(scales)
library(fmsb)

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  #fig.width = 15,       # Set plot width in inches
  #fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load & Check Data
```{r}

getwd()
setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels")
# load data
df<- read.csv("../data/cleaned/all_with_imputation_weights_site_ra.csv")

# check it out
str(df) # checking df structure
summary(df) #checking df contents

```

```{r }

# replace any values below LOQ with imputed LOQ (.5 x LOD), LOD= 0.05
df$sod[df$sod <= 0] <- 0.0025

# check
summary(df)

```

# Creating the Beliaeff and Burgeot IBR Index with biomarkers and morphometrics
## Normalizing the data
### DO NOT RUN Again except for comparison purposes

Step 1 - Normalize biomarker values. Use z-scores/ min-max scaling per site or sample to put them on a comparable scale.
Step 2 -Transform morphometrics to reduce scale variability.
Step 3 - Check for outliers. Extreme values may skew integrated scores.
## Optional 
Step 0 - Separate IBRs: IBR biomarkers, IBR morphometrics, IBR both (do this after step 3)
Step 1 - Score biomarkers and morphometrics. Use grand mean or the reference site - Penn Cove Ref Site.
Step 2 - Find deviation from the reference. A positive deviation is more stressed and a negative deviation is less stress.
Step 3 - Plot it. Radar plots and polygon area plots will support calculations.
```{r}

# select and scale biomarkers and raw morphometrics
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

ibr_scaled <- scale(ibr_vars) # standardize (Z-score)

ibr_scores <- abs(as.data.frame(ibr_scaled)) # convert to absolute Z-scores (deviation) 

df$IBR <- rowSums(ibr_scores) # calculate IBR Index (sum of absolute scores)

# plot
ibr_box<- ggplot(df, aes(x = site_name, y = IBR)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Integrated Biomarker Response (IBR) by Site",
       y = "IBR Index", x = "Site") +
  theme(axis.text.x= element_text(angle= 45, hjust= 1))

#save and view
ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibr_box.png", plot= ibr_box, width = 16, height = 14, dpi = 300)

print(ibr_box)

```
# Radar Plots
## Test sample
```{r}

#install.packages("fmsb")
library("fmsb")

# Select and scale
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm, ci1, ci2, ci3)

scaled_df <- scale(ibr_vars)
abs_scores <- abs(as.data.frame(scaled_df))

# Add sample/site identifiers back
abs_scores$site_name <- df$site_name
abs_scores$sample_id <- df$sample_id

# Choose one sample (e.g., row 1)
sample <- abs_scores[1, 1:11]  # only the numeric variables
max_min <- rbind(apply(abs_scores[ ,1:11], 2, max),
                 apply(abs_scores[ ,1:11], 2, min))
radar_data <- rbind(max_min, sample)

radarchart(radar_data,
           axistype = 1,
           pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.8, 0.4),
           plwd = 2,
           title = paste("IBR Radar Plot for Sample", abs_scores$sample_id[1]))

```

## All individual samples
```{r}

library(fmsb)

# Prepare absolute z-score data
ibr_vars <- df %>%
  select(p450, sod, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm, ci1, ci2, ci3)

scaled <- scale(ibr_vars)
abs_scores <- abs(as.data.frame(scaled))

# Add ID columns back
abs_scores$sample_id <- df$sample_id
abs_scores$site_name <- df$site_name

# Loop through each sample and create radar plots
for (i in 1:nrow(abs_scores)) {
  sample_data <- abs_scores[i, 1:11]
  max_min <- rbind(apply(abs_scores[ ,1:11], 2, max),
                   apply(abs_scores[ ,1:11], 2, min))
  radar_data <- rbind(max_min, sample_data)
  
  # Save to file
  png(paste0("radar_sample_", abs_scores$sample_id[i], ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.8, 0.4, 0.4),
             plwd = 2,
             title = paste("Sample", abs_scores$sample_id[i], "\nSite:", abs_scores$site_name[i]))
  dev.off()
}

```

## Site plots
```{r}

# Calculate site-level mean absolute Z-scores
site_means <- abs_scores %>%
  group_by(site_name) %>%
  summarise(across(1:11, mean, na.rm = TRUE))

# Loop through each site
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:12]  # skip site_name column
  max_min <- rbind(apply(site_means[ ,2:12], 2, max),
                   apply(site_means[ ,2:12], 2, min))
  radar_data <- rbind(max_min, site_data)
  
  # Save to file
  png(paste0("radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("Mean IBR Profile\nSite:", site_means$site_name[i]))
  dev.off()
}

```


# Creating the Mattos IBRv2i Index with biomarkers and morphometrics
Step 0: Identify reference site (Penn Cove Ref)
Step 1 - Normalize biomarker and morphometric values. Use z-scores/ min-max scaling per site or sample to put them on a comparable scale and retain the directionality of responses based on ref site.

## Optional Extra IBRv2i breakdown
Step 0 - Separate IBRs: IBR biomarkers, IBR morphometrics, IBR both (do this after step 3)
Step 1 - Score biomarkers and morphometrics. Use grand mean or the reference site - Penn Cove Ref Site.
Step 2 - Find deviation from the reference. A positive deviation is more stressed and a negative deviation is less stress.
Step 3 - Plot it. Radar plots and polygon area plots will support calculations.
```{r}

# make categorical columns factors
df$site_number <- as.factor(df$site_number)
df$sample_id <- as.factor(df$sample_id)
df$reporting_area <- as.factor(df$reporting_area)

# Select and scale biomarkers and raw morphometrics
ibr_vars <- df %>%
  select(p450, sod, weight_initial_g, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

# ID reference site/samples
ref_df <- df %>% filter(site_number == 3)
ref_vars <- ref_df %>%
  select(p450, sod, weight_initial_g, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm)

# Calculate reference means and SDs
ref_means <- colMeans(ref_vars, na.rm = TRUE)
ref_sds <- apply(ref_vars, 2, sd, na.rm = TRUE)

# Z-score relative to reference site
z_scores <- sweep(ibr_vars, 2, ref_means, FUN = "-")
z_scores <- sweep(z_scores, 2, ref_sds, FUN = "/")
z_scores <- as.data.frame(z_scores)

# Add back identifying columns
z_scores <- bind_cols(
  df %>% select(latitude, longitude, site_name, sample_id, reporting_area),
  z_scores
)

# IBRv2i = mean of reference-based Z-scores (preserving sign)
z_scores$IBRv2i <- rowMeans(z_scores %>% select(p450:width_mm), na.rm = TRUE)

# Optional: also store in the original df
df$IBRv2i <- z_scores$IBRv2i


#plot
ibrv2i_box<- ggplot(df, aes(x = site_name, y = IBRv2i)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "IBRv2i by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# plot with Penn Cove indicated
df$highlight <- ifelse(df$site_name == "Penn Cove Reference", "Penn Cove Reference", "Other")

ibrv2i_box2 <- ggplot(df, aes(x = site_name, y = IBRv2i, fill = highlight)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Penn Cove Reference" = "red", "Other" = "white")) +
  theme_minimal() +
  labs(title = "IBRv2i by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")  # hides legend if not needed


# save and view
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibrv2i_box_reference_indicated.png", plot= ibrv2i_box2, width = 16, height = 14, dpi = 300)

print(ibrv2i_box)
print(ibrv2i_box2)

```

# Plotting All Sites IBRv2i
## Do Not Run
```{r}

z_scores$sample_id <- df$sample_id
z_scores$site_name <- df$site_name

# Site-averaged Z-scores
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(1:8, mean, na.rm = TRUE))

# Loop through sites
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)

  png(paste0("IBRv2i_radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}

```

# Plotting All Samples IBRv2i
## Only run if spot checking
```{r}

library(fmsb)

# Loop through each sample
for (i in 1:nrow(z_scores)) {
  
  # Extract sample data
  sample_data <- z_scores[i, 1:8]
  
  # Get axis limits (min/max across all samples)
  max_min <- rbind(apply(z_scores[, 1:8], 2, max, na.rm = TRUE),
                   apply(z_scores[, 1:8], 2, min, na.rm = TRUE))
  
  # Combine for radar input
  radar_data <- rbind(max_min, sample_data)
  
  # Set output file name (in current directory)
  filename <- paste0("radar_sample_", z_scores$sample_id[i], ".png")
  
  # Save plot
  png(filename, width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.8, 0.2, 0.2, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSample:", z_scores$sample_id[i],
                           "\nSite:", z_scores$site_name[i]))
  dev.off()
}


```

# All sites w/ reference site overlay, IBRv2i
```{r}

# Step 1: Create the reference profile
reference_profile <- z_scores %>%
  filter(sample_id %in% 41:44) %>%
  select(p450, sod, weight_initial_g, weight_change_g, weight_final_g, shell, length_mm, height_mm, width_mm) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

# Step 2: Create site-level means for the same variables
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(p450:width_mm, mean, na.rm = TRUE)) %>%
  ungroup()

# Step 3: Define radar axis limits (same across all sites)
max_vals <- apply(site_means[, 2:10], 2, max, na.rm = TRUE)
min_vals <- apply(site_means[, 2:10], 2, min, na.rm = TRUE)

# Step 4: Loop through each site to plot
for (i in 1:nrow(site_means)) {
  site <- site_means[i, ]
  site_profile <- site[, 2:10]  # extract only the measurement columns
  
  # Skip if identical to reference
  if (all.equal(as.numeric(site_profile), as.numeric(reference_profile)) == TRUE) next
  
  # Combine into radar data
  radar_data <- rbind(max_vals, min_vals, reference_profile, site_profile)
  rownames(radar_data) <- c("Max", "Min", "Reference", as.character(site$site_name))
  
  # Save plot
  filename <- paste0("radar_site_with_ref_", gsub(" ", "_", site$site_name), ".png")
  png(filename, width = 600, height = 600)
  
  radarchart(radar_data,
             axistype = 1,
             pcol = c("gray40", "steelblue"),
             pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
             plwd = 2,
             plty = 1,
             title = paste("IBRv2i Radar Plot with Reference\nSite:", site$site_name))
  
  legend("topright", legend = c("Reference (Penn Cove)", site$site_name),
         col = c("gray40", "steelblue"), lty = 1, lwd = 2, bty = "n")
  
  dev.off()
}

```

# Plotting Sites by Reporting Areas w/ Reference, IBRv2i
```{r}

library(dplyr)
library(fmsb)

# Prep: site-level means and reporting area info
site_means <- z_scores %>%
  group_by(site_name, reporting_area) %>%
  summarise(across(p450:width_mm, mean, na.rm = TRUE), .groups = "drop")

# Define limits across ALL sites (for consistent radar scale)
global_max <- apply(site_means[, 3:11], 2, max, na.rm = TRUE)
global_min <- apply(site_means[, 3:11], 2, min, na.rm = TRUE)

# Create reference profile from Penn Cove samples
reference_profile <- z_scores %>%
  filter(sample_id %in% 41:44) %>%
  summarise(across(p450:width_mm, mean, na.rm = TRUE), .groups = "drop")

# Now loop through each reporting area
reporting_areas <- unique(site_means$reporting_area)

for (area in reporting_areas) {
  area_sites <- site_means %>% filter(reporting_area == area)
  n_sites <- nrow(area_sites)
  n_pages <- ceiling(n_sites / 6)

  for (page in 1:n_pages) {
    # Subset to current group of up to 6 sites
    start_i <- (page - 1) * 6 + 1
    end_i <- min(start_i + 5, n_sites)
    sites_subset <- area_sites[start_i:end_i, ]

    # Set output
    filename <- paste0("radar_group_", gsub(" ", "_", area), "_page_", page, ".png")
    png(filename, width = 1200, height = 800)
    par(mfrow = c(2, 3), mar = c(2, 2, 4, 2))  # layout: 2 rows x 3 cols

    for (i in 1:nrow(sites_subset)) {
      site_data <- sites_subset[i, 3:11]  # numeric columns only

      # Format data for radarchart
      chart_data <- rbind(global_max, global_min, reference_profile, site_data)
      rownames(chart_data) <- c("Max", "Min", "Reference", as.character(sites_subset$site_name[i]))

      radarchart(chart_data,
                 axistype = 1,
                 pcol = c("gray40", "steelblue"),
                 pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
                 plwd = 2,
                 plty = 1,
                 cglcol = "grey80",
                 title = sites_subset$site_name[i])
    }

    mtext(paste("Reporting Area:", area), side = 3, line = -2, outer = TRUE, cex = 1.5)
    dev.off()
  }
}

```


# Separate the IBRs and run the same things
# Run PCAs/ NMDSs/ Permanova/ Correlations + Heatmaps/ Kendall's Tau
```{r}



```


# Stats
PERMANOVA
```{r}

```

# Visualization
## Poster viz
```{r}

#map
library(sf)
library(readr)
library(patchwork)
library(viridis)
library(rnaturalearth)
library(ggsignif)

df$reporting_area <- as.factor(df$reporting_area)

# Convert to spatial
sites_sf <- st_as_sf(df, coords = c("longitude", "latitude"), crs = 4326)

coast <- ne_states(country = "united states of america", returnclass = "sf") %>%
  filter(name == "Washington")

# Major city coordinates for labels
cities <- data.frame(
  city = c("Seattle", "Tacoma", "Olympia", "Bellingham", "Port Angeles", "Hama Hama"),
  lon = c(-122.33, -122.45, -122.90, -122.48, -123.43, -123.16),
  lat = c(47.61, 47.25, 47.00, 48.73, 48.08, 47.63)
)

# Define color palette
area_colors <- viridis::viridis(length(levels(df$reporting_area)), option = "turbo")

# Map
map_plot <- ggplot() +
  geom_sf(data = coast, fill = "gray90", color = "white") +
  geom_sf(data = sites_sf, aes(color = reporting_area), size = 3, alpha = 0.9) +
  geom_text(data = cities, aes(x = lon, y = lat, label = city),
            color = "black", fontface = "bold", size = 4) +
  scale_color_manual(values = area_colors, name = "Reporting Area") + 
  coord_sf(xlim = c(-124, -121), ylim = c(46.5, 49.0)) +
  theme_void() +
  labs(title = "Sampled Sites by Reporting Area")

# Filter data for IBRv2i < 5
df_filtered <- df %>% filter(IBRv2i < 5)
df_filtered <- df_filtered %>%
  mutate(is_reference = ifelse(site_name == "Penn Cove Reference", "Reference", "Other"))

# Boxplot
box_plot <- ggplot(df_filtered, aes(x = reporting_area, y = IBRv2i, fill = reporting_area)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.2, color = "black", size = 1.5) +
  scale_fill_manual(values = area_colors, guide = "none") + 
  coord_cartesian(ylim = c(-3, 5.5)) +
  geom_signif(comparisons = list(
      c("11", "7"),
      c("13", "7"),
      c("6", "7"),
      c("10", "13")
    ),
    annotations = c("p = 0.0093", "p = 0.0013", "p = 0.0487", "p = 0.0199"),
    y_position = c(5.2, 4.8, 4.5, 4.2),
    tip_length = 0.02,
    textsize = 4.5
  ) +
  theme_minimal(base_size = 13) + 
  theme(
    axis.line = element_line(size = 1.2, color = "black"),
    axis.text = element_text(size= 25, face = "bold"),
    axis.title = element_text(size= 30, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(x = "Reporting Area", y = "IBRv2i Score")
  
# Combine
combo_plot <- map_plot + box_plot +
  plot_layout(ncol = 2, widths = c(1.3, 1)) +
  plot_annotation(title = "Integrated Biomarker Response Overview")

# Display
print(map_plot)
print(box_plot)

# Export 
ggsave("/Users/cmantegna/Documents/GitHub/WDFWmussels/output/poster_map.png", map_plot, width = 14, height = 8, dpi = 300)

ggsave("/Users/cmantegna/Documents/GitHub/WDFWmussels/output/poster_box.png", box_plot, width = 14, height = 8, dpi = 300)


```
## KW for plotting
```{r}

# Make sure reporting_area is a factor
df$reporting_area <- as.factor(df$reporting_area)

# Run Kruskal-Wallis test
kruskal_result <- kruskal.test(IBRv2i ~ reporting_area, data = df)
print(kruskal_result)

# Run Dunn's post hoc test
dunn_result <- dunnTest(IBRv2i ~ reporting_area, data = df, method = "holm")
print(dunn_result)


```



PCA/ NMDS
Correlation heatmaps
```{r}

```

