---
title: "10- Creating an IBR & Contaminant Index"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

# libraries
library(knitr) # output fotmatting
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(dplyr) # data wrangling
library(vegan) # ecological stats 
library(cluster) # grouping metrics - VERIFY still needed
library(pgirmess) # stats - KW
library(ggplot2) # plots
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(FSA) # post hoc test - Dunn's Test       
library(rstatix) # VERIFY what this is for      
library(car) # VERIFY what this is for  
library(RVAideMemoire) # post hoc test for permanova
library(rcompanion) # KW testing
library(scales) # scaling data for IBR - works with data wrangling packages
library(fmsb) # polygon calculations for the radars

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

```

# Load & Check Data
```{r}

getwd()
#setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels") # something here isn't working right - check out why

df1<- read.csv("../data/all_with_imputations_and_identifiers.csv")
chlordane<- read.csv("../data/indices/test_chlordane.csv")
ddt<- read.csv("../data/indices/test_ddt.csv")
hch<- read.csv("../data/indices/test_hch.csv")
metal<- read.csv("../data/indices/test_metal.csv")
pbde<- read.csv("../data/indices/test_pbde.csv")
pcb<- read.csv("../data/indices/test_pcb.csv")
pesticide<- read.csv("../data/indices/test_pesticide.csv")

str(df1) # checking df structure (classes of each column of data)
summary(df1) #checking df contents (quick stats for each column)

```

```{r }

df1$sod[df1$sod <= 0] <- 0.0025 # replace any values below LOQ with imputed LOQ (.5 x LOD), LOD= 0.05

summary(df1) # check

```

# Creating IBRv2i Index with biomarkers and raw morphometrics (no condition indices)
## df prep
```{r}

# make categorical columns factors
df1$site_number <- as.factor(df1$site_number)
df1$sample_id <- as.factor(df1$sample_id)
df1$reporting_area <- as.factor(df1$reporting_area)

```

## fixing site names & trimming white space
```{r}

chlordane$site[chlordane$site == "Blair Waterway #2"] <- "Blair Waterway Two"
chlordane$site[chlordane$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
chlordane$site[chlordane$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
chlordane$site[chlordane$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
chlordane$site[chlordane$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
chlordane$site[chlordane$site == "Purdy - Dexters"] <- "Purdy, Dexters"
chlordane$site[chlordane$site == "S of Skunk Island"] <- "South of Skunk Island"
chlordane$site[chlordane$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

ddt$site[ddt$site == "Blair Waterway #2"] <- "Blair Waterway Two"
ddt$site[ddt$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
ddt$site[ddt$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
ddt$site[ddt$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
ddt$site[ddt$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
ddt$site[ddt$site == "Purdy - Dexters"] <- "Purdy, Dexters"
ddt$site[ddt$site == "S of Skunk Island"] <- "South of Skunk Island"
ddt$site[ddt$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

hch$site[hch$site == "Blair Waterway #2"] <- "Blair Waterway Two"
hch$site[hch$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
hch$site[hch$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
hch$site[hch$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
hch$site[hch$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
hch$site[hch$site == "Purdy - Dexters"] <- "Purdy, Dexters"
hch$site[hch$site == "S of Skunk Island"] <- "South of Skunk Island"
hch$site[hch$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

metal$site[metal$site == "Blair Waterway #2"] <- "Blair Waterway Two"
metal$site[metal$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
metal$site[metal$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
metal$site[metal$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
metal$site[metal$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
metal$site[metal$site == "Purdy - Dexters"] <- "Purdy, Dexters"
metal$site[metal$site == "S of Skunk Island"] <- "South of Skunk Island"
metal$site[metal$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"
metal$analyte[metal$analyte == "mercuryTotal"] <- "mercury"
metal$analyte[metal$analyte == "Zinc"] <- "zinc"

pbde$site[pbde$site == "Blair Waterway #2"] <- "Blair Waterway Two"
pbde$site[pbde$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
pbde$site[pbde$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
pbde$site[pbde$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
pbde$site[pbde$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
pbde$site[pbde$site == "Purdy - Dexters"] <- "Purdy, Dexters"
pbde$site[pbde$site == "S of Skunk Island"] <- "South of Skunk Island"
pbde$site[pbde$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

pcb$site[pcb$site == "Blair Waterway #2"] <- "Blair Waterway Two"
pcb$site[pcb$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
pcb$site[pcb$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
pcb$site[pcb$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
pcb$site[pcb$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
pcb$site[pcb$site == "Purdy - Dexters"] <- "Purdy, Dexters"
pcb$site[pcb$site == "S of Skunk Island"] <- "South of Skunk Island"
pcb$site[pcb$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

pesticide$site[pesticide$site == "Blair Waterway #2"] <- "Blair Waterway Two"
pesticide$site[pesticide$site == "Comm Bay Skookum"] <- "Commencement Bay Skookum"
pesticide$site[pesticide$site == "Comm Bay, Dick Gilmur Launch"] <- "Commencement Bay, Dick Gilmur Launch"
pesticide$site[pesticide$site == "Comm Bay, Milwaukee Waterway"] <- "Commencement Bay, Milwaukee Waterway"
pesticide$site[pesticide$site == "Meyer's Point - Henderson Inlet"] <- "Meyer's Point, Henderson Inlet"
pesticide$site[pesticide$site == "Purdy - Dexters"] <- "Purdy, Dexters"
pesticide$site[pesticide$site == "S of Skunk Island"] <- "South of Skunk Island"
pesticide$site[pesticide$site == "Suquamish, Stormwater Outfall"] <- "Suquamish Stormwater Outfall"

df1 <- df1 %>%
  mutate(across(where(is.character), str_trim))

chlordane <- chlordane %>%
  mutate(across(where(is.character), str_trim))

ddt <- ddt %>%
  mutate(across(where(is.character), str_trim))

hch <- hch %>%
  mutate(across(where(is.character), str_trim))

metal <- metal %>%
  mutate(across(where(is.character), str_trim))

pbde <- pbde %>%
  mutate(across(where(is.character), str_trim))

pcb <- pcb %>%
  mutate(across(where(is.character), str_trim))

pesticide <- pesticide %>%
  mutate(across(where(is.character), str_trim))

```

## Reducing all sample df to site-level
```{r}

# cleaning up site names and then averaging by site for new df
#data$site_name <- trimws(data$site_name)
#data <- data %>%
#  filter(!is.na(site_name), site_name != "")

df1 <- df1 %>%
  group_by(site_name, reporting_area) %>% 
  summarise(
    latitude= first(latitude),
    longitude= first(longitude),
    p450 = mean(p450, na.rm = TRUE),
    sod = mean(sod, na.rm = TRUE),
    shell = mean(shell, na.rm = TRUE),
    weight_initial = mean(weight_initial_g, na.rm = TRUE),
    weight_change = mean(weight_change_g, na.rm = TRUE),
    weight_final = mean(weight_final_g, na.rm = TRUE),
    length = mean(length_mm, na.rm = TRUE),
    height = mean(height_mm, na.rm = TRUE),
    width = mean(width_mm, na.rm = TRUE),
    ci1 = mean(ci1, na.rm = TRUE),
    ci2 = mean(ci2, na.rm = TRUE),
    ci3 = mean(ci3, na.rm = TRUE)
  ) %>%
  ungroup()

head(df1)
#write.csv(df1, "/Users/cmantegna/Documents/Github/WDFWmussels/data/cleaned/avg_site_level.csv", row.names = FALSE)

```

## replace undetected analyte values 
```{r}

chlordane<- chlordane %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

ddt<- ddt %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

hch<- hch %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

metal<- metal %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

pbde<- pbde %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

pcb<- pcb %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

pesticide<- pesticide %>%
  mutate(flagged = grepl("U|B|I", qualifier),
         wet_value = ifelse(flagged, 0, wet_value))

```

## pivoting analytes
```{r}

chlordane_wide <- chlordane %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

ddt_wide <- ddt %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

hch_wide <- hch %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

metal_wide <- metal_df %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

pbde_wide <- pbde %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

pcb_wide <- pcb %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )

pesticide_wide <- pesticide %>%
  pivot_wider(
    names_from = analyte,      # this becomes the new column names
    values_from = wet_value    # this fills the values
  )


```


## IBR - biomarker, morphometric, combined
```{r}

# define variables used in building the ibr - 2 biomarkers and 7 morphometric values, CI left out because it is derived from raw measurements that are included in the ibr

biomarkers<- c("p450", "sod")
morphometrics<- c("shell", "weight_initial_g", "weight_change_g", "weight_final_g", "length_mm", "height_mm", "width_mm")

# define the reference groups - only using Penn Cove
# penn cove= site 3, using in index 1 and 3, hc holly = site 29, using in index 2 and 3, broad spit = site 30, using in index 2 and 3

#### Penn Cove Only 
ref_df <- df1 %>% filter(site_name == "Penn Cove Reference") # define reference site

# not using since I'm separating biomarkers from morphometrics to balance their weight in the index
#ref_vars <- ref_df %>%
#  select(p450, sod, shell, weight_initial_g, weight_change_g, weight_final_g, length_mm, height_mm, width_mm) 

ref_bio_means <- colMeans(ref_df[, biomarkers], na.rm = TRUE) # reference site means - biomarkers
ref_bio_sds   <- apply(ref_df[, biomarkers], 2, sd, na.rm = TRUE) # reference site sds - biomarkers
ref_morph_means <- colMeans(ref_df[, morphometrics], na.rm = TRUE) # same as above - morphometrics
ref_morph_sds   <- apply(ref_df[, morphometrics], 2, sd, na.rm = TRUE)  # same as above - morphometrics

bio_z <- sweep(df[, biomarkers], 2, ref_bio_means, "-") # z-scores, maintaining sign - biomarkers
bio_z <- sweep(bio_z, 2, ref_bio_sds, "/") # z-scores- actual conversion from the centered values calculated above
morph_z <- sweep(df[, morphometrics], 2, ref_morph_means, "-")
morph_z <- sweep(morph_z, 2, ref_morph_sds, "/")
# add to OG df or create a new one and name the columns correctly
ibr_df <- df %>%
  select(latitude, longitude, site_name, reporting_area) %>%
  bind_cols(bio_z, morph_z) %>%
  mutate(
    ibr_biomarker = rowMeans(bio_z, na.rm = TRUE),
    ibr_morphometric = rowMeans(morph_z, na.rm = TRUE),
    ibr_combined = rowMeans(cbind(ibr_biomarker, ibr_morphometric), na.rm = TRUE))



```

## adding condition indices back to df
```{r}

# define variables
condition <- c("ci1", "ci2", "ci3")

#### Penn Cove Only
ref_df <- df %>% filter(site_name == "Penn Cove Reference") # define reference site
ref_ci_means <- colMeans(ref_df[, condition], na.rm = TRUE) # reference means
ref_ci_sds   <- apply(ref_df[, condition], 2, sd, na.rm = TRUE) # reference sds
condition_z <- sweep(df[, condition], 2, ref_ci_means, "-") # z-scores relative to reference site - preserving sign
condition_z <- sweep(condition_z, 2, ref_ci_sds, "/") # z-scores
ibr_df <- bind_cols(ibr_df, condition_z) # add the new columns to the df
ibr_df <- ibr_df %>%
 mutate(
    ci1_1 = condition_z$ci1,
    ci2_1 = condition_z$ci2,
    ci3_1 = condition_z$ci3,)


# save it
#write.csv(ibr_df, "/Users/cmantegna/Documents/GitHub/WDFWmussels/data/cleaned/complete_cleaned_ibr.csv", row.names = FALSE)

```

## Contaminant Index - chlordane, ddt, hch, metal, pbde, pcb, pesticide
```{r}



```




# Plotting
## Box plots, Separate Indices per Site
```{r}

#df1<- read.csv("../data/cleaned/ibr_1.csv")


# pivot only the index columns - adjusted for each df
ibr3_long <- df3 %>%
  pivot_longer(cols = c(ibr3_biomarkers, ibr3_morphometrics, ibr3_overall),
               names_to = "index_type",
               values_to = "index_value") %>%
  mutate(index_type = case_when(
    index_type == "ibr3_biomarkers" ~ "Biomarkers",
    index_type == "ibr3_morphometrics" ~ "Morphometrics",
    index_type == "ibr3_overall" ~ "Overall",
    TRUE ~ index_type))  # fallback in case of unexpected names

#### plotting ibr1 with Penn Cove reference site
all_index<- ggplot(ibr1_long, aes(x = site_name, y = index_value,
                     color = site_name == "Penn Cove Reference")) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  facet_wrap(~ index_type, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none") +
  labs(title = "IBRv2i Indices by Site",
       x = "Site", y = "IBRv2i Value")

# save and view
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibr1_all_indices.png", plot= all_index, width = 16, height = 14, dpi = 300)

print(all_index)



```
## Radars, reference overlay
## All Sites, Do Not Run
```{r}

z_scores$sample_id <- df$sample_id
z_scores$site_name <- df$site_name

# Site-averaged Z-scores
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(1:8, mean, na.rm = TRUE))

# Loop through sites
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)

  png(paste0("IBRv2i_radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}

```

## All Samples,Only run if spot checking
```{r}

library(fmsb)

# Loop through each sample
for (i in 1:nrow(z_scores)) {
  
  # Extract sample data
  sample_data <- z_scores[i, 1:8]
  
  # Get axis limits (min/max across all samples)
  max_min <- rbind(apply(z_scores[, 1:8], 2, max, na.rm = TRUE),
                   apply(z_scores[, 1:8], 2, min, na.rm = TRUE))
  
  # Combine for radar input
  radar_data <- rbind(max_min, sample_data)
  
  # Set output file name (in current directory)
  filename <- paste0("radar_sample_", z_scores$sample_id[i], ".png")
  
  # Save plot
  png(filename, width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.8, 0.2, 0.2, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSample:", z_scores$sample_id[i],
                           "\nSite:", z_scores$site_name[i]))
  dev.off()
}


```

## All sites w/ reference site overlay
```{r}

#### IBR_1
# Variables to include in radar plots
vars_to_use <- c("p450_ibr1", "sod_ibr1", "shell_ibr1", "weight_initial_g_ibr1", "weight_final_g_ibr1", "weight_change_g_ibr1", "length_mm_ibr1", "height_mm_ibr1", "width_mm_ibr1")

# create the reference profile from Penn Cove 
reference_profile <- z_scores %>%
  filter(sample_id %in% 41:44) %>%
  select(all_of(vars_to_use)) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

# create site-level means 
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE)) %>%
  ungroup()

# define radar axis limits (same across all sites)
max_vals <- apply(site_means[, vars_to_use], 2, max, na.rm = TRUE)
min_vals <- apply(site_means[, vars_to_use], 2, min, na.rm = TRUE)

# convert reference profile to a named numeric vector
ref_vec <- unlist(reference_profile[1, ], use.names = TRUE)

# loop through each site to create radar plots
for (i in 1:nrow(site_means)) {
  site <- site_means[i, ]
  site_profile <- unlist(site[, vars_to_use], use.names = TRUE)
  
  # Combine into radar data
  radar_data <- as.data.frame(rbind(max_vals, min_vals, ref_vec, site_profile))
  rownames(radar_data) <- c("Max", "Min", "Reference", as.character(site$site_name))
  
  # Save radar plot
  filename <- paste0("radar_site_with_ref_", gsub(" ", "_", site$site_name), ".png")
  png(filename, width = 600, height = 600)
  
  radarchart(radar_data,
             axistype = 1,
             pcol = c("gray40", "steelblue"),
             pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
             plwd = 2,
             plty = 1,
             title = paste("IBRv2i Radar Plot with Reference\nSite:", site$site_name))
  
  legend("topright", legend = c("Penn Cove Reference", site$site_name),
         col = c("gray40", "steelblue"), lty = 1, lwd = 2, bty = "n")
  
  dev.off()
}


```

## Sites by Reporting Areas w/ Reference
```{r}

# manually adjusting for each ibr
# Define variables to include (excluding condition indices and weight final)
vars_to_use <- c("p450_ibr3", "sod_ibr3", "shell_ibr3", "weight_initial_g_ibr3", "weight_change_g_ibr3", "length_mm_ibr3", "height_mm_ibr3", "width_mm_ibr3")

# Site-level means (if you have multiple entries per site)
site_means <- df3 %>%
  group_by(site_name, reporting_area) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE), .groups = "drop")

# Global min/max for radar chart scaling
global_max <- apply(site_means[, vars_to_use], 2, max, na.rm = TRUE)
global_min <- apply(site_means[, vars_to_use], 2, min, na.rm = TRUE)

# Reference profile from one or more reference sites
reference_sites <- c( "Penn Cove Reference", "Hood Canal Holly", "Broad Spit (Fisherman's Point)")  # or use site_number
reference_profile <- df3 %>%
  filter(site_name %in% reference_sites) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE), .groups = "drop") %>%
  unlist(use.names = TRUE)

# Proceed with the rest of the radar chart code
reporting_areas <- unique(site_means$reporting_area)

for (area in reporting_areas) {
  area_sites <- site_means %>% filter(reporting_area == area)
  n_sites <- nrow(area_sites)
  n_pages <- ceiling(n_sites / 6)

  for (page in 1:n_pages) {
    start_i <- (page - 1) * 6 + 1
    end_i <- min(start_i + 5, n_sites)
    sites_subset <- area_sites[start_i:end_i, ]

    filename <- paste0("radar_group_", gsub(" ", "_", area), "_page_", page, ".png")
    png(filename, width = 1200, height = 800)
    par(mfrow = c(2, 3), mar = c(2, 2, 4, 2), oma = c(0, 0, 4, 0))

    for (i in 1:nrow(sites_subset)) {
      site_data <- unlist(sites_subset[i, vars_to_use], use.names = TRUE)
      chart_data <- as.data.frame(rbind(global_max, global_min, reference_profile, site_data))
      rownames(chart_data) <- c("Max", "Min", "Reference", as.character(sites_subset$site_name[i]))

      radarchart(chart_data,
                 axistype = 1,
                 pcol = c("gray40", "steelblue"),
                 pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
                 plwd = 2,
                 plty = 1,
                 cglcol = "grey80",
                 title = sites_subset$site_name[i])
    }

    mtext(paste("Reporting Area:", area), side = 3, line = 1, outer = TRUE, cex = 1.0)
    dev.off()
  }
}

```





