---
title: "10- Creating an IBR Index"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

# libraries
library(knitr) # output fotmatting
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(dplyr) # data wrangling
library(vegan) # ecological stats 
library(cluster) # grouping metrics - VERIFY still needed
library(pgirmess) # stats - KW
library(ggplot2) # plots
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(FSA) # post hoc test - Dunn's Test       
library(rstatix) # VERIFY what this is for      
library(car) # VERIFY what this is for  
library(RVAideMemoire) # post hoc test for permanova
library(rcompanion) # KW testing
library(scales) # scaling data for IBR - works with data wrangling packages
library(fmsb) # polygon calculations for the radars

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

```

# Load & Check Data
```{r}

getwd()
#setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels") # something here isn't working right - check out why

df<- read.csv("../data/all_with_imputation_weights_site_ra.csv") # load data

str(df) # checking df structure (classes of each column of data)
summary(df) #checking df contents (quick stats for each column)

```

```{r }

df$sod[df$sod <= 0] <- 0.0025 # replace any values below LOQ with imputed LOQ (.5 x LOD), LOD= 0.05

summary(df) # check

```

# Creating IBRv2i Index with biomarkers and raw morphometrics (no condition indices)
```{r}

# make categorical columns factors
df$site_number <- as.factor(df$site_number)
df$sample_id <- as.factor(df$sample_id)
df$reporting_area <- as.factor(df$reporting_area)

```

```{r}

# define variables used in building the ibr - 2 biomarkers and 7 morphometric values, CI left out because it is derived from raw measurements that are included in the ibr

biomarkers<- c("p450", "sod")
morphometrics<- c("shell", "weight_initial_g", "weight_change_g", "weight_final_g", "length_mm", "height_mm", "width_mm")

# define the reference groups
# penn cove= site 3, using in index 1 and 3, hc holly = site 29, using in index 2 and 3, broad spit = site 30, using in index 2 and 3

#### Penn Cove Only 
ref_df <- df %>% filter(site_number == 3) # define reference site

# not using since I'm separating biomarkers from morphometrics to balance their weight in the index
#ref_vars <- ref_df %>%
#  select(p450, sod, shell, weight_initial_g, weight_change_g, weight_final_g, length_mm, height_mm, width_mm) 

ref_bio_means <- colMeans(ref_df[, biomarkers], na.rm = TRUE) # reference site means - biomarkers
ref_bio_sds   <- apply(ref_df[, biomarkers], 2, sd, na.rm = TRUE) # reference site sds - biomarkers
ref_morph_means <- colMeans(ref_df[, morphometrics], na.rm = TRUE) # same as above - morphometrics
ref_morph_sds   <- apply(ref_df[, morphometrics], 2, sd, na.rm = TRUE)  # same as above - morphometrics

bio_z <- sweep(df[, biomarkers], 2, ref_bio_means, "-") # z-scores, maintaining sign - biomarkers
bio_z <- sweep(bio_z, 2, ref_bio_sds, "/") # z-scores- actual conversion from the centered values calculated above
morph_z <- sweep(df[, morphometrics], 2, ref_morph_means, "-")
morph_z <- sweep(morph_z, 2, ref_morph_sds, "/")
# add to OG df or create a new one and name the columns correctly
ibr_df <- df %>%
  select(latitude, longitude, sample_id,site_number, site_name, reporting_area) %>%
  bind_cols(bio_z, morph_z) %>%
  mutate(
    ibr1_biomarkers = rowMeans(bio_z, na.rm = TRUE),
    ibr1_morphometrics = rowMeans(morph_z, na.rm = TRUE),
    ibr1_overall = rowMeans(cbind(ibr1_biomarkers, ibr1_morphometrics), na.rm = TRUE))

#### Penn Cove & HC Sites 
ref_df <- df %>% filter(site_number %in% c(3, 29, 30)) # define reference sites
ref_vars <- ref_df %>%
  select(p450, sod, shell, weight_initial_g, weight_change_g, weight_final_g, length_mm, height_mm, width_mm)

ref_bio_means <- colMeans(ref_df[, biomarkers], na.rm = TRUE) # reference site means - bio
ref_bio_sds   <- apply(ref_df[, biomarkers], 2, sd, na.rm = TRUE) # reference site sds - bio
ref_morph_means <- colMeans(ref_df[, morphometrics], na.rm = TRUE) 
ref_morph_sds   <- apply(ref_df[, morphometrics], 2, sd, na.rm = TRUE) 

bio_z <- sweep(df[, biomarkers], 2, ref_bio_means, "-") # z-scores prep
bio_z <- sweep(bio_z, 2, ref_bio_sds, "/") # z-score completion
morph_z <- sweep(df[, morphometrics], 2, ref_morph_means, "-")
morph_z <- sweep(morph_z, 2, ref_morph_sds, "/")
# add to df
ibr_df <- bind_cols(ibr_df, bio_z, morph_z)
ibr_df <- ibr_df %>%
 mutate(
    ibr3_biomarkers = rowMeans(bio_z, na.rm = TRUE),
    ibr3_morphometrics = rowMeans(morph_z, na.rm = TRUE),
    ibr3_overall = rowMeans(cbind(ibr3_biomarkers, ibr3_morphometrics), na.rm = TRUE))

#### HC Sites Only
ref_df <- df %>% filter(site_number %in% c(29, 30)) # define reference sites
ref_vars <- ref_df %>%
  select(p450, sod, shell, weight_initial_g, weight_change_g, weight_final_g, length_mm, height_mm, width_mm)

ref_bio_means <- colMeans(ref_df[, biomarkers], na.rm = TRUE) # refernce site means
ref_bio_sds   <- apply(ref_df[, biomarkers], 2, sd, na.rm = TRUE) # reference site sds
ref_morph_means <- colMeans(ref_df[, morphometrics], na.rm = TRUE)
ref_morph_sds   <- apply(ref_df[, morphometrics], 2, sd, na.rm = TRUE)

bio_z <- sweep(df[, biomarkers], 2, ref_bio_means, "-") # z-scores
bio_z <- sweep(bio_z, 2, ref_bio_sds, "/")
morph_z <- sweep(df[, morphometrics], 2, ref_morph_means, "-")
morph_z <- sweep(morph_z, 2, ref_morph_sds, "/")
# combine into final df with signed values
ibr_df <- bind_cols(ibr_df, bio_z, morph_z)
ibr_df <- ibr_df %>%
 mutate(
    ibr2_biomarkers = rowMeans(bio_z, na.rm = TRUE),
    ibr2_morphometrics = rowMeans(morph_z, na.rm = TRUE),
    ibr2_overall = rowMeans(cbind(ibr2_biomarkers, ibr2_morphometrics), na.rm = TRUE))

# saved the complete df in the next block after adding the condition indices back to the df

```

## adding condition indices back to df
```{r}

# define variables
condition <- c("ci1", "ci2", "ci3")

#### Penn Cove Only
ref_df <- df %>% filter(site_name == "Penn Cove Reference") # define reference site
ref_ci_means <- colMeans(ref_df[, condition], na.rm = TRUE) # reference means
ref_ci_sds   <- apply(ref_df[, condition], 2, sd, na.rm = TRUE) # reference sds
condition_z <- sweep(df[, condition], 2, ref_ci_means, "-") # z-scores relative to reference site - preserving sign
condition_z <- sweep(condition_z, 2, ref_ci_sds, "/") # z-scores
ibr_df <- bind_cols(ibr_df, condition_z) # add the new columns to the df
ibr_df <- ibr_df %>%
 mutate(
    ci1_1 = condition_z$ci1,
    ci2_1 = condition_z$ci2,
    ci3_1 = condition_z$ci3,)

#### HC Sites Only
ref_df <- df %>% filter(site_number %in% c(29, 30)) # define reference sites, switched to the site numbers instead of names to catch them all - one of the broad spit samples keeps getting dropped - the name may be typed different
ref_ci_means <- colMeans(ref_df[, condition], na.rm = TRUE) # reference means
ref_ci_sds   <- apply(ref_df[, condition], 2, sd, na.rm = TRUE) # reference sds
condition_z <- sweep(df[, condition], 2, ref_ci_means, "-") # z-score prep relative to reference site
condition_z <- sweep(condition_z, 2, ref_ci_sds, "/") # z-scores completion
ibr_df <- bind_cols(ibr_df, condition_z) # add the new columns to the df
ibr_df <- ibr_df %>%
 mutate(
    ci1_2 = condition_z$ci1,
    ci2_2 = condition_z$ci2,
    ci3_2 = condition_z$ci3,)

#### Penn Cove & HC Sites
ref_df <- df %>% filter(site_number %in% c(3, 29, 30)) # define reference sites
ref_ci_means <- colMeans(ref_df[, condition], na.rm = TRUE) # reference means
ref_ci_sds   <- apply(ref_df[, condition], 2, sd, na.rm = TRUE) # reference sds
condition_z <- sweep(df[, condition], 2, ref_ci_means, "-") # z-scores prep
condition_z <- sweep(condition_z, 2, ref_ci_sds, "/") # z-score completion
ibr_df <- bind_cols(ibr_df, condition_z) # add the new columns to the full df
ibr_df <- ibr_df %>%
 mutate(
    ci1_3 = condition_z$ci1,
    ci2_3 = condition_z$ci2,
    ci3_3 = condition_z$ci3,)

# save it
#write.csv(ibr_df, "/Users/cmantegna/Documents/GitHub/WDFWmussels/data/cleaned/complete_cleaned_ibr.csv", row.names = FALSE)

```

# Plotting
Separate Indices per Site, box plot
```{r}

df1<- read.csv("../data/cleaned/ibr_1.csv")
df2<- read.csv("../data/cleaned/ibr_2.csv")
df3<- read.csv("../data/cleaned/ibr_3.csv")

# pivot only the index columns
ibr_long <- ibrv2i_df %>%
  pivot_longer(cols = c(IBRv2i_biomarkers, IBRv2i_morphometrics, IBRv2i_overall),
               names_to = "index_type",
               values_to = "index_value") %>%
  mutate(index_type = case_when(
    index_type == "IBRv2i_biomarkers" ~ "Biomarkers",
    index_type == "IBRv2i_morphometrics" ~ "Morphometrics",
    index_type == "IBRv2i_overall" ~ "Overall",
    TRUE ~ index_type  # fallback in case of unexpected names
  ))


all_index<- ggplot(ibr_long, aes(x = site_name, y = index_value,
                     color = site_name == "Penn Cove Reference")) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  facet_wrap(~ index_type, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none") +
  labs(title = "IBRv2i Indices by Site",
       x = "Site", y = "IBRv2i Value")

# save and view
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibrv2i_box_all_index_reference_indicated.png", plot= all_index, width = 16, height = 14, dpi = 300)

print(all_index)

```

```{r}




# plot with Penn Cove indicated
df$highlight <- ifelse(df$site_name == "Penn Cove Reference", "Penn Cove Reference", "Other")

ibr1_box <- ggplot(df, aes(x = site_name, y = ibr, fill = highlight)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Penn Cove Reference" = "red", "Other" = "white")) +
  theme_minimal() +
  labs(title = "IBR_1 by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")  # hides legend if not needed

# plot with Penn Cove indicated
df$highlight <- ifelse(df$site_name == "Penn Cove Reference", "Penn Cove Reference", "Other")

ibr2_box <- ggplot(df, aes(x = site_name, y = IBRv2i, fill = highlight)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Penn Cove Reference" = "red", "Other" = "white")) +
  theme_minimal() +
  labs(title = "IBRv2i by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")  # hides legend if not needed

# plot with Penn Cove indicated
df$highlight <- ifelse(df$site_name == "Penn Cove Reference", "Penn Cove Reference", "Other")

ibr3_box <- ggplot(df, aes(x = site_name, y = IBRv2i, fill = highlight)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Penn Cove Reference" = "red", "Other" = "white")) +
  theme_minimal() +
  labs(title = "IBRv2i by Site", y = "IBRv2i Index", x = "Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")  # hides legend if not needed

# save and view
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/ibrv2i_box_reference_indicated.png", plot= ibrv2i_box2, width = 16, height = 14, dpi = 300)

print(ibrv2i_box)
print(ibrv2i_box2)


```


# Plotting - Only moving forward with IBRv2i indices
## All Sites, Do Not Run
```{r}

z_scores$sample_id <- df$sample_id
z_scores$site_name <- df$site_name

# Site-averaged Z-scores
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(1:8, mean, na.rm = TRUE))

# Loop through sites
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)

  png(paste0("IBRv2i_radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}

```

## All Samples,Only run if spot checking
```{r}

library(fmsb)

# Loop through each sample
for (i in 1:nrow(z_scores)) {
  
  # Extract sample data
  sample_data <- z_scores[i, 1:8]
  
  # Get axis limits (min/max across all samples)
  max_min <- rbind(apply(z_scores[, 1:8], 2, max, na.rm = TRUE),
                   apply(z_scores[, 1:8], 2, min, na.rm = TRUE))
  
  # Combine for radar input
  radar_data <- rbind(max_min, sample_data)
  
  # Set output file name (in current directory)
  filename <- paste0("radar_sample_", z_scores$sample_id[i], ".png")
  
  # Save plot
  png(filename, width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.8, 0.2, 0.2, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSample:", z_scores$sample_id[i],
                           "\nSite:", z_scores$site_name[i]))
  dev.off()
}


```

## All sites w/ reference site overlay
```{r}

df1<- read.csv("../data/cleaned/complete_cleaned_ibr.csv")
df2
df3

# Variables to include in radar plots
vars_to_use <- c("p450", "sod", "weight_initial_g", "weight_change_g", 
                 "weight_final_g", "shell", "length_mm", "height_mm", "width_mm")

# Step 1: Create the reference profile from Penn Cove samples (e.g., sample_id 41:44)
reference_profile <- z_scores %>%
  filter(sample_id %in% 41:44) %>%
  select(all_of(vars_to_use)) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

# Step 2: Create site-level means for the same variables
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE)) %>%
  ungroup()

# Step 3: Define radar axis limits (same across all sites)
max_vals <- apply(site_means[, vars_to_use], 2, max, na.rm = TRUE)
min_vals <- apply(site_means[, vars_to_use], 2, min, na.rm = TRUE)

# Convert reference profile to a named numeric vector
ref_vec <- unlist(reference_profile[1, ], use.names = TRUE)

# Step 4: Loop through each site to create radar plots
for (i in 1:nrow(site_means)) {
  site <- site_means[i, ]
  site_profile <- unlist(site[, vars_to_use], use.names = TRUE)
  
  # Skip if site profile matches reference exactly
  if (isTRUE(all.equal(site_profile, ref_vec))) next
  
  # Combine into radar data
  radar_data <- as.data.frame(rbind(max_vals, min_vals, ref_vec, site_profile))
  rownames(radar_data) <- c("Max", "Min", "Reference", as.character(site$site_name))
  
  # Save radar plot
  filename <- paste0("radar_site_with_ref_", gsub(" ", "_", site$site_name), ".png")
  png(filename, width = 600, height = 600)
  
  radarchart(radar_data,
             axistype = 1,
             pcol = c("gray40", "steelblue"),
             pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
             plwd = 2,
             plty = 1,
             title = paste("IBRv2i Radar Plot with Reference\nSite:", site$site_name))
  
  legend("topright", legend = c("Penn Cove Reference", site$site_name),
         col = c("gray40", "steelblue"), lty = 1, lwd = 2, bty = "n")
  
  dev.off()
}


```

## Sites by Reporting Areas w/ Reference
```{r}

# Define variables to include (excluding condition indices)
vars_to_use <- c("p450", "sod", "weight_initial_g", "weight_change_g", 
                 "weight_final_g", "shell", "length_mm", "height_mm", "width_mm")

# Prep: site-level means and reporting area info
site_means <- z_scores %>%
  group_by(site_name, reporting_area) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE), .groups = "drop")

# Global min/max across all sites
global_max <- apply(site_means[, vars_to_use], 2, max, na.rm = TRUE)
global_min <- apply(site_means[, vars_to_use], 2, min, na.rm = TRUE)

# Reference profile from Penn Cove samples
reference_profile <- z_scores %>%
  filter(sample_id %in% 41:44) %>%
  summarise(across(all_of(vars_to_use), mean, na.rm = TRUE), .groups = "drop") %>%
  unlist(use.names = TRUE)

# Unique reporting areas
reporting_areas <- unique(site_means$reporting_area)

# Loop over each reporting area
for (area in reporting_areas) {
  area_sites <- site_means %>% filter(reporting_area == area)
  n_sites <- nrow(area_sites)
  n_pages <- ceiling(n_sites / 6)

  for (page in 1:n_pages) {
    # Subset to group of up to 6 sites
    start_i <- (page - 1) * 6 + 1
    end_i <- min(start_i + 5, n_sites)
    sites_subset <- area_sites[start_i:end_i, ]

    # Set output
    filename <- paste0("radar_group_", gsub(" ", "_", area), "_page_", page, ".png")
    png(filename, width = 1200, height = 800)
    par(mfrow = c(2, 3), mar = c(2, 2, 4, 2), oma = c(0, 0, 4, 0))  # 2x3 layout

    for (i in 1:nrow(sites_subset)) {
      site_data <- unlist(sites_subset[i, vars_to_use], use.names = TRUE)
      
      # Assemble radar data as data frame
      chart_data <- as.data.frame(rbind(global_max, global_min, reference_profile, site_data))
      rownames(chart_data) <- c("Max", "Min", "Reference", as.character(sites_subset$site_name[i]))

      radarchart(chart_data,
                 axistype = 1,
                 pcol = c("gray40", "steelblue"),
                 pfcol = c(rgb(0.4, 0.4, 0.4, 0.3), rgb(0.2, 0.5, 0.8, 0.4)),
                 plwd = 2,
                 plty = 1,
                 cglcol = "grey80",
                 title = sites_subset$site_name[i])
    }

    # Add group label at the top
    mtext(paste("Reporting Area:", area), side = 3, line = 1, outer = TRUE, cex = 1.5)
    dev.off()
  }
}

```





