---
title: "10.2- Running Stats at the Sample and Site Levels"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

# libraries
library(car) # VERIFY what this is for 
library(cluster) # grouping metrics - VERIFY still needed
library(devtools) # installing bespoke packages
library(dplyr) # data wrangling
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(fmsb) # polygon calculations for the radars
library(FSA) # post hoc test - Dunn's Test   
library(ggplot2) # plots
library(knitr) # output formatting
library(multcompView) # used for the letter assignment
library(pgirmess) # stats - KW
library(rcompanion) # for annotation of permanova
library(rstatix) # VERIFY what this is for      
library(RVAideMemoire) # post hoc test for permanova
library(scales) # scaling data for IBR - works with data wrangling packages
library(sf) # mapping - needed for converting to spatial data
library(spdep) # building spatial geometric components for analysis
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(vegan) # ecological stats 

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

```

# Load & Check Data
```{r}

getwd()
#setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels") # something here isn't working right - check out why

sample_data<- read.csv("../data/index_df/complete_sample_level.csv")
site_data<- read.csv("../data/index_df/complete_site_level.csv")

```

# Data prep
```{r}

# make site_number & reporting area a factor in both dfs

sample_data$reporting_area <- as.factor(sample_data$reporting_area)
site_data$reporting_area <- as.factor(site_data$reporting_area)

sample_data$site_number <- as.factor(sample_data$site_number)
site_data$site_number <- as.factor(site_data$site_number)

```

## add contaminant indices to sample- level df
### NOT needed - do not use
```{r}

site_df<- read.csv("../data/index_df/contaminant_indices_only.csv")

all_data_complete_indices <- all_data %>%
  left_join(site_df, by = "site_name")

head(all_data_complete_indices)

# write it out
#write.csv(all_data_complete_indices, "/Users/cmantegna/Documents/Github/WDFWmussels/data/index_df/complete_sample_level.csv", row.names = FALSE) 

```

# Normality, similarity and variance all tested in 10.1 to validate groupings

# KW 
## sample - level, run 5.11.25
### commented out lines indicate non- significant results
```{r}

#KW

kruskal.test(sod ~ site_name, data = sample_data) # p-value = 0.001777
#kruskal.test(sod ~ reporting_area, data = sample_data) # p-value = 0.3136

kruskal.test(shell ~ site_name, data = sample_data) # p-value = 1.092e-08
kruskal.test(shell ~ reporting_area, data = sample_data) # p-value = 1.545e-06

kruskal.test(ci1 ~ site_name, data = sample_data) # p-value = 0.001056
#kruskal.test(ci1 ~ reporting_area, data = sample_data) # p-value = 0.7855

kruskal.test(ci2 ~ site_name, data = sample_data) # p-value = 2.191e-08
#kruskal.test(ci2 ~ reporting_area, data = sample_data) # p-value = 0.3396

kruskal.test(ci3 ~ site_name, data = sample_data) # p-value = 0.0001364
kruskal.test(ci3 ~ reporting_area, data = sample_data) # p-value = 0.02201

kruskal.test(ibr_bio ~ site_name, data = sample_data) # p-value = 1.652e-05
kruskal.test(ibr_bio ~ reporting_area, data = sample_data) # p-value = 0.01583

kruskal.test(ibr_morph ~ site_name, data = sample_data) # p-value = 8.555e-05
kruskal.test(ibr_morph ~ reporting_area, data = sample_data) # p-value = 8.025e-05

kruskal.test(ibr_combined ~ site_name, data = sample_data) # p-value = 0.0002026
kruskal.test(ibr_combined ~ reporting_area, data = sample_data) # p-value = 0.002755

```

# Dunn's
## sod - site
```{r}
# not significant
dunn_result <- dunnTest(sod ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

```
## shell - site & reporting area
```{r}

library(multcompView) # used for the letter assignment

# site - significant
dunn_result <- dunnTest(shell ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letter designation
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}
cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)
dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

head(dunn_table_with_cld)
#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_shell_site.csv", row.names = FALSE) # write it out


# reporting area - significant
dunn_result <- dunnTest(shell ~ reporting_area, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)

print(dunn_table) # view
groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letter designation
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}
cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)
dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

head(dunn_table_with_cld)
#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_shell_ra.csv", row.names = FALSE) # write it out

```

## ci1 - site
```{r}

# site - significant
dunn_result <- dunnTest(ci1 ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letter designation
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}
cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)
dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

head(dunn_table_with_cld)
#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_ci1_site.csv", row.names = FALSE) # write it out

```

## ci2 - site
```{r}

# site - not significant
dunn_result <- dunnTest(ci2 ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)

print(dunn_table) # view

```

## ci3 - site & reporting area
```{r}

# site is not significant
dunn_result <- dunnTest(ci3 ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

# reporting area is significant (2)
dunn_result <- dunnTest(ci3 ~ reporting_area, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letter designation
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

head(dunn_table_with_cld)
#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_ci3_ra.csv", row.names = FALSE) # write it out

```

## ibr_bio - site & reporting area
```{r}

# site is not significant
dunn_result <- dunnTest(ibr_bio ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

# reporting area is significant
dunn_result <- dunnTest(ibr_bio ~ reporting_area, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letters
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)
dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)
head(dunn_table_with_cld)
#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_ibrbio_ra.csv", row.names = FALSE) # write it out

```

## ibr_morph - site & reporting area
```{r}

# site is not significant
dunn_result <- dunnTest(ibr_morph ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

# reporting area is significant
dunn_result <- dunnTest(ibr_morph ~ reporting_area, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letters
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}
cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_ibrmorph_ra.csv", row.names = FALSE) # write it out

```

## ibr_combined - site & reporting area
```{r}

# site is not significant
dunn_result <- dunnTest(ibr_combined ~ site_name, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view


# reporting area is significant
dunn_result <- dunnTest(ibr_combined ~ reporting_area, data = sample_data, method = "bh") 
dunn_df <- dunn_result$res %>%
  dplyr::select(Comparison, Z, P.unadj, P.adj) %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = " - ") %>%
  dplyr::mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))
dunn_table <- dunn_df %>%
  dplyr::mutate(across(c(P.unadj, P.adj), round, digits = 4)) %>%
  dplyr::arrange(P.adj)
print(dunn_table) # view

groups <- unique(c(dunn_df$Group1, dunn_df$Group2)) # add letters
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))
for (i in 1:nrow(dunn_df)) {
  g1 <- dunn_df$Group1[i]
  g2 <- dunn_df$Group2[i]
  p <- dunn_df$P.adj[i]
  pval_matrix[g1, g2] <- p
  pval_matrix[g2, g1] <- p
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters
cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)
dunn_table_with_cld <- dunn_table %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

#write.csv(dunn_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/dunn_ibrcombined_ra.csv", row.names = FALSE) # write it out

```

# ANOVA + post hoc
## individual sample - level
## p450
```{r}

anova_site <- aov(p450 ~ site_name, data = sample_data) # p= 1.25e-07
anova_ra <- aov(p450 ~ reporting_area, data = sample_data) # p= 3.83e-06

summary(anova_site)
summary(anova_ra)

#post - site is significant
tukey_site <- TukeyHSD(anova_site, "site_name")
tukey_df <- as.data.frame(tukey_site$site_name)

tukey_df <- tukey_df %>%
  tibble::rownames_to_column("Comparison") %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = "-") %>%
  rename(Diff = diff, Lower = lwr, Upper = upr, P.adj = `p adj`) %>%
  mutate(across(everything(), ~trimws(as.character(.)))) %>%  
  mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

groups <- unique(c(tukey_df$Group1, tukey_df$Group2))
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))

for (i in 1:nrow(tukey_df)) {
  g1 <- tukey_df$Group1[i]
  g2 <- tukey_df$Group2[i]
  pval <- as.numeric(tukey_df$P.adj[i])
  pval_matrix[g1, g2] <- pval
  pval_matrix[g2, g1] <- pval
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters

cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

tukey_table_with_cld <- tukey_df %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

view(tukey_table_with_cld) #view
#write.csv(tukey_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/tukey_p450_site.csv", row.names = FALSE) #save

#post - reporting area
tukey_ra <- TukeyHSD(anova_ra, "reporting_area")
tukey_df <- as.data.frame(tukey_ra$reporting_area)

tukey_df <- tukey_df %>%
  tibble::rownames_to_column("Comparison") %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = "-") %>%
  rename(Diff = diff, Lower = lwr, Upper = upr, P.adj = `p adj`) %>%
  mutate(across(everything(), ~trimws(as.character(.)))) %>%  
  mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

groups <- unique(c(tukey_df$Group1, tukey_df$Group2))
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))

for (i in 1:nrow(tukey_df)) {
  g1 <- tukey_df$Group1[i]
  g2 <- tukey_df$Group2[i]
  pval <- as.numeric(tukey_df$P.adj[i])
  pval_matrix[g1, g2] <- pval
  pval_matrix[g2, g1] <- pval
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters

cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

tukey_table_with_cld <- tukey_df %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

view(tukey_table_with_cld) #view
#write.csv(tukey_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/tukey_p450_ra.csv", row.names = FALSE) #save

```

# Correlations, Spearman & Kendalls
## site-level
### all indices
```{r}

colnames(site_data)
# Define your variable sets
metrics_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# List any ID columns to exclude here
id_columns <- c("site_name", "site_number",  "reporting_area", "latitude", "longitude", "heavy_metal", "nutrient_metal", "total_metal")

# Build analyte_vars by excluding both metrics and IDs
analyte_vars <- c("chlordane", "ddt", "hch", "pah_lmw", "pah_hmw", "pah_add", "total_pah", "pbde", "pcb", "pesticide", "mixture_index_sum", "heavy_metal_ng", "nutrient_metal_ng", "mixture_index_pca")

# set p-value & initialize empty list
alpha <- 0.05
results_list <- list()

# Loop over metric–analyte pairs across sites
for (metric in metrics_vars) {
  for (analyte in analyte_vars) {
    
    x <- site_data[[metric]]
    y <- site_data[[analyte]]
    
    valid_idx <- complete.cases(x, y)
    x_valid <- x[valid_idx]
    y_valid <- y[valid_idx]
    
    if (length(x_valid) < 4) next  # Skip if too few values
    
    spearman <- cor.test(x_valid, y_valid, method = "spearman")
    kendall <- cor.test(x_valid, y_valid, method = "kendall")
    
    results_list[[paste(metric, analyte, sep = "_")]] <- data.frame(
      Metric = metric,
      Analyte = analyte,
      Spearman_r = as.numeric(spearman$estimate),
      Spearman_p = as.numeric(spearman$p.value),
      Kendall_tau = as.numeric(kendall$estimate),
      Kendall_p = as.numeric(kendall$p.value)
    )
  }
}

# Combine results
corr_full <- bind_rows(results_list)

# Add significance stars
corr_full <- corr_full %>%
  mutate(
    Spearman_sig = case_when(
      Spearman_p <= 0.001 ~ "***",
      Spearman_p <= 0.01  ~ "**",
      Spearman_p <= 0.05  ~ "*",
      TRUE ~ ""
    ),
    Kendall_sig = case_when(
      Kendall_p <= 0.001 ~ "***",
      Kendall_p <= 0.01  ~ "**",
      Kendall_p <= 0.05  ~ "*",
      TRUE ~ ""
    )
  )

View(corr_full)
#write.csv(corr_full, "/Users/cmantegna/Documents/Github/WDFWmussels/output/correlation_all_indices.csv", row.names = FALSE)

```

### individual analytes
```{r}

#correlation_data<- read.csv("../data/index_df/site_level_all_analytes_scaled.csv")

metrics_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# identifying columns to exclude from analysis
id_columns <- c("site_name", "site_number",  "reporting_area", "latitude", "longitude", "ai_z_arsenic", "ai_z_cadmium", "ai_z_copper", "ai_z_lead", "ai_z_mercury", "ai_z_zinc")

# converting metals to match other contaminants - metals are reported in mg/kg and all others are ng/g
# 1 mg/kg = 1,000 ng/g
correlation_data$ai_z_arsenic_ng <- correlation_data$ai_z_arsenic * 1000
correlation_data$ai_z_cadmium_ng <- correlation_data$ai_z_cadmium * 1000
correlation_data$ai_z_copper_ng <- correlation_data$ai_z_copper * 1000
correlation_data$ai_z_lead_ng <- correlation_data$ai_z_lead * 1000
correlation_data$ai_z_mercury_ng <- correlation_data$ai_z_mercury * 1000
correlation_data$ai_z_zinc_ng <- correlation_data$ai_z_zinc * 1000

#write.csv(correlation_data, "/Users/cmantegna/Documents/Github/WDFWmussels/data/site_level_corrected_metals_all_analytes.csv", row.names = FALSE)

# Build analyte_vars by excluding both metrics and IDs
analyte_vars <- colnames(correlation_data)[!(colnames(correlation_data) %in% c(metrics_vars, id_columns))]

# set p-value & initialize empty list
alpha <- 0.05
results_list <- list()

# Loop over metric–analyte pairs across sites
for (metric in metrics_vars) {
  for (analyte in analyte_vars) {
    
    x <- correlation_data[[metric]]
    y <- correlation_data[[analyte]]
    
    valid_idx <- complete.cases(x, y)
    x_valid <- x[valid_idx]
    y_valid <- y[valid_idx]
    
    if (length(x_valid) < 4) next  # Skip if too few values
    
    spearman <- cor.test(x_valid, y_valid, method = "spearman")
    kendall <- cor.test(x_valid, y_valid, method = "kendall")
    
    results_list[[paste(metric, analyte, sep = "_")]] <- data.frame(
      Metric = metric,
      Analyte = analyte,
      Spearman_r = as.numeric(spearman$estimate),
      Spearman_p = as.numeric(spearman$p.value),
      Kendall_tau = as.numeric(kendall$estimate),
      Kendall_p = as.numeric(kendall$p.value)
    )
  }
}

# Combine results
corr_full <- bind_rows(results_list)

# Add significance stars
corr_full <- corr_full %>%
  mutate(
    Spearman_sig = case_when(
      Spearman_p <= 0.001 ~ "***",
      Spearman_p <= 0.01  ~ "**",
      Spearman_p <= 0.05  ~ "*",
      TRUE ~ ""
    ),
    Kendall_sig = case_when(
      Kendall_p <= 0.001 ~ "***",
      Kendall_p <= 0.01  ~ "**",
      Kendall_p <= 0.05  ~ "*",
      TRUE ~ ""
    )
  )

View(corr_full)
#write.csv(corr_full, "/Users/cmantegna/Documents/Github/WDFWmussels/output/correlation_individual_analytes.csv", row.names = FALSE)

```

# Spatial Analysis
## Global Moran's I
```{r}

# fix this in post - this can be a loop, but I will need help making it work properly; this is why it is manually done

# all measured metrics
# p450
site_data <- site_sf %>% filter(!is.na(p450)) # filter out any NAs
summary(site_data$p450) # check for variance and more than 3 rows for analysis
coords <- st_coordinates(site_data) # get coordinated from included sites
nb <- knn2nb(knearneigh(coords, k = 4)) # build neighbor list and weights
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$p450)[, 1] # scale results and run morans
moran.test(x_scaled, lw)

# sod
site_data <- site_sf %>% filter(!is.na(sod))
summary(site_data$sod)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$sod)[, 1]
moran.test(x_scaled, lw)

# shell
site_data <- site_sf %>% filter(!is.na(shell))
summary(site_data$shell)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$shell)[, 1]
moran.test(x_scaled, lw)

# ci1
site_data <- site_sf %>% filter(!is.na(ci1))
summary(site_data$ci1)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci1)[, 1]
moran.test(x_scaled, lw)

# ci2
site_data <- site_sf %>% filter(!is.na(ci2))
summary(site_data$ci2)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci2)[, 1]
moran.test(x_scaled, lw)

# ci3
site_data <- site_sf %>% filter(!is.na(ci3))
summary(site_data$ci3)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci3)[, 1]
moran.test(x_scaled, lw)


# ibr_bio
site_data <- site_sf %>% filter(!is.na(ibr_bio))
summary(site_data$ibr_bio)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_bio)[, 1]
moran.test(x_scaled, lw)

# ibr_morph
site_data <- site_sf %>% filter(!is.na(ibr_morph))
summary(site_data$ibr_morph)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_morph)[, 1]
moran.test(x_scaled, lw)

# ibr_combined
site_data <- site_sf %>% filter(!is.na(ibr_combined))
summary(site_data$ibr_combined)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_combined)[, 1]
moran.test(x_scaled, lw)

# All contaminant indices
# mixture index, sum
site_data <- site_sf %>% filter(!is.na(mixture_index_sum))
summary(site_data$mixture_index_sum)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$mixture_index_sum)[, 1]
moran.test(x_scaled, lw)

# mixture index, pca
site_data <- site_sf %>% filter(!is.na(mixture_index_pca))
summary(site_data$mixture_index_pca)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$mixture_index_pca)[, 1]
moran.test(x_scaled, lw)

#chlordane
site_data <- site_sf %>% filter(!is.na(chlordane))
summary(site_data$chlordane)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$chlordane)[, 1]
moran.test(x_scaled, lw)

# UPDATE the CODE from HERE
#ddt
site_data <- site_sf %>% filter(!is.na(ddt))
summary(site_data$ddt)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ddt)[, 1]
moran.test(x_scaled, lw)

#hch
site_data <- site_sf %>% filter(!is.na(hch))
summary(site_data$hch)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$hch)[, 1]
moran.test(x_scaled, lw)

#metal - nutrients
site_data <- site_sf %>% filter(!is.na(nutrient_metal_ng))
summary(site_data$nutrient_metal_ng)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$nutrient_metal_ng)[, 1]
moran.test(x_scaled, lw)

#metal - heavy
site_data <- site_sf %>% filter(!is.na(heavy_metal_ng))
summary(site_data$heavy_metal_ng)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$heavy_metal_ng)[, 1]
moran.test(x_scaled, lw)

#pah_lmw
site_data <- site_sf %>% filter(!is.na(pah_lmw))
summary(site_data$pah_lmw)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_lmw)[, 1]
moran.test(x_scaled, lw)

#pah_hmw
site_data <- site_sf %>% filter(!is.na(pah_hmw))
summary(site_data$pah_hmw)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_hmw)[, 1]
moran.test(x_scaled, lw)

#pah_add
site_data <- site_sf %>% filter(!is.na(pah_add))
summary(site_data$pah_add)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_add)[, 1]
moran.test(x_scaled, lw)

#pbde
site_data <- site_sf %>% filter(!is.na(pbde))
summary(site_data$pbde)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pbde)[, 1]
moran.test(x_scaled, lw)

#pcb
site_data <- site_sf %>% filter(!is.na(pcb))
summary(site_data$pcb)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pcb)[, 1]
moran.test(x_scaled, lw)

#pesticides
site_data <- site_sf %>% filter(!is.na(pesticide))
summary(site_data$pesticide)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pesticide)[, 1]
moran.test(x_scaled, lw)

```

## LISA - Local Indicators of Spatial Association
```{r}

library(sf)
library(spdep)

# convert df to a spatial object
site_sf <- st_as_sf(site_data, coords = c("longitude", "latitude"), crs = 4326)
site_sf <- st_transform(site_sf, crs = 3857)  # projected CRS for distance-based neighbors

# create spatial neighbors and weights (4 nearest neighbors)
coords <- st_coordinates(site_sf)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

# response variables to test
bio_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# contaminant indices 
index_vars <- c("chlordane", "ddt", "hch", "pah_lmw", "pah_hmw", "pah_add", "pbde", "pcb", "pesticide", "heavy_metal_ng", "nutrient_metal_ng", "mixture_index_sum", "mixture_index_pca")

# LOOP FOR LOCAL MORAN'S I 

# create output folder
dir.create("lisa_outputs", showWarnings = FALSE)

for (var in c(bio_vars, index_vars)) {
  x <- site_sf[[var]]
  lisa <- localmoran(scale(x)[,1], lw)

  # Add LISA results to spatial dataframe
  site_sf[[paste0(var, "_lisa_I")]] <- lisa[,1]
  site_sf[[paste0(var, "_lisa_p")]] <- lisa[,5]

  # Define cluster types
  site_sf[[paste0(var, "_lisa_cluster")]] <- case_when(
    x > mean(x, na.rm = TRUE) & lisa[,1] > 0 ~ "High-High",
    x < mean(x, na.rm = TRUE) & lisa[,1] > 0 ~ "Low-Low",
    lisa[,1] < 0 ~ "Outlier",
    TRUE ~ "Not Significant"
  )
  
  # Save spatial cluster results (optional)
  st_write(site_sf[, c("site_name", paste0(var, "_lisa_cluster"))],
           paste0("lisa_outputs/LISA_", var, ".geojson"), delete_dsn = TRUE)
}

```

## plotting LISA output
```{r}

getwd()
library(sf)
lisa_p450<- st_read("lisa_outputs/LISA_p450.geojson")
names(lisa_p450)
head(lisa_p450)

```

