---
title: "11- Visualizations"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

# libraries
library(car) # VERIFY what this is for 
library(cluster) # grouping metrics - VERIFY still needed
library(devtools) # installing bespoke packages
library(dplyr) # data wrangling
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(fmsb) # polygon calculations for the radars
library(FSA) # post hoc test - Dunn's Test   
library(ggplot2) # plots
library(knitr) # output formatting
library(multcompView) # used for the letter assignment
library(pgirmess) # stats - KW
library(rcompanion) # for annotation of permanova
library(rstatix) # VERIFY what this is for      
library(RVAideMemoire) # post hoc test for permanova
library(scales) # scaling data for IBR - works with data wrangling packages
library(sf) # mapping - needed for converting to spatial data
library(spdep) # building spatial geometric components for analysis
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(vegan) # ecological stats 

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

```

# Load & Check Data
```{r}

getwd()
#setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels") # something here isn't working right - check out why

sample_data<- read.csv("../data/index_df/complete_sample_level.csv")
site_data<- read.csv("../data/index_df/site_with_mixture_index.csv")

```

# Data prep
```{r}

# make site_number & reporting area a factor in both dfs

sample_data$reporting_area <- as.factor(sample_data$reporting_area)
site_data$reporting_area <- as.factor(site_data$reporting_area)

sample_data$site_number <- as.factor(sample_data$site_number)
site_data$site_number <- as.factor(site_data$site_number)

```

# Don't forget to update this radar plot code
```{r}

## Radars, reference overlay - reference line 757 in file 99
## All Sites, Do Not Run


z_scores$sample_id <- df$sample_id
z_scores$site_name <- df$site_name

# Site-averaged Z-scores
site_means <- z_scores %>%
  group_by(site_name) %>%
  summarise(across(1:8, mean, na.rm = TRUE))

# Loop through sites
for (i in 1:nrow(site_means)) {
  site_data <- site_means[i, 2:9]
  max_min <- rbind(apply(site_means[ ,2:9], 2, max),
                   apply(site_means[ ,2:9], 2, min))
  radar_data <- rbind(max_min, site_data)
  
  png(paste0("IBRv2i_radar_site_", gsub(" ", "_", site_means$site_name[i]), ".png"), width = 600, height = 600)
  radarchart(radar_data,
             axistype = 1,
             pcol = "darkblue", pfcol = rgb(0.2, 0.4, 0.9, 0.4),
             plwd = 2,
             title = paste("IBRv2i Radar Plot\nSite:", site_means$site_name[i]))
  dev.off()
}

```

# Box plots by reporting area - sample- level
## Facet wrapped for review - not reporting
### biomarkers 
```{r}

# biomarkers - p450 & sod
bio_df <- sample_data %>%
  select(site_name, reporting_area, p450, sod) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Biomarkers, Scaled", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_biomarkers_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

### morphometrics
```{r}

# morphometrics - shell, ci1, ci3
bio_df <- sample_data %>%
  select(site_name, reporting_area, shell, ci1, ci3) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Morphometrics, Scaled", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_morphometrics_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

### raw measurements - width, height, length
```{r}

# width, height, length
bio_df <- sample_data %>%
  select(site_name, reporting_area, height, length, width) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Raw Measurements (height, length, width), Scaled", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_measurements_hlw_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

### raw measurements - weights
```{r}

# weights - note the misspelling of initial weight column name, fix it
bio_df <- sample_data %>%
  select(site_name, reporting_area, weigtht_initial, weight_final, weight_change) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Raw Measurements (weights), Scaled", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_measurements_weights_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

### IBR
```{r}

# IBRs
bio_df <- sample_data %>%
  select(site_name, reporting_area, ibr_bio, ibr_morph, ibr_combined) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Integrated Biomarker Response, Scores", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_ibr_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

# Box plots, sample- level by site
## IBRs - facet wrapped plots
```{r}

# Reshape IBR data to long format
ibr_long <- sample_data %>%
  pivot_longer(cols = c(ibr_bio, ibr_morph, ibr_combined),
               names_to = "ibr_type",
               values_to = "ibr_value") %>%
  mutate(ibr_type = recode(ibr_type,
                           "ibr_bio" = "Biomarker",
                           "ibr_morph" = "Morphometric",
                           "ibr_combined" = "Combined"))

ibr_long$ibr_type <- factor(ibr_long$ibr_type,
                            levels = c("Biomarker", "Morphometric", "Combined"))

# Plot by site
ibr_site<- ggplot(ibr_long, aes(x = site_name, y = ibr_value, fill = ibr_type)) +
  geom_boxplot() +
  #facet_wrap(~ibr_type, scales = "free") +
  theme_bw() +
  labs(x = "Site", y = "IBR Score", title = "IBR Scores by Site") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot by reporting area
ibr_ra<- ggplot(ibr_long, aes(x = reporting_area, y = ibr_value, fill = ibr_type, 
                              color = site_name == "Penn Cove Reference")) +
  geom_boxplot() +
  facet_wrap(~ ibr_type, ncol = 1, scales = "fixed") +  # change to "fixed" if comparing directly
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_bw() +
  labs(x = "Reporting Area", y = "IBR Score", title = "IBR Scores by Reporting Area") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# save and view
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/all_ibr_ra_facet_boxplots.png", plot= ibr_ra, width = 16, height = 14, dpi = 300)

print(ibr_site)
print(ibr_ra)

```

## IBRs 
```{r}
# ibr with Penn Cove reference site
ibr_long$ibr_type <- factor(ibr_long$ibr_type,
                            levels = c("Biomarker", "Morphometric", "Combined"))

all_index <- ggplot(ibr_long, aes(x = site_name, y = ibr_value,
                                  color = site_name == "Penn Cove Reference")) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  facet_wrap(~ ibr_type, ncol = 1, scales = "fixed") +  # change to "fixed" if comparing directly
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "IBRv2i Indices by Site (Reference site in red)",
       x = "Site", y = "IBRv2i Score")


# save and view
ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/all_ibr_boxplot.png", plot= all_index, width = 16, height = 14, dpi = 300)

print(all_index)

```


# Boxplots, site-level
## double check i can do a box plot by site or must group by reporting area since it's only one value per site
### contaminant indices, 1 of 
```{r}

# chlordanes, hch, 
# ddt, pesticides
# all 3 metals
# all 3 pahs
# pbde, pcb
# mixtures

bio_df <- sample_data %>%
  select(site_name, reporting_area, height, length, width) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Raw Measurements (height, length, width), Scaled", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

# view & save
print(facet_bio)
#ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/may_figures/boxplot_measurements_hlw_ra_facet.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

```

```{r}
# ========== Morphometrics & IBR (morph) ==========
morph_df <- df %>%
  select(site_name, reporting_area, shell, ci1, ibr1morph, ibr1overall) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_morph<- ggplot(morph_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Morphometrics and IBRs", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/morph_metrics_facet_boxplot.png", plot= facet_morph, width = 16, height = 14, dpi = 300)

print(facet_morph)

```

#UPDATE THIS
# Box plot, facet wrapped by reporting area
## Metrics and IBRs grouped by reporting area
```{r}

# prep
df$reporting_area <- as.factor(df$reporting_area)
df <- df %>%
  mutate(across(where(is.character), str_trim))


# ========== Functional Biomarkers & IBR (bio) ==========
bio_df <- df %>%
  select(site_name, reporting_area, p450, sod, ibr1bio, ibr1overall) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_bio<- ggplot(bio_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Biomarkers and IBRs", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/bio_metrics_facet_boxplot.png", plot= facet_bio, width = 16, height = 14, dpi = 300)

print(facet_bio)

# ========== Morphometrics & IBR (morph) ==========
morph_df <- df %>%
  select(site_name, reporting_area, shell, ci1, ibr1morph, ibr1overall) %>%
  pivot_longer(cols = -c(site_name, reporting_area), 
               names_to = "metric", 
               values_to = "value") %>%
  mutate(site_name = factor(site_name, levels = sort(unique(site_name))))

facet_morph<- ggplot(morph_df, aes(x = site_name, y = value, fill = metric)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  facet_wrap(~ reporting_area, scales = "free_x") +
  labs(title = "Morphometrics and IBRs", 
       x = "Site", y = "Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

ggsave(filename= "/Users/cmantegna/Documents/GitHub/WDFWmussels/output/morph_metrics_facet_boxplot.png", plot= facet_morph, width = 16, height = 14, dpi = 300)

print(facet_morph)

```


# Pulling Mapping Data from USGS
```{r}

download.file(
  url = "https://prd-tnm.s3.amazonaws.com/index.html?prefix=StagedProducts/Elevation/1m/Projects/WA_Western_North_2016/TIFF/",
  destfile = "wa_north_DEM.tif",
  mode = "wb"
)


```

