---
title: "12- Creating Visualzations of Stats Results"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

# libraries
library(knitr) # output fotmatting
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(dplyr) # data wrangling
library(vegan) # ecological stats 
library(cluster) # grouping metrics - VERIFY still needed
library(pgirmess) # stats - KW
library(ggplot2) # plots
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(FSA) # post hoc test - Dunn's Test       
library(rstatix) # VERIFY what this is for      
library(car) # VERIFY what this is for  
library(RVAideMemoire) # post hoc test for permanova
library(rcompanion) # for annotation of permanova
library(scales) # scaling data for IBR - works with data wrangling packages
library(fmsb) # polygon calculations for the radars
library(sf) # spatial plotting
library(spdep) # spatial plotting
library(tmap) # making maps
library(patchwork) # making pretty grids of plots
library(viridis) # colors for plotting
library(rnaturalearth) # to pull the map
library(ggsignif) # adding significance indicators to plots

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

```

# Load & Check Data
```{r}

getwd()
#setwd("/Users/cmantegna/Documents/GitHub/WDFWmussels") # something here isn't working right - check out why

ibr_avg<- read.csv("../data/cleaned/avg_ibr1.csv") # penn cove reference site only
analyte_long<- read.csv("../data/cleaned/analytes_long.csv")
metal_long<- read.csv("../data/cleaned/metals_long.csv")
ibr_analyte<- read.csv("../data/cleaned/avg_ibr1_analytess.csv")
ibr_metal<- read.csv("../data/cleaned/avg_ibr1_metals.csv")
#moran_analyte<- read.csv("../output/tables/moran_analyte.csv")
#moran_metal<- read.csv("../output/tables/moran_metal.csv")
#spatial_analyte<- read.csv("../output/tables/spatial_correlation_biomarkers_analytes.csv")
#spatial_metal<- read.csv("../output/tables/moran_analyte.csv")

# clean up names if needed
#df1 <- df1 %>%
#  mutate(across(where(is.character), trimws))


```

# Moran's Plots
## data prep
```{r}

# add significance flag
moran_analyte <- moran_analyte %>%
  mutate(significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))

spatial_analyte <- spatial_analyte %>%
  mutate(significant = ifelse(p_value < 0.05, "Significant", "Not Significant"))


```

## contaminant class, bar plot 
```{r}

ggplot(moran_analyte, aes(x = reorder(class, moran_i), y = moran_i, fill = significant)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Significant" = "darkred", "Not Significant" = "gray")) +
  labs(title = "Moran's I for Contaminant Classes",
       x = "Contaminant Class",
       y = "Moran's I",
       fill = "Significance") +
  theme_minimal()

```

## biomarkers + morphometrics, bar plot
```{r}

ggplot(spatial_analyte, aes(x = reorder(metric, moran_i), y = moran_i, fill = significant)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Significant" = "darkblue", "Not Significant" = "gray")) +
  labs(title = "Moran's I for Biomarkers and Morphometrics",
       x = "Metric",
       y = "Moran's I",
       fill = "Significance") +
  theme_minimal()

```

## combined comparison plot
```{r}

# Combine and format both datasets
combined_df <- bind_rows(
  moran_analyte %>%
    rename(name = class) %>%
    mutate(group = "Analyte"),
  spatial_analyte %>%
    rename(name = metric) %>%
    mutate(group = "Biomarker/Morphometric")
)

ggplot(combined_df, aes(x = reorder(name, moran_i), y = moran_i, fill = group)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Comparison of Moran's I: Analytes vs Biomarkers/Morphometrics",
       x = "Variable",
       y = "Moran's I",
       fill = "Group") +
  theme_minimal()

```

## LISA map of Moran's results
### data prep
```{r}

# appending chemical class to the analyte names so that I can group classes for plotting & converting to spatial data

analyte_long <- analyte_long %>%
  mutate(analyte_class = paste0(analyte, "_", class))

analyte_wide <- analyte_long %>%
  select(site_name, latitude, longitude, analyte_class, dry_value) %>%
  pivot_wider(names_from = analyte_class, values_from = dry_value)

```

### creating WA basemap
```{r}

# convert to spatial
analyte_sf <- st_as_sf(analyte_wide, coords = c("longitude", "latitude"), crs = 4326)

coast <- ne_states(country = "united states of america", returnclass = "sf") %>%
  filter(name == "Washington")

# major city coordinates for labels
cities <- data.frame(
  city = c("Seattle", "Tacoma", "Olympia", "Bellingham", "Port Angeles", "Hama Hama"),
  lon = c(-122.33, -122.45, -122.90, -122.48, -123.43, -123.16),
  lat = c(47.61, 47.25, 47.00, 48.73, 48.08, 47.63)
)

```

## maps for each analyte class - this is not working correctly yet
```{r}

# Set spatial weights 
#coords <- st_coordinates(analyte_sf)
#nb <- dnearneigh(coords, d1 = 0, d2 = 50000)  # adjust d2 as needed
#lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Extract all analyte_class column names
analyte_columns <- colnames(analyte_sf)[!(colnames(analyte_sf) %in% c("site_name", "geometry"))]

# Filter analyte columns to avoid empty or non-numeric ones
analyte_columns_filtered <- analyte_columns %>%
  keep(~ is.numeric(analyte_sf[[.x]]) && sum(!is.na(analyte_sf[[.x]])) >= 4)

###### troubleshooting the LISA setup for the map
#table(card(nb))  # This will show how many sites have 0, 1, 2... neighbors
#sd(analyte_sf$acenaphthene_lmw, na.rm = TRUE)
#all(st_is_valid(analyte_sf))
#nrow(analyte_sf) == nrow(st_coordinates(analyte_sf))
#colnames(analyte_sf) <- make.names(colnames(analyte_sf))
####### the names that start with numbers may have been throwing this whole thing off

base_map<- coast

make_lisa_plot <- function(data_full, analyte_name, base_map, save_path = "/Users/cmantegna/Documents/GitHub/WDFWmussels/output") {
  # Subset data to non-NA values for the current analyte
  data <- data_full %>% filter(!is.na(.data[[analyte_name]]))
  
  if (nrow(data) < 4) {
    message("Skipping ", analyte_name, ": too few valid rows.")
    return(NULL)
  }
  
  # Recalculate spatial weights
  coords <- st_coordinates(data)
  nb <- dnearneigh(coords, d1 = 0, d2 = 50000)
  if (any(card(nb) == 0)) {
    message("Skipping ", analyte_name, ": some points have no neighbors.")
    return(NULL)
  }
  lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
  
  # Run local Moran's I
  result <- tryCatch({
    localmoran(data[[analyte_name]], lw, zero.policy = TRUE)
  }, error = function(e) {
    message("Error running localmoran for ", analyte_name, ": ", e$message)
    return(NULL)
  })
  
  if (is.null(result) || !"Pr(z > 0)" %in% colnames(result)) {
    message("No valid output from localmoran for: ", analyte_name)
    return(NULL)
  }
  
  # Add results
  data$Ii <- result[, "Ii"]
  data$p_value <- result[, "Pr(z > 0)"]
  
  # Extract analyte and class from the name
  parts <- str_split(analyte_name, "_", simplify = TRUE)
  analyte <- parts[1]
  chem_class <- parts[ncol(parts)]  # assumes class is at the end
  
  # Classify clusters
  data <- data %>%
    mutate(cluster = case_when(
      .data[[analyte_name]] > mean(.data[[analyte_name]], na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "High-High",
      .data[[analyte_name]] < mean(.data[[analyte_name]], na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "Low-Low",
      .data[[analyte_name]] > mean(.data[[analyte_name]], na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "High-Low",
      .data[[analyte_name]] < mean(.data[[analyte_name]], na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "Low-High",
      TRUE ~ "Not Significant"
    ))
  
  # Plot
  p <- ggplot() +
    geom_sf(data = base_map, fill = "white", color = "black") +
    geom_sf(data = data, aes(color = cluster), size = 3) +
    scale_color_manual(values = c(
      "High-High" = "red",
      "Low-Low" = "blue",
      "High-Low" = "orange",
      "Low-High" = "lightblue",
      "Not Significant" = "gray"
    )) +
    labs(
      title = paste("LISA Cluster Map for", analyte, "| Class:", chem_class),
      color = "Cluster"
    ) +
    theme_minimal()
  
  print(p)
  
  save_path<- "/Users/cmantegna/Documents/GitHub/WDFWmussels/code"
  # Save if save_path is provided
  if (!is.null(save_path)) {
    filename <- paste0("lisa_", analyte_name, ".png")
    ggsave(filename = file.path(save_path, filename), plot = p, width = 8, height = 6)
  }
}

# Create output folder
output_folder <- "LISA_Plots"
dir.create(output_folder, showWarnings = FALSE)

# Loop through all analyte columns
for (analyte in analyte_columns_filtered) {
  message("Running: ", analyte)
  make_lisa_plot(analyte_sf, analyte, base_map = coast, save_path = output_folder)
}

```

## maps for the metals
### data prep
```{r}

metal_long <- metal_long %>%
  mutate(analyte = make.names(analyte))

# Pivot to wide format
metals_wide <- metal_long %>%
  select(site_name, latitude, longitude, analyte, dry_value) %>%
  pivot_wider(names_from = analyte, values_from = dry_value) %>%
  distinct()

# Check column names
colnames(metals_wide)

```

```{r}

#metals_sf <- st_as_sf(metals_wide, coords = c("longitude", "latitude"), crs = 4326)
#coords <- st_coordinates(metals_sf)
#nb <- dnearneigh(coords, d1 = 0, d2 = 50000)  # Adjust d2 as needed
#table(card(nb))  # Confirm no 0s

#lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Get list of all analyte columns
analyte_cols <- colnames(metals_sf)[!(colnames(metals_sf) %in% c("site_name", "geometry"))]

# Create output folder
output_folder <- "Metals_LISA_Plots"
dir.create(output_folder, showWarnings = FALSE)

# Loop over each analyte
for (analyte in analyte_cols) {
  message("Running LISA for: ", analyte)

  x <- metals_sf[[analyte]]

  if (sum(!is.na(x)) < 4 || sd(x, na.rm = TRUE) == 0) {
    message("Skipping ", analyte, ": too few values or no variation.")
    next
  }

  result <- tryCatch({
    localmoran(x, lw, zero.policy = TRUE)
  }, error = function(e) {
    message("Error running localmoran for ", analyte, ": ", e$message)
    return(NULL)
  })

  # Stop here if localmoran failed or malformed
  if (is.null(result) || !"Pr(z > 0)" %in% colnames(result)) {
    message("Skipping ", analyte, ": LISA output missing expected columns.")
    next
  }

  # Safe to proceed
  metals_sf$Ii <- result[, "Ii"]
  metals_sf$p_value <- result[, "Pr(z > 0)"]

  metals_sf <- metals_sf %>%
    mutate(cluster = case_when(
      .data[[analyte]] > mean(.data[[analyte]], na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "High-High",
      .data[[analyte]] < mean(.data[[analyte]], na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "Low-Low",
      .data[[analyte]] > mean(.data[[analyte]], na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "High-Low",
      .data[[analyte]] < mean(.data[[analyte]], na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "Low-High",
      TRUE ~ "Not Significant"
    ))

  p <- ggplot() +
    geom_sf(data = coast, fill = "white", color = "black") +
    geom_sf(data = metals_sf, aes(color = cluster), size = 3) +
    scale_color_manual(values = c(
      "High-High" = "red",
      "Low-Low" = "blue",
      "High-Low" = "orange",
      "Low-High" = "lightblue",
      "Not Significant" = "gray"
    )) +
    labs(
      title = paste("LISA Cluster Map for", analyte),
      color = "Cluster"
    ) +
    theme_minimal()

  ggsave(
    filename = file.path(output_folder, paste0("lisa_", analyte, ".png")),
    plot = p, width = 8, height = 6
  )
}

```

### troubleshooting metals - same problems as the analytes results above
```{r}

#using zinc since i know it was a significant moran's result
summary(metals_sf$zinc)
sd(metals_sf$zinc, na.rm = TRUE)

z <- metals_sf$zinc
result <- localmoran(z, lw, zero.policy = TRUE)
print(result)
colnames(result)

unique(metals_sf$zinc)
summary(metals_sf$zinc)
sd(metals_sf$zinc, na.rm = TRUE)


```

### new metals LISA
```{r}

# Recalculate neighbors and weights
coords <- st_coordinates(metals_sf)
nb <- dnearneigh(coords, d1 = 0, d2 = 25000)  # try a shorter radius like 25 km
table(card(nb))

# Binary style weights
lw_bin <- nb2listw(nb, style = "B", zero.policy = TRUE)

# Try again for zinc
z <- metals_sf$zinc
z_result <- localmoran(z, lw_bin, zero.policy = TRUE)

# Inspect result
head(z_result)
summary(z_result)
str(z_result)


```

```{r}

metals_sf$Ii <- z_result[, "Ii"]
metals_sf$p_value <- z_result[, "Pr(z > 0)"]

metals_sf <- metals_sf %>%
  mutate(cluster = case_when(
    zinc > mean(zinc, na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "High-High",
    zinc < mean(zinc, na.rm = TRUE) & Ii > 0 & p_value < 0.05 ~ "Low-Low",
    zinc > mean(zinc, na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "High-Low",
    zinc < mean(zinc, na.rm = TRUE) & Ii < 0 & p_value < 0.05 ~ "Low-High",
    TRUE ~ "Not Significant"
  ))

ggplot() +
  geom_sf(data = coast, fill = "white", color = "black") +
  geom_sf(data = metals_sf, aes(color = cluster), size = 3) +
  scale_color_manual(values = c(
    "High-High" = "red", "Low-Low" = "blue",
    "High-Low" = "orange", "Low-High" = "lightblue",
    "Not Significant" = "gray"
  )) +
  labs(title = "Zinc LISA with Binary Weights (25 km)", color = "Cluster") +
  theme_minimal()


```




