---
title: "3.0- Analysis without indices, site-level"
output:
  pdf_document: 
    fig_width: 20
    fig_height: 9
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE, echo=FALSE}

knitr::opts_chunk$set(
  root.dir = here::here(),
  echo = TRUE,         # show code chunks
  eval = TRUE,         # evaluate code chunks
  warning = FALSE,     # hide warnings
  message = FALSE,     # hide messages
  #fig.width = 15,       # set plot width in inches
  #fig.height = 9,      # set plot height in inches
  fig.align = "center" # slign plots to the center in output doc/ slide/ whatever
)

# libraries
library(car) # VERIFY what this is for 
library(cluster) # grouping metrics - VERIFY still needed
library(devtools) # installing bespoke packages
library(dplyr) # data wrangling
library(factoextra) # pca/ nmds/ tweaking radars
library(FactoMineR) # pca/ nmds/ tweaking radars
library(fmsb) # polygon calculations for the radars
library(FSA) # post hoc test - Dunn's Test   
library(ggplot2) # plots
library(knitr) # output formatting
library(multcompView) # used for the letter assignment
library(pgirmess) # stats - KW
library(rcompanion) # for annotation of permanova
library(rstatix) # VERIFY what this is for      
library(RVAideMemoire) # post hoc test for permanova
library(scales) # scaling data for IBR - works with data wrangling packages
library(sf) # mapping - needed for converting to spatial data
library(spdep) # building spatial geometric components for analysis
library(tidyr) # data wrangling
library(tidyverse) # data wrangling
library(vegan) # ecological stats 

```

# Load & Check Data
```{r}

getwd()

site_data<- read.csv("../data/start_here/avg_site_level_no_indices.csv")

```

# Data prep
```{r}

# make reporting area a factor 
site_data$reporting_area <- as.factor(site_data$reporting_area)
site_data$site_name <- as.factor(site_data$site_name)

```

## add contaminant indices to sample- level df
### NOT needed - do not use
```{r}

site_df<- read.csv("../data/index_df/contaminant_indices_only.csv")

all_data_complete_indices <- all_data %>%
  left_join(site_df, by = "site_name")

head(all_data_complete_indices)

# write it out
#write.csv(all_data_complete_indices, "/Users/cmantegna/Documents/Github/WDFWmussels/data/index_df/complete_sample_level.csv", row.names = FALSE) 

```

# KW 
## site & reporting area, run 9.26.25
### no significant result
```{r}

# both biomarkers
kruskal.test(p450 ~ sod, data = site_data) # p-value = 0.478

# p450
kruskal.test(p450 ~ site_name, data = site_data) # p-value = 0.478
kruskal.test(p450 ~ reporting_area, data = site_data) # p-value = 0.07831

# sod
kruskal.test(sod ~ site_name, data = site_data) # p-value = 0.478
kruskal.test(sod ~ reporting_area, data = site_data) # p-value = 0.7474

# ci1
kruskal.test(ci1 ~ site_name, data = site_data) # p-value = 0.478
kruskal.test(ci1 ~ reporting_area, data = site_data) # p-value = 0.9762

# ci2
kruskal.test(ci2 ~ site_name, data = site_data) # p-value = 0.478
kruskal.test(ci2 ~ reporting_area, data = site_data) # p-value = 0.821

```

# ANOVA + post hoc
## shell and ci3
```{r}

#shell
anova_site <- aov(shell ~ site_name, data = site_data) # p-value= 0.01119
anova_ra <- aov(shell ~ reporting_area, data = site_data) # p-value= 0.0312
summary(anova_site)
summary(anova_ra)

#ci3
anova_site <- aov(ci3 ~ site_name, data = site_data) # p-value= 1.923e-5
anova_ra <- aov(ci3 ~ reporting_area, data = site_data) # p-value= 0.386
summary(anova_site)
summary(anova_ra)

#post - shell, site=  not significant
tukey_site <- TukeyHSD(anova_site, "site_name")
tukey_df <- as.data.frame(tukey_site$site_name)

tukey_df <- tukey_df %>%
  tibble::rownames_to_column("Comparison") %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = "-") %>%
  rename(Diff = diff, Lower = lwr, Upper = upr, P.adj = `p adj`) %>%
  mutate(across(everything(), ~trimws(as.character(.)))) %>%  
  mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

groups <- unique(c(tukey_df$Group1, tukey_df$Group2))
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))

for (i in 1:nrow(tukey_df)) {
  g1 <- tukey_df$Group1[i]
  g2 <- tukey_df$Group2[i]
  pval <- as.numeric(tukey_df$P.adj[i])
  pval_matrix[g1, g2] <- pval
  pval_matrix[g2, g1] <- pval
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters

cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

tukey_table_with_cld <- tukey_df %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

view(tukey_table_with_cld) #view
# write.csv(tukey_table_with_cld, "../output/tukey_shell_site.csv", row.names = FALSE) #save

#post - shell=  not significant
tukey_ra <- TukeyHSD(anova_ra, "reporting_area")
tukey_df <- as.data.frame(tukey_ra$reporting_area)

tukey_df <- tukey_df %>%
  tibble::rownames_to_column("Comparison") %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = "-") %>%
  rename(Diff = diff, Lower = lwr, Upper = upr, P.adj = `p adj`) %>%
  mutate(across(everything(), ~trimws(as.character(.)))) %>%  
  mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

groups <- unique(c(tukey_df$Group1, tukey_df$Group2))
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))

for (i in 1:nrow(tukey_df)) {
  g1 <- tukey_df$Group1[i]
  g2 <- tukey_df$Group2[i]
  pval <- as.numeric(tukey_df$P.adj[i])
  pval_matrix[g1, g2] <- pval
  pval_matrix[g2, g1] <- pval
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters

cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

tukey_table_with_cld <- tukey_df %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

view(tukey_table_with_cld) #view
# write.csv(tukey_table_with_cld, "../output/tukey_shell_ra.csv", row.names = FALSE) #save

#post - reporting area
tukey_ra <- TukeyHSD(anova_ra, "reporting_area")
tukey_df <- as.data.frame(tukey_ra$reporting_area)

tukey_df <- tukey_df %>%
  tibble::rownames_to_column("Comparison") %>%
  tidyr::separate(Comparison, into = c("Group1", "Group2"), sep = "-") %>%
  rename(Diff = diff, Lower = lwr, Upper = upr, P.adj = `p adj`) %>%
  mutate(across(everything(), ~trimws(as.character(.)))) %>%  
  mutate(Significant = ifelse(P.adj < 0.05, "Yes", "No"))

groups <- unique(c(tukey_df$Group1, tukey_df$Group2))
pval_matrix <- matrix(1, nrow = length(groups), ncol = length(groups),
                      dimnames = list(groups, groups))

for (i in 1:nrow(tukey_df)) {
  g1 <- tukey_df$Group1[i]
  g2 <- tukey_df$Group2[i]
  pval <- as.numeric(tukey_df$P.adj[i])
  pval_matrix[g1, g2] <- pval
  pval_matrix[g2, g1] <- pval
}

cld_letters <- multcompLetters(pval_matrix < 0.05)$Letters

cld_df <- data.frame(
  Group = names(cld_letters),
  CLD = cld_letters,
  stringsAsFactors = FALSE
)

tukey_table_with_cld <- tukey_df %>%
  left_join(cld_df, by = c("Group1" = "Group")) %>%
  rename(Group1_CLD = CLD) %>%
  left_join(cld_df, by = c("Group2" = "Group")) %>%
  rename(Group2_CLD = CLD)

view(tukey_table_with_cld) #view
#write.csv(tukey_table_with_cld, "/Users/cmantegna/Documents/Github/WDFWmussels/output/tukey_p450_ra.csv", row.names = FALSE) #save

```

# Correlations, Spearman & Kendalls
## site-level
### all indices
```{r}

colnames(site_data)
# Define your variable sets
metrics_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# List any ID columns to exclude here
id_columns <- c("site_name", "site_number",  "reporting_area", "latitude", "longitude", "heavy_metal", "nutrient_metal", "total_metal")

# Build analyte_vars by excluding both metrics and IDs
analyte_vars <- c("chlordane", "ddt", "hch", "pah_lmw", "pah_hmw", "pah_add", "total_pah", "pbde", "pcb", "pesticide", "mixture_index_sum", "heavy_metal_ng", "nutrient_metal_ng", "mixture_index_pca")

# set p-value & initialize empty list
alpha <- 0.05
results_list <- list()

# Loop over metric–analyte pairs across sites
for (metric in metrics_vars) {
  for (analyte in analyte_vars) {
    
    x <- site_data[[metric]]
    y <- site_data[[analyte]]
    
    valid_idx <- complete.cases(x, y)
    x_valid <- x[valid_idx]
    y_valid <- y[valid_idx]
    
    if (length(x_valid) < 4) next  # Skip if too few values
    
    spearman <- cor.test(x_valid, y_valid, method = "spearman")
    kendall <- cor.test(x_valid, y_valid, method = "kendall")
    
    results_list[[paste(metric, analyte, sep = "_")]] <- data.frame(
      Metric = metric,
      Analyte = analyte,
      Spearman_r = as.numeric(spearman$estimate),
      Spearman_p = as.numeric(spearman$p.value),
      Kendall_tau = as.numeric(kendall$estimate),
      Kendall_p = as.numeric(kendall$p.value)
    )
  }
}

# Combine results
corr_full <- bind_rows(results_list)

# Add significance stars
corr_full <- corr_full %>%
  mutate(
    Spearman_sig = case_when(
      Spearman_p <= 0.001 ~ "***",
      Spearman_p <= 0.01  ~ "**",
      Spearman_p <= 0.05  ~ "*",
      TRUE ~ ""
    ),
    Kendall_sig = case_when(
      Kendall_p <= 0.001 ~ "***",
      Kendall_p <= 0.01  ~ "**",
      Kendall_p <= 0.05  ~ "*",
      TRUE ~ ""
    )
  )

View(corr_full)
#write.csv(corr_full, "/Users/cmantegna/Documents/Github/WDFWmussels/output/correlation_all_indices.csv", row.names = FALSE)

```

### individual analytes
```{r}

#correlation_data<- read.csv("../data/index_df/site_level_all_analytes_scaled.csv")

metrics_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# identifying columns to exclude from analysis
id_columns <- c("site_name", "site_number",  "reporting_area", "latitude", "longitude", "ai_z_arsenic", "ai_z_cadmium", "ai_z_copper", "ai_z_lead", "ai_z_mercury", "ai_z_zinc")

# converting metals to match other contaminants - metals are reported in mg/kg and all others are ng/g
# 1 mg/kg = 1,000 ng/g
correlation_data$ai_z_arsenic_ng <- correlation_data$ai_z_arsenic * 1000
correlation_data$ai_z_cadmium_ng <- correlation_data$ai_z_cadmium * 1000
correlation_data$ai_z_copper_ng <- correlation_data$ai_z_copper * 1000
correlation_data$ai_z_lead_ng <- correlation_data$ai_z_lead * 1000
correlation_data$ai_z_mercury_ng <- correlation_data$ai_z_mercury * 1000
correlation_data$ai_z_zinc_ng <- correlation_data$ai_z_zinc * 1000

#write.csv(correlation_data, "/Users/cmantegna/Documents/Github/WDFWmussels/data/site_level_corrected_metals_all_analytes.csv", row.names = FALSE)

# Build analyte_vars by excluding both metrics and IDs
analyte_vars <- colnames(correlation_data)[!(colnames(correlation_data) %in% c(metrics_vars, id_columns))]

# set p-value & initialize empty list
alpha <- 0.05
results_list <- list()

# Loop over metric–analyte pairs across sites
for (metric in metrics_vars) {
  for (analyte in analyte_vars) {
    
    x <- correlation_data[[metric]]
    y <- correlation_data[[analyte]]
    
    valid_idx <- complete.cases(x, y)
    x_valid <- x[valid_idx]
    y_valid <- y[valid_idx]
    
    if (length(x_valid) < 4) next  # Skip if too few values
    
    spearman <- cor.test(x_valid, y_valid, method = "spearman")
    kendall <- cor.test(x_valid, y_valid, method = "kendall")
    
    results_list[[paste(metric, analyte, sep = "_")]] <- data.frame(
      Metric = metric,
      Analyte = analyte,
      Spearman_r = as.numeric(spearman$estimate),
      Spearman_p = as.numeric(spearman$p.value),
      Kendall_tau = as.numeric(kendall$estimate),
      Kendall_p = as.numeric(kendall$p.value)
    )
  }
}

# Combine results
corr_full <- bind_rows(results_list)

# Add significance stars
corr_full <- corr_full %>%
  mutate(
    Spearman_sig = case_when(
      Spearman_p <= 0.001 ~ "***",
      Spearman_p <= 0.01  ~ "**",
      Spearman_p <= 0.05  ~ "*",
      TRUE ~ ""
    ),
    Kendall_sig = case_when(
      Kendall_p <= 0.001 ~ "***",
      Kendall_p <= 0.01  ~ "**",
      Kendall_p <= 0.05  ~ "*",
      TRUE ~ ""
    )
  )

View(corr_full)
#write.csv(corr_full, "/Users/cmantegna/Documents/Github/WDFWmussels/output/correlation_individual_analytes.csv", row.names = FALSE)

```

# Spatial Analysis
## Global Moran's I
```{r}

# fix this in post - this can be a loop, but I will need help making it work properly; this is why it is manually done

# all measured metrics
# p450
site_data <- site_sf %>% filter(!is.na(p450)) # filter out any NAs
summary(site_data$p450) # check for variance and more than 3 rows for analysis
coords <- st_coordinates(site_data) # get coordinated from included sites
nb <- knn2nb(knearneigh(coords, k = 4)) # build neighbor list and weights
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$p450)[, 1] # scale results and run morans
moran.test(x_scaled, lw)

# sod
site_data <- site_sf %>% filter(!is.na(sod))
summary(site_data$sod)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$sod)[, 1]
moran.test(x_scaled, lw)

# shell
site_data <- site_sf %>% filter(!is.na(shell))
summary(site_data$shell)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$shell)[, 1]
moran.test(x_scaled, lw)

# ci1
site_data <- site_sf %>% filter(!is.na(ci1))
summary(site_data$ci1)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci1)[, 1]
moran.test(x_scaled, lw)

# ci2
site_data <- site_sf %>% filter(!is.na(ci2))
summary(site_data$ci2)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci2)[, 1]
moran.test(x_scaled, lw)

# ci3
site_data <- site_sf %>% filter(!is.na(ci3))
summary(site_data$ci3)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ci3)[, 1]
moran.test(x_scaled, lw)


# ibr_bio
site_data <- site_sf %>% filter(!is.na(ibr_bio))
summary(site_data$ibr_bio)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_bio)[, 1]
moran.test(x_scaled, lw)

# ibr_morph
site_data <- site_sf %>% filter(!is.na(ibr_morph))
summary(site_data$ibr_morph)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_morph)[, 1]
moran.test(x_scaled, lw)

# ibr_combined
site_data <- site_sf %>% filter(!is.na(ibr_combined))
summary(site_data$ibr_combined)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ibr_combined)[, 1]
moran.test(x_scaled, lw)

# All contaminant indices
# mixture index, sum
site_data <- site_sf %>% filter(!is.na(mixture_index_sum))
summary(site_data$mixture_index_sum)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$mixture_index_sum)[, 1]
moran.test(x_scaled, lw)

# mixture index, pca
site_data <- site_sf %>% filter(!is.na(mixture_index_pca))
summary(site_data$mixture_index_pca)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$mixture_index_pca)[, 1]
moran.test(x_scaled, lw)

#chlordane
site_data <- site_sf %>% filter(!is.na(chlordane))
summary(site_data$chlordane)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$chlordane)[, 1]
moran.test(x_scaled, lw)

# UPDATE the CODE from HERE
#ddt
site_data <- site_sf %>% filter(!is.na(ddt))
summary(site_data$ddt)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$ddt)[, 1]
moran.test(x_scaled, lw)

#hch
site_data <- site_sf %>% filter(!is.na(hch))
summary(site_data$hch)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$hch)[, 1]
moran.test(x_scaled, lw)

#metal - nutrients
site_data <- site_sf %>% filter(!is.na(nutrient_metal_ng))
summary(site_data$nutrient_metal_ng)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$nutrient_metal_ng)[, 1]
moran.test(x_scaled, lw)

#metal - heavy
site_data <- site_sf %>% filter(!is.na(heavy_metal_ng))
summary(site_data$heavy_metal_ng)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$heavy_metal_ng)[, 1]
moran.test(x_scaled, lw)

#pah_lmw
site_data <- site_sf %>% filter(!is.na(pah_lmw))
summary(site_data$pah_lmw)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_lmw)[, 1]
moran.test(x_scaled, lw)

#pah_hmw
site_data <- site_sf %>% filter(!is.na(pah_hmw))
summary(site_data$pah_hmw)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_hmw)[, 1]
moran.test(x_scaled, lw)

#pah_add
site_data <- site_sf %>% filter(!is.na(pah_add))
summary(site_data$pah_add)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pah_add)[, 1]
moran.test(x_scaled, lw)

#pbde
site_data <- site_sf %>% filter(!is.na(pbde))
summary(site_data$pbde)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pbde)[, 1]
moran.test(x_scaled, lw)

#pcb
site_data <- site_sf %>% filter(!is.na(pcb))
summary(site_data$pcb)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pcb)[, 1]
moran.test(x_scaled, lw)

#pesticides
site_data <- site_sf %>% filter(!is.na(pesticide))
summary(site_data$pesticide)
coords <- st_coordinates(site_data)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")
x_scaled <- scale(site_data$pesticide)[, 1]
moran.test(x_scaled, lw)

```

## LISA - Local Indicators of Spatial Association
```{r}

library(sf)
library(spdep)

# convert df to a spatial object
site_sf <- st_as_sf(site_data, coords = c("longitude", "latitude"), crs = 4326)
site_sf <- st_transform(site_sf, crs = 3857)  # projected CRS for distance-based neighbors

# create spatial neighbors and weights (4 nearest neighbors)
coords <- st_coordinates(site_sf)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

# response variables to test
bio_vars <- c("p450", "sod", "shell", "ci1", "ci2", "ci3", "weight_initial", "weight_change", "weight_final", "length", "width", "height", "ibr_bio", "ibr_morph", "ibr_combined")

# contaminant indices 
index_vars <- c("chlordane", "ddt", "hch", "pah_lmw", "pah_hmw", "pah_add", "pbde", "pcb", "pesticide", "heavy_metal_ng", "nutrient_metal_ng", "mixture_index_sum", "mixture_index_pca")

# LOOP FOR LOCAL MORAN'S I 

# create output folder
dir.create("lisa_outputs", showWarnings = FALSE)

for (var in c(bio_vars, index_vars)) {
  x <- site_sf[[var]]
  lisa <- localmoran(scale(x)[,1], lw)

  # Add LISA results to spatial dataframe
  site_sf[[paste0(var, "_lisa_I")]] <- lisa[,1]
  site_sf[[paste0(var, "_lisa_p")]] <- lisa[,5]

  # Define cluster types with p-value significance threshold
site_sf[[paste0(var, "_lisa_cluster")]] <- case_when(
  x > mean(x, na.rm = TRUE) & lisa[,1] > 0 & lisa[,5] < 0.05 ~ "High-High",
  x < mean(x, na.rm = TRUE) & lisa[,1] > 0 & lisa[,5] < 0.05 ~ "Low-Low",
  lisa[,1] < 0 & lisa[,5] < 0.05 ~ "Outlier",
  TRUE ~ "Not Significant"
)

  # Save spatial cluster results with p-values included
st_write(site_sf[, c("site_name", paste0(var, "_lisa_cluster"), paste0(var, "_lisa_p"))],
         paste0("lisa_outputs/LISA_", var, ".geojson"), delete_dsn = TRUE)
}

```

## plotting LISA output
### testing p450 to spot check LISA results
```{r}

getwd()
library(sf)
lisa_p450<- st_read("lisa_outputs/LISA_p450.geojson")
names(lisa_p450)
head(lisa_p450)

```

## mapping p450 LISA clusters to visualize the results before moving to the next steps
```{r}

library(tmap)

# basic LISA plot
tm_shape(lisa_p450) +
  tm_symbols(col = "p450_lisa_cluster",
             palette = "Set1",
             size = 0.5,
             title.col = "LISA Cluster") +
  tm_layout(main.title = "LISA Clusters for P450 Biomarker")

# only signficant clusters
tm_shape(lisa_p450) +
  tm_symbols(col = "p450_lisa_cluster",
             palette = c("High-High" = "red",
                         "Low-Low" = "blue",
                         "Outlier" = "purple",
                         "Not Significant" = "grey80"),
             size = 0.5,
             title.col = "Cluster Type") +
  tm_layout(main.title = "Significant LISA Clusters (P450)")

# table of results
table(lisa_p450$p450_lisa_cluster)

# adding WA basemap for another way to see the clusters
library(rnaturalearth)

# grab WA
us_states <- ne_states(country = "united states of america", returnclass = "sf")
wa_outline <- us_states[us_states$name == "Washington", ]

library(tmap)

tm_shape(wa_outline) +
  tm_borders(col = "gray30") +
tm_shape(lisa_p450) +
  tm_symbols(col = "p450_lisa_cluster",
             palette = c("High-High" = "red",
                         "Low-Low" = "blue",
                         "Outlier" = "purple",
                         "Not Significant" = "gray80"),
             size = 0.5,
             title.col = "Cluster Type") +
  tm_layout(main.title = "LISA Clusters for P450 Biomarker")


```

## creating LISA summary table
```{r}

# Directly from Chat GPT

library(stringr)

# Get list of all LISA geojson files
lisa_files <- list.files("lisa_outputs", pattern = "\\.geojson$", full.names = TRUE)

# Create a summary table for each
lisa_summary <- lapply(lisa_files, function(file) {
  data <- st_read(file, quiet = TRUE)
  
  # Extract the variable name from the file name
  var_name <- str_extract(basename(file), "(?<=LISA_).*(?=\\.geojson)")
  
  # Identify cluster column (should be the only one with 'cluster' in name)
  cluster_col <- names(data)[grepl("cluster", names(data))][1]
  
  # Summarize cluster counts
  summary <- data %>%
    st_drop_geometry() %>%
    count(!!sym(cluster_col), name = "n") %>%
    mutate(variable = var_name, cluster = !!sym(cluster_col)) %>%
    select(variable, cluster, n)
  
  return(summary)
})

# Combine all summaries into one dataframe
all_lisa_summary <- bind_rows(lisa_summary)

#write.csv(all_lisa_summary, "/Users/cmantegna/Documents/Github/WDFWmussels/output/lisa_summary.csv", row.names = FALSE) # write it out

```

