---
title: "Morphometrics Table Values"
output:
  
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 12
  pdf_document: 
    fig_width: 10
    fig_height: 7
---

# Setup

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 12,       # Set plot width in inches
  fig.height = 8,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r, eval= FALSE}
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

### Note:

For *data* the units are listed below. Weight = g\
Length, width, height = mm\
p450, SOD = activity/ (mg/protein)\
Condition factor, economic factor = unitless\
For *pah*, *indv*, and *allana* the units are ng/g\
For *metal* the units are mg/kg

```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarker_thickness_condition.csv")

morph <- readxl::read_excel("/Users/cmantegna/Documents/Biomarker Data Analysis/all data table.xlsx", sheet = "Sheet1")

#pah has complete site values and different summed pah analyte whole tissue values 
#pah<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_pah_analytes.csv")

#indv has complete site values and individual named pah analyte whole tissue values 
#indv<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_analytes.csv") 

#metal<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/metal.csv")

#all<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/all_analytes.csv")

#man_df <-read.csv("/Users/cmantegna/Documents/WDFWmussels/data/man_df.csv")

```

# fix values

```{r}
# Data contains 0's and must be adjusted in this order to preserve all usable data.

#sod
#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
data$SOD[data$SOD <= 0] <- 0.0025

#p450
#remove any p450 values that are 0 - those are true 0's not non-detectable. I am replacing with na so I don't lose the entire row of data, including the SOD values.
data$p450[data$p450 <= 0] <- NA
```

#data for basic stats table for biomarkers, condition factor and shell thickness

```{r}
# Convert 'condition_factor' and 'avg_thickness' to numeric
data$condition_factor <- as.numeric(data$condition_factor)
data$avg_thickness <- as.numeric(data$avg_thickness)

```

#quartiles and median values \# adjusting p450 to only include values in the 2.5 - 97.5 percentile

```{r}
#install.packages("dplyr")
library(dplyr)

```

```{r}
library(dplyr)

# Define the calculate_stats function to handle NA and provide clear output
calculate_stats <- function(x) {
  valid_x = x[!is.na(x)]  # Remove NA values before calculations
  range_val <- range(valid_x)
  percentile_range <- quantile(valid_x, probs = c(0.025, 0.975), na.rm = TRUE)
  median_val <- median(valid_x, na.rm = TRUE)
  c(Range_Min = range_val[1], Range_Max = range_val[2], 
    Percentile_2.5 = percentile_range[1], Percentile_97.5 = percentile_range[2], 
    Median = median_val)
}

# Sample data assuming df is your original dataframe
# Here we ensure to use your actual column names for metrics

result <- data %>%
  group_by(site_name) %>%
  summarise(across(c(p450, SOD, condition_factor, avg_thickness), 
                   .fns = list(
                     Range_Min = ~min(.x, na.rm = TRUE),
                     Range_Max = ~max(.x, na.rm = TRUE),
                     Percentile_2.5 = ~quantile(.x, probs = 0.025, na.rm = TRUE),
                     Percentile_97.5 = ~quantile(.x, probs = 0.975, na.rm = TRUE),
                     Median = ~median(.x, na.rm = TRUE)
                   ),
                   .names = "{.col}_{.fn}"))

```

```{r}
print(result)

write.csv(result, "morphometricinfo.csv", row.names = FALSE)
```

```{r}
library(dplyr)

# Assuming df is your data frame and metric1, metric2, metric3, metric4 are your columns
result2 <- data %>%
  summarise(across(c(p450, SOD, condition_factor, avg_thickness), calculate_stats, .names = "{.col}_{.fn}"))

print(result2)
write.csv(result2, "morphometricinfo2.csv", row.names = FALSE)

```

```{r}
library(readxl)
#install.packages("openxlsx")
library(openxlsx)

```

```{r}

library(knitr)
#install.packages("kableExtra")
library("kableExtra")

# Create a simple table using kable and make headers span across columns
kable_styled <- kable(morph, "html", escape = FALSE, row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  add_header_above(c(" " = 1, "Condition Factor" = 3, "Shell Thickness (mm)" = 3, "p450 (activity mg-1 protein)"=3, "SOD (activity mg-1 protein)" =3))   

# Print the styled table 
kable_styled

```
