---
title: "01- Data Exploration"
output:
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
  pdf_document: 
    fig_width: 20
    fig_height: 9
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}
library(tinytex)
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

Weight = mg\
Length, width, height = mm\
p450, SOD = activity/ (mg/protein)\
Condition factor, economic factor = unitless

```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
#data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarkerfull.csv")

#alldata has the site names, biomarkers, condition factor, average thickness and analyte data - each row is an individual sample
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/p450data.csv")
sdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/soddata.csv")

```

## fix zero's in the data frame and alldata frame

```{r}

# Data contains 0's and must be adjusted in this order to preserve all usable data.

#sod
#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
sdata$sod[data$sod <= 0] <- 0.0025
#alldata$SOD[alldata$SOD <= 0] <- 0.0025
#p450
#remove any p450 values that are 0 - those are true 0's not non-detectable. I am replacing with na so I don't lose the entire row of data, including the SOD values.
#data$p450[data$p450 <= 0] <- NA
#alldata$p450[alldata$p450 <= 0] <- NA

#write.csv(alldata, "/Users/cmantegna/Documents/WDFWmussels/data/alldata.csv")
```

```{r}
# check the data frame
summary(sdata)
#str(alldata)
```

# boxplot of biomarker data, p450 by site

```{r}
#p450
pplot<- ggplot(data, aes(x = site_name, y = p450)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(pplot)

#SOD
#splot<- ggplot(data, aes(x = site_name, y = SOD)) +
 # geom_boxplot() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

#print(splot)
```

# ranked box plot, p450 by site
```{r}
#order the sites by value
data_ordered <- data[order(data$p450),]

#create a factor with the ordered site names
data_ordered$site_name <- factor(data_ordered$site_name, levels = unique(data_ordered$site_name))

#plot with ordered site names
ranked<- ggplot(data_ordered, aes(x = site_name, y = p450)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels if needed

print(ranked)
```

# boxplot of p450 by reporting area
```{r}

data$reporting_area <- as.character(data$reporting_area)

#p450
pplot<- ggplot(data, aes(x = reporting_area, y = p450)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(pplot)

```

# boxplot of p450 ranked by reporting area
```{r}

data$reporting_area <- as.character(data$reporting_area)

#order the sites by value
data_ordered <- data[order(data$p450),]

#create a factor with the ordered site names
data_ordered$reporting_area <- factor(data_ordered$reporting_area, levels = unique(data_ordered$reporting_area))

#plot with ordered site names
rankedra<- ggplot(data_ordered, aes(x = reporting_area, y = p450)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels if needed

print(rankedra)
```

```{r}
#sod
sdata$sod <- as.numeric(sdata$sod)
sdata$reporting_area <- as.character(sdata$reporting_area)

splot<- ggplot(sdata, aes(x = reporting_area, y = sod)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

print(splot)

#SOD
#splot<- ggplot(data, aes(x = site_name, y = SOD)) +
 # geom_boxplot() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

#print(splot)
```
# Violin plots
```{r}
#p450
ggplot(data, aes(x = as.factor(site_name), y = p450)) +
  geom_violin() +
  labs(title = "Variation of p450 within Sites", x = "Site", y = "p450 Value") +
  theme_minimal()


# SOD
ggplot(data, aes(x = as.factor(site_name), y = SOD)) +
  geom_violin() +
  labs(title = "Variation of SOD within Sites", x = "Site", y = "SOD Value") +
  theme_minimal()

```


# Histograms of p450, SOD, condition_factor and avg_thickness

### Only avg_thickness looks normally distributed.

```{r}
# p450 basic histogram + basic density plot
hist(alldata$p450)
plot(density(alldata$p450, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# SOD basic histogram + basic density plot
hist(alldata$SOD)
plot(density(alldata$SOD, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Condition_factor basic histogram + basic density plot
hist(alldata$condition_factor)
plot(density(alldata$condition_factor, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

# Avg_thickness basic histogram + basic density plot
hist(alldata$avg_thickness)
plot(density(alldata$avg_thickness, na.rm= TRUE), main="Density Plot", xlab="p450 Value")

```

# Shapiro-Wilkes test for normality

### Only avg_thickness is normally distributed.

```{r}
shapiro.test(data$p450)
#shapiro.test(alldata$SOD)
#shapiro.test(alldata$condition_factor)
#shapiro.test(alldata$avg_thickness)

```

# Kruskal-Wallis, p450

### p450 and site have a statistically significant relationship, p\< 0.0000008077

```{r}
kruskal.test(p450 ~ site_name, data = data)
kruskal.test(p450 ~ reporting_area, data = data)
#kruskal.test(p450 ~ condition_factor, data = alldata)
#kruskal.test(p450 ~ avg_thickness, data = alldata)

```

# Kruskal-Wallis, SOD

### SOD and site have a statistically significant relationship, p\< 0.0000005669

```{r}
#kruskal.test(SOD ~ site_name, data = alldata)
#kruskal.test(SOD ~ p450, data = alldata)
#kruskal.test(SOD ~ condition_factor, data = alldata)
#kruskal.test(SOD ~ avg_thickness, data = alldata)

```

# Post hoc test. Kruskal-Wallac Multiple Comparisons

### p450 and site_name show no true differences despite K-W test result.

# significant differences across reporting area pairwise matches:
8.2-12, 10-13, 11-12, and 12-13

Reporting Areas Are:
6 - East Juan de Fuca Strait
7 - San Juan Islands
8.1 - Deception Pass, Hope Island, and Skagit Bay
8.2 - Port Susan and Port Gardner
9 - Admiralty Inlet
10 - Seattle-Bremerton
11 - Tacoma-Vashon
12 - Hood Canal
13 - South Puget Sound

```{r}
library(pgirmess)

mc_site<- as.data.frame(kruskalmc(p450 ~ site_name, data = data, method = "bonferroni"))
mc_reporting<- as.data.frame(kruskalmc(p450 ~ reporting_area, data = data, method = "bonferroni")) 

head(mc_site)
head(mc_reporting)

```
