---
title: "03- Mapping Biomarkers"
output:
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
  pdf_document: 
    fig_width: 20
    fig_height: 9

---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}

#install.packages(c("ggplot2", "sf", "viridis", "rnaturalearth", "rnaturalearthdata"))
library(tidyr)
#library(tidyverse)
library(ggplot2)
library(vegan)
library(sf)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(ggmap)
#install.packages("elevatr")

```

# Load data
```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/reportingareas_pahgroups_for_maps.csv")

#alldata has the site names, biomarkers, condition factor, average thickness and analyte data - each row is an individual sample
#alldata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/alldata.csv")

```

# Map of Washington State

```{r}
world <- ne_states(country = "united states of america", returnclass = "sf")
washington_map <- world[world$name == "Washington", ]


```

# adjust data to a spacial data frame
```{r}

data_sf <- st_as_sf(data, coords = c("longitude", "latitude"), crs = 4326)

```

# reporting areas - color coded - plain background
```{r}

xlim <- c(-124, -122)  # longitude bounds
ylim <- c(47, 49)  # latitude bounds

ggplot() +
  geom_sf(data = washington_map, fill = "lightblue", color = "black") +  # Base map of Puget Sound
  geom_sf(data = data_sf, aes(color = as.factor(reporting_area)), size = 3) +  # Site points
  scale_color_manual(values = scales::hue_pal()(length(unique(data_sf$reporting_area)))) +  # Distinct colors
  theme_minimal() +
  labs(color = "Reporting Area", title = "Sites in Reporting Area") +
  theme(legend.position = "right")

```

# reporting areas, puget sound with topographic overlay
```{r}

library(ggplot2)
library(sf)
library(rnaturalearth)
library(elevatr)
library(raster)
library(viridis)

# Get the Washington state boundary (as a substitute for Puget Sound region)
washington_map <- ne_states(country = "united states of america", returnclass = "sf") %>%
  filter(name == "Washington")

# Set the bounding box (longitude and latitude range for Puget Sound area)
xlim <- c(-124, -122)  # longitude bounds
ylim <- c(47, 49)      # latitude bounds

# Create a data frame for the bounding box corner points
bbox_df <- data.frame(
  x = c(xlim[1], xlim[2], xlim[1], xlim[2]),
  y = c(ylim[1], ylim[1], ylim[2], ylim[2])
)

# Convert the bounding box data frame to an sf object (spatial points)
bbox_sf <- st_as_sf(bbox_df, coords = c("x", "y"), crs = 4326)

# Fetch elevation data for the bounding box using get_elev_raster
elevation_data <- get_elev_raster(locations = bbox_sf, z = 10, prj = "+proj=longlat +datum=WGS84")

# Convert the raster elevation data to a data frame for plotting with ggplot2
elevation_df <- as.data.frame(as(elevation_data, "SpatialPixelsDataFrame"))

# Rename the columns for ggplot2 compatibility
names(elevation_df) <- c("x", "y", "elevation")

# Plot the topographic map with site data
ggplot() +
  geom_raster(data = elevation_df, aes(x = x, y = y, fill = elevation)) +  # Elevation background
  geom_sf(data = washington_map, fill = NA, color = "black") +  # Washington state boundary
  geom_sf(data = data_sf, aes(color = as.factor(reporting_area)), size = 3) +  # Site points
  scale_fill_viridis_c(option = "C") +  # Color scale for elevation
  scale_color_manual(values = scales::hue_pal()(length(unique(data_sf$reporting_area)))) +  # Distinct colors
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +  # Properly set xlim and ylim
  theme_minimal() +
  labs(color = "Reporting Area", fill = "Elevation", title = "Sites with Topographic Background") +
  theme(legend.position = "right")

dem_data <- raster("/Users/cmantegna/Downloads/Puget_Sound_CoNED_Topobathy_DEM_1m/Puget_Sound_CoNED_Topobathy_DEM_1m.tif")
```

# new way to get a topographic overlay
```{r}


# Load the terra package
library(terra)

# Load the large DEM file using terra
#dem_data <- rast("path_to_your_download/Puget_Sound_CoNED_Topobathy_DEM_1m.tif")
dem_data <- rast("/Users/cmantegna/Desktop/Puget_Sound_CoNED_Topobathy_DEM_1m.tif")

# Plot a quick view of the DEM to check if it's loaded correctly
p <- plot(dem_data)

saveRDS(dem_data, "/Users/cmantegna/Documents/WDFWmussels/data/dem_data.rds")

# Save the raster object using save (if saving multiple objects)
save(dem_data, file = "/Users/cmantegna/Documents/WDFWmussels/data/dem_data2.RData")


# Convert the raster data to a data frame (this could take time due to size)
dem_df <- as.data.frame(dem_data, xy = TRUE)

# Use ggplot2 to plot the DEM with your site data
library(ggplot2)
ggplot() +
  geom_raster(data = dem_df, aes(x = x, y = y, fill = layer)) +  # Plot DEM as raster
  scale_fill_viridis_c(option = "C") +  # Color scale for elevation
  theme_minimal() +
  labs(fill = "Elevation", title = "Topobathymetric DEM of Puget Sound")


```




EVERYTHING BELOW THIS LINE IS OLD CODE. DO NOT NEED TO RERUN IT.
# p450 map- WA state

```{r}

pmap<- ggplot() + 
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point(data = alldata, aes(x = longitude, y = latitude, color = p450), size = 2) +
  scale_color_viridis(option = "C", name = "p450") +
  theme_minimal() +
  labs(title = "Washington State - p450 Data")

print(pmap)
ggsave(plot=pmap, filename="/Users/cmantegna/Documents/WDFWmussels/output/p450map.png", width=15, height=8)
```

# p450 map - Puget Sound

```{r}
#zoom into puget sound region & note the legend, lighter colors are higher values
xlim <- c(-124, -122)  # longitude bounds
ylim <- c(47, 49)  # latitude bounds

sodpugetsound<- ggplot() + 
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point(data = alldata, aes(x = longitude, y = latitude, color = p450), size = 3) +
  scale_color_viridis(option = "C", name = "SOD") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)+
  theme_minimal() +
  labs(title = "Washington State - SOD Data")

print(sodpugetsound)
```

# p450 map by regions

### fix the coordinates and maps, not a priority 4/21

```{r}
library(ggplot2)
library(dplyr)
library(sf)  

# Prepare data for plotting by adding a region column
data <- alldata %>%
  mutate(
    region = case_when(
      latitude >= 47.78184 & latitude <= 48.8208 ~ "North",
      latitude >= 47.38566 & latitude <= 47.7296079 ~ "Central",
      latitude >= 47.052362 & latitude <= 47.3539722 ~ "South"
    )
  ) %>%
  filter(!is.na(region) & !is.na(p450))

# Plotting p450 values by region over the Washington state map
ggplot(data = data, aes(x = longitude, y = latitude, color = p450)) +
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(~ region) +
  labs(title = "p450 Levels by Region in Washington State",
       x = "Longitude",
       y = "Latitude",
       color = "p450 Levels") +
  theme_minimal()
```

# SOD map- WA state

```{r}

smap<- ggplot() + 
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point(data = alldata, aes(x = longitude, y = latitude, color = SOD), size = 3) +
  scale_color_viridis(option = "C", name = "SOD") +
  theme_minimal() +
  labs(title = "Washington State - SOD Data")

print(smap)
ggsave(plot=smap, filename="/Users/cmantegna/Documents/WDFWmussels/output/sodmap.png", width=15, height=8)

```

#SOD map - Puget Sound

```{r}
#zoom into puget sound region & note the legend, lighter colors are higher values
xlim <- c(-124, -122)  # longitude bounds
ylim <- c(47, 49)  # latitude bounds

sodpugetsound<- ggplot() + 
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point(data = alldata, aes(x = longitude, y = latitude, color = SOD), size = 3) +
  scale_color_viridis(option = "C", name = "SOD") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)+
  theme_minimal() +
  labs(title = "Washington State - SOD Data")

print(sodpugetsound)
```

# SOD map by regions

### fix the coordinates and maps, not a priority 4/21

```{r}
library(ggplot2)
library(dplyr)
library(sf)  

# Prepare data for plotting by adding a region column
data <- alldata %>%
  mutate(
    region = case_when(
      latitude >= 47.78184 & latitude <= 48.8208 ~ "North",
      latitude >= 47.38566 & latitude <= 47.7296079 ~ "Central",
      latitude >= 47.052362 & latitude <= 47.3539722 ~ "South"
    )
  ) %>%
  filter(!is.na(region) & !is.na(SOD))

# Plotting p450 values by region over the Washington state map
ggplot(data = data, aes(x = longitude, y = latitude, color = SOD)) +
  geom_sf(data = washington_map, fill = "lightgrey", color = "white") +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap(~ region) +
  labs(title = "p450 Levels by Region in Washington State",
       x = "Longitude",
       y = "Latitude",
       color = "SOD Activity") +
  theme_minimal()
```
