---
title: "05- Analytes Data for Coorelation with Biomarker Values"
output:
  html_document: 
    fig_width: 8
  pdf_document: 
    fig_width: 8
    fig_height: 5
---

### Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 8,       # Set plot width in inches
  fig.height = 5,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r, eval= FALSE}

library(tidyr)
library(tidyverse)
library(ggplot2)
library(vegan)
library(reshape2)

```

# Load data

```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarkerfull.csv")

pah<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/sum_analytes.csv")

indv<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_analytes.csv") 

```

### Review setup

```{r}

str(data)
str(pah)
str(indv)

```

```{r}
summary(data)
summary(pah)
summary(indv)

```

# Linear Regression
# GLM

#### both model types are applied becasue I am unsure where to go - the GLM feels most correct because it is more flexible to non- normally distributed data.

```{r}
library(reshape2)
#merge data frames and reshape for input.

#merge by latitude and longitude
merged_df <- merge(data, pah, by = c("latitude", "longitude"))

#reshape to get the analytes into their own columns with the DryValue as their values
reshaped_df <- dcast(merged_df, latitude + longitude + p450 + SOD ~ Analyte, value.var = "DryValue")

print(reshaped_df)

```
# Linear Regression- p450 as response 
```{r}

lm_model <- lm(p450 ~ SOD + SumPAHs + SumPAHs16 + SumPAHs42_DMNcorrected + SumPAHsHMW + SumPAHsLMW, data = reshaped_df)
print(summary(lm_model))

```
# GLM - p450 as response
```{r}

glm_model <- glm(p450 ~ SOD + SumPAHs + SumPAHs16 + SumPAHs42_DMNcorrected + SumPAHsHMW + SumPAHsLMW, family = poisson(), data = reshaped_df)
print(summary(glm_model))

```
# LR- p450 as response
#### Replicating with individual analytes to see if anything really sticks out.
```{r}
#merge and reshape just like above
#merge data frames and reshape for input.
#fix column name difference so the merge can happen
colnames(indv)[colnames(indv) == "Latitude"] <- "latitude"
colnames(indv)[colnames(indv) == "Longitude"] <- "longitude"

#merge by latitude and longitude
merged_df2 <- merge(data, indv, by = c("latitude", "longitude"))

#reshape to get the analytes into their own columns with the DryValue as their values
reshaped2_df <- dcast(merged_df2, latitude + longitude + p450 + SOD ~ Analyte, value.var = "DryValue")

print(reshaped2_df)
```

```{r}
#get the column names so I don't have to individually type each one

all_columns <- names(reshaped2_df)

# Remove the columns you don't want to include in the model
excluded_columns <- c('p450', 'latitude', 'longitude', 'SOD')
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Enclose each column name in backticks to handle special characters
independent_columns <- sapply(independent_columns, function(x) paste0("`", x, "`"))

# Create a string representing the formula
formula_str <- paste("p450 ~", paste(independent_columns, collapse = " + "))

# Convert the string to a formula object
formula <- as.formula(formula_str)

# Now you can use this formula in your model
#lm_model <- lm(formula, data = reshaped_df)
#summary(lm_model)


```

```{r}

indvlm_model <- lm(formula, data = reshaped2_df)
print(summary(indvlm_model))

```

# GLM- p450 as response
```{r}

indvglm_model<- glm(formula, data = reshaped2_df, family = poisson())
print(summary(indvglm_model))

```

