---
title: "07- PCA + Pearson, SOD"
output:
  pdf_document: 
    fig_width: 10
    fig_height: 7
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 10
---

# Setup

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 8,       # Set plot width in inches
  fig.height = 5,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r, eval= FALSE}
library(tidyr)
library(tidyverse)
library(ggplot2)
library(vegan)
#library(tinytex)

```

# Load data

### Note:

For *data* the units are listed below. Weight = g\
Length, width, height = mm\
p450, SOD = activity/ (mg/protein)\
Condition factor, economic factor = unitless\
For *pah*, *indv*, and *allana* the units are ng/g\
For *metal* the units are mg/kg

```{r}

getwd()
#data has all sites, coordinates, p450, sod, condition factor, economic factor data
data<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/biomarkerfull.csv")

#pah has complete site values and different summed pah analyte whole tissue values 
pah<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/sum_analytes.csv")

#indv has complete site values and individual named pah analyte whole tissue values 
indv<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/individual_analytes.csv") 

metal<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/metal.csv")

allana<- read.csv("/Users/cmantegna/Documents/Biomarker Data Analysis/allana.csv")

```

```{r}
# Review data frame structure

#str(metal)
#str(allana)
#str(indv)

```

```{r}
# Review basic data types and stats

#summary(data)
#summary(pah)
#summary(indv)

```

```{r}

head(data)
head(metal)
head(allana)
```

# Data frame manipulations

### Adjusting biomarker values for accurate stats

```{r}
# Data contains 0's and must be adjusted in this order to preserve all usable data.

#sod
#replace any SOD values at or below 0 with half of the lower detection limit of .005 (.005*.5). Lower detection limit determined by assay protocol by the manufacturer, Cayman.
data$SOD[data$SOD <= 0] <- 0.0025

#p450
#remove any p450 values that are 0 - those are true 0's not non-detectable. I am replacing with na so I don't lose the entire row of data, including the SOD values.
data$p450[data$p450 <= 0] <- NA

```

### Data adjustment for analysis- SOD & p450

```{r}
#Average the

library(dplyr)

#simplifying the dataframe for joining with next steps
averaged_data <- data %>%
  group_by(site_number, latitude, longitude, site_name) %>%  
  summarise(
    avg_p450 = mean(p450, na.rm = TRUE),  
    avg_SOD = mean(SOD, na.rm = TRUE)
  ) %>%
  ungroup()  # Remove grouping for the new dataframe

print(averaged_data)

library(reshape2)
#merge data frames and reshape for input.
colnames(allana)[colnames(allana) == "SiteName"] <- "site_name"
merged_df <- merge(averaged_data, allana, by = c("site_name"), all.x = TRUE)

#reshape to get the analytes into their own columns with the DryValue as their values
reshaped_df <- dcast(merged_df, site_name + site_number +latitude + longitude + avg_p450 + avg_SOD ~ Analyte, value.var = "DryValue")

head(reshaped_df)

```

### Data frame for SOD GLM & PCA- All summed analytes and metals

```{r}

#create a table without the avg_p450 and NA column for SOD work
cols_to_keep <- colnames(reshaped_df)[!colnames(reshaped_df) %in% c("avg_p450", "NA")]

sod_all <- reshaped_df[, cols_to_keep]

head(sod_all)

#create a table without the avg_SOD and NA column for p450 work
#cols_to_keep2 <- colnames(reshaped2_df)[!colnames(reshaped2_df) %in% c("avg_SOD", "arsenic", "cadmium", "copper", "lead", "mercuryTotal", "Sum40CBs", "SumBDEs", "SumCHLDs", "SumDDTs", "SumHCHs", "SumPCBs2x17", "Zinc", "NA")]

#p450PAH <- reshaped2_df[, cols_to_keep2]
#p450PAH$plogdata <- plogdata

#print(p450PAH)

```
#SOD Pearson- summed analytes + metals

```{r}
#get the column names from sod_all so I don't have to individually type each one

all_columns <- names(sod_all)

# Remove the columns you don't want to include in the model
excluded_columns <- c('latitude', 'longitude', 'site_name', 'site_number')
independent_columns <- all_columns[!all_columns %in% excluded_columns]

# Enclose each column name in backticks to handle special characters
independent_columns <- sapply(independent_columns, function(x) paste0("`", x, "`"))

# Create a string representing the formula
formula_str <- paste("avg_SOD ~", paste(independent_columns, collapse = " + "))

# Convert the string to a formula object
formula <- as.formula(formula_str)

#SODall_glm<- glm(formula, data = sod_all, family = poisson())
#print(summary(SODall_glm))
```

```{r}
library(corrplot)

# Extract variable names from the formula
variables <- all.vars(formula)

# Subset the dataframe 'sod_all' using the extracted variables
subset_data <- sod_all[, variables]

# Compute Pearson correlation for each pair of variables
correlation_results <- cor(subset_data, method = "pearson", use = "complete.obs")

# View the correlation matrix
print(correlation_results)

corrplot(correlation_results, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

```

# SOD PCA - all analytes + metals

```{r}
# PCA Plot with biomarkers
#install.packages("FactoMineR")
#install.packages("factoextra")
library('FactoMineR')
library("factoextra")


# Remove NAs from the dataset
df_clean <- na.omit(sod_all)

# Selecting the relevant variables for PCA
pca_data <- df_clean[, c("avg_SOD", "arsenic", "cadmium", "copper", "lead", "mercuryTotal", "Sum40CBs", "SumBDEs", "SumCHLDs", "SumDDTs", "SumHCHs", "SumPAHs", "SumPAHs16", "SumPAHs42_DMNcorrected", "SumPAHsHMW", "SumPAHsLMW", "SumPCBs2x17", "Zinc")]

# Performing PCA
pca_res <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Plotting the PCA
pcaplot<- fviz_pca_biplot(pca_res, label = "var", col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                repel = TRUE)  # Avoid text overlapping (slow if many points)



print(pcaplot)
#ggsave(plot=pcaplot, filename="/Users/cmantegna/Documents/WDFWmussels/output/pca.png", width=15, height=8)

```





