---
title: "Pivoting & merging data frames"
output:
  html_document: 
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    fig_width: 20
  pdf_document: 
    fig_width: 20
    fig_height: 9
---

# Directory and doc rules

```{r, setup, eval=TRUE, include=TRUE}

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 20,       # Set plot width in inches
  fig.height = 9,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

```

# Load packages

```{r}
library(tidyr)
library(tidyverse)
library(vegan)

```

# Load data

```{r}

#getwd()
pdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgP450.csv")
sdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/avgSOD.csv")
#mdata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/morphometricdata.csv")
idata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/not_for_analysis/individual.csv")
odata<- read.csv("/Users/cmantegna/Documents/WDFWmussels/data/not_for_analysis/othersummed.csv")

```

```{r}
summary(odata)
```

```{r}
summary(idata)
```

# Pivot data table

```{r}
#individual
df_wide <- idata %>%
  pivot_wider(names_from = analyte, values_from = dryvalue)

# Print the reshaped data frame
print(df_wide)
write.csv(df_wide, "/Users/cmantegna/Documents/WDFWmussels/data/individual_wide.csv", row.names = FALSE)

#other summed
df_wide2 <- odata %>%
  pivot_wider(names_from = analyte , values_from = dryvalue )

# Print the reshaped data frame
print(df_wide2)
write.csv(df_wide2, "/Users/cmantegna/Documents/WDFWmussels/data/othersummed_wide.csv", row.names = FALSE)
```

# Merge 
## All of the data frames are merged but only the p450_analytes2 has been adjusted for analysis. all others need to be adjusted before completing the next merge step.
```{r}
#p450
merged_df <- merge(pdata, df_wide, by = "site_name", all = TRUE)

# Print the merged data frame
head(merged_df)
write.csv(merged_df, "/Users/cmantegna/Documents/WDFWmussels/data/avgP450_analytes.csv", row.names = FALSE)

#sod
merged_df <- merge(sdata, df_wide, by = "site_name", all = TRUE)

# Print the merged data frame
head(merged_df)
write.csv(merged_df, "/Users/cmantegna/Documents/WDFWmussels/data/avgSOD_analytes.csv", row.names = FALSE)

#morphometric
#merged_df <- merge(mdata, df_wide, by = "site_name", all = TRUE)

# Print the merged data frame
#print(merged_df)
#write.csv(merged_df, "/Users/cmantegna/Documents/WDFWmussels/data/morphometric_analytes2.csv", row.names = FALSE)

```

## All of the data frames are merged but only the p450_analytes3 has been adjusted for analysis. all others need to be adjusted before completing the next merge step.
```{r}
#p450
merged_df2 <- merge(pdata, df_wide2, by = "site_name", all = TRUE)

# Print the merged data frame
head(merged_df2)
#write.csv(merged_df2, "/Users/cmantegna/Documents/WDFWmussels/data/p450_analytes3.csv", row.names = FALSE)

#sod
merged_df2 <- merge(sdata, df_wide2, by = "site_name", all = TRUE)

# Print the merged data frame
head(merged_df2)
#write.csv(merged_df2, "/Users/cmantegna/Documents/WDFWmussels/data/sod_analytes3.csv", row.names = FALSE)

#morphometric
merged_df2 <- merge(mdata, df_wide2, by = "site_name", all = TRUE)

# Print the merged data frame
print(merged_df2)
#write.csv(merged_df2, "/Users/cmantegna/Documents/WDFWmussels/data/morphometric_analytes2.csv", row.names = FALSE)
```
